---
title: "Modeling - Take 1"
format: html
author: Sujan Kakumanu
code-fold: show
---

```{r setup}
#| message: false
#| warning: false

options(scipen = 999)

if(!require(pacman)){install.packages("pacman"); library(pacman, quietly = T)}
p_load(tidyverse, sf, viridis, lubridate)
```

```{r setup-2}
baseline <- read.csv("../ridership_eda/baseline_model_stop_trip.csv")
```


## Cleanup of Baseline Data

We're working on cleaning up the baseline feature file for easier joining to AIS and Swiftly. Before then, doing some direct cleanup here.
```{r baseline-cleanup}
baseline_clean <- baseline %>%
  filter(model_eligible == "True") %>%
  mutate(
    date      = as.Date(str_extract(trip_key, "^\\d{4}-\\d{2}-\\d{2}")),
    trip_id   = str_extract(trip_key, "(?<=\\|)\\d+(?=\\|)"),
    route     = str_extract(trip_key, "(?<=\\|)[A-Z]+(?=\\|)"),
    direction = str_extract(trip_key, "[A-Z]+$")
  )
```
## Model Development

We build a linear regression model to understand factors associated with arrival
lateness. We use a temporal train/test split and time-based cross validation to prevent data leakage.
Random cross validation is inappropriate here because trips close in time share
similar conditions — a random split would allow the model to learn from the future.

### Data Preparation
```{r model-prep}
p_load(tidymodels)

# Remove rows with NA outcome
model_data <- baseline_clean %>%
  filter(!is.na(arr_late_min)) %>%
  select(
    arr_late_min,
    dep_hour,
    schedule_season,
    schedule_daytype,
    route,
    direction,
    stop_sequence,
    vessel_capacity,
    hdwy_prev_min,
    turn_slack_sched_min,
    prev_terminal_late_pos,
    prev_stop_late_pos,
    boardings_wake_obs,
    onboard_obs,
    date
  )

nrow(model_data)
summary(model_data$arr_late_min)
```

### Train/Test Split
```{r train-test-split}
# Temporal split: train on everything before June 2025, test on June 2025
train_data <- model_data %>% filter(date < ymd("2025-06-01"))
test_data  <- model_data %>% filter(date >= ymd("2025-06-01"))

nrow(train_data)
nrow(test_data)
```

### Time-Based Cross Validation
```{r cv-folds}
# Leave one month out at a time — prevents leakage across time periods
train_data <- train_data %>%
  mutate(cv_month = format(date, "%Y-%m"))

cv_folds <- group_vfold_cv(train_data, group = cv_month, v = 10)
cv_folds
```

### Recipe
```{r recipe}
model_rec <- recipe(arr_late_min ~ ., data = train_data) %>%
  update_role(date, new_role = "id") %>%
  update_role(cv_month, new_role = "id") %>%
  step_impute_median(hdwy_prev_min, turn_slack_sched_min,
                     prev_terminal_late_pos, prev_stop_late_pos) %>%
  step_dummy(all_nominal_predictors()) %>%
  step_zv(all_predictors()) %>%
  step_normalize(all_numeric_predictors())
```

### Model Specification and Workflow
```{r lm-workflow}
lm_spec <- linear_reg() %>%
  set_engine("lm")

lm_wf <- workflow() %>%
  add_recipe(model_rec) %>%
  add_model(lm_spec)
```

### Fit with Cross Validation
```{r lm-fit}
control <- control_resamples(save_pred = TRUE)
metrics <- metric_set(rmse, rsq, mae)

lm_tuned <- fit_resamples(
  lm_wf,
  resamples = cv_folds,
  control   = control,
  metrics   = metrics
)

collect_metrics(lm_tuned)
```

### Validate on Test Set
```{r lm-validate}
# Drop cv_month before final fit
model_rec_final <- recipe(arr_late_min ~ ., data = train_data %>% select(-cv_month)) %>%
  update_role(date, new_role = "id") %>%
  step_impute_median(hdwy_prev_min, turn_slack_sched_min,
                     prev_terminal_late_pos, prev_stop_late_pos) %>%
  step_dummy(all_nominal_predictors()) %>%
  step_zv(all_predictors()) %>%
  step_normalize(all_numeric_predictors())

lm_wf_final <- workflow() %>%
  add_recipe(model_rec_final) %>%
  add_model(lm_spec)

data_split <- make_splits(
  list(analysis   = which(model_data$date < ymd("2025-06-01")),
       assessment = which(model_data$date >= ymd("2025-06-01"))),
  model_data
)

lm_val_fit <- last_fit(lm_wf_final, split = data_split, metrics = metrics)
collect_metrics(lm_val_fit)

collect_metrics(lm_tuned)
```

## Model Validation - More Thorough

```{r extract-predictions}
val_predictions <- lm_val_fit %>%
  collect_predictions() %>%
  bind_cols(test_data %>% select(route, stop_sequence, 
                                  schedule_season, schedule_daytype,
                                  dep_hour, vessel_capacity))  %>%
  mutate(residual = arr_late_min - .pred)
```

```{r error-by-route}
val_predictions %>%
  group_by(route) %>%
  summarise(
    rmse = sqrt(mean(residual^2)),
    mae  = mean(abs(residual)),
    bias = mean(residual),
    n    = n()
  ) %>%
  arrange(desc(rmse)) %>%
  knitr::kable(caption = "Model Error by Route", digits = 2)
```

```{r error-by-hour}
val_predictions %>%
  group_by(dep_hour) %>%
  summarise(bias = mean(residual), .groups = "drop") %>%
  ggplot(aes(x = dep_hour, y = bias)) +
  geom_col(fill = "steelblue") +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(title = "Model Bias by Hour of Day",
       subtitle = "Positive = model underpredicts lateness",
       x = "Departure Hour", y = "Mean Residual (minutes)") +
  theme_minimal()
```

```{r error-by-season}
val_predictions %>%
  group_by(schedule_season, schedule_daytype) %>%
  summarise(
    rmse = sqrt(mean(residual^2)),
    bias = mean(residual),
    n    = n(),
    .groups = "drop"
  ) %>%
  arrange(desc(rmse)) %>%
  knitr::kable(caption = "Model Error by Season and Day Type", digits = 2)
```
### Model Validation: Key Takeaways

**Overall performance** — RMSE of 2.44 minutes, MAE of 1.80 minutes, and R² of 0.57 on the holdout set. The train/holdout gap is negligible, indicating the model generalizes well without overfitting.

**By route** — GI shows the largest bias (+2.21 min), meaning the model consistently underpredicts lateness there. RW has the highest error overall but near-zero bias, suggesting high variance the current features cannot capture. ER performs well despite being the highest ridership route.

**By hour** — Bias is small in magnitude across all hours (under 0.5 minutes). The model slightly overpredicts lateness at hour 5 and slightly underpredicts across most of the midday and afternoon window. No severe hourly failure.

**By season and day type** — The holdout covers only SUMMER_1 given the June 2025 test window. Weekends show slightly higher RMSE than weekdays, consistent with leisure travel adding unpredictability. Bias is small across all groups.

**NEXT STEPS** — GI underprediction points to factors not yet captured. AIS traffic features and weather/tide data could be areas of improvement, particularly for RW and GI.

```{r}
lm_val_fit %>%
  extract_fit_parsnip() %>%
  tidy()
```
