---
title: "Modeling - Take 2"
format: html
author: Sujan Kakumanu
code-fold: show
---
```{r setup}
#| message: false
#| warning: false
options(scipen = 999)
if(!require(pacman)){install.packages("pacman"); library(pacman, quietly = T)}
p_load(tidyverse, sf, viridis, lubridate, tidymodels, themis)
```

## Overview

This document builds a predictive model for severe ferry delays, defined as arrivals
more than 10 minutes late. The model is designed to generate forecasts 48 hours in
advance, meaning all features must be knowable before the day of operation. Features
derived from same-day observations — such as actual boardings or real-time schedule
adherence — are explicitly excluded.

The outcome variable is binary: `is_severely_late` (1 if `arr_late_min > 10`, 0
otherwise). Given that severe delays represent only 2.2% of trips, we address class
imbalance via downsampling and evaluate performance using ROC AUC, precision, recall,
and F1 rather than accuracy.

## Data Loading
```{r load-data}
baseline <- read.csv("../ridership_eda/baseline_model_stop_trip.csv")

historical_averages <- read.csv("../ais-eda/AIS_historical_averages.csv") %>%
  mutate(
    stop_id     = as.character(stop_id),
    day_of_week = as.character(day_of_week)
  )
```

## Data Cleaning

```{r clean-data}
model_data <- baseline %>%
  mutate(
    date      = as.Date(str_extract(trip_key, "^\\d{4}-\\d{2}-\\d{2}")),
    trip_id   = str_extract(trip_key, "(?<=\\|)\\d+(?=\\|)"),
    route     = str_extract(trip_key, "(?<=\\|)[A-Z]+(?=\\|)"),
    direction = str_extract(trip_key, "[A-Z]+$")
  ) %>%
  filter(model_eligible == "True") %>%
  filter(!is.na(arr_late_min)) %>%
  mutate(
    stop_id     = as.character(as.integer(str_extract(segment_key, "\\d+$"))),
    day_of_week = as.character(wday(date, label = TRUE))
  ) %>%
  left_join(
    historical_averages,
    by = c("stop_id", "dep_hour" = "hour",
           "schedule_season", "schedule_daytype"),
    relationship = "many-to-many"
  ) %>%
  filter(day_of_week.x == day_of_week.y) %>%
  select(-day_of_week.y) %>%
  rename(day_of_week = day_of_week.x) %>%
  mutate(is_severely_late = factor(
    if_else(arr_late_min > 10, "yes", "no"),
    levels = c("yes", "no")
  ))

glimpse(model_data)
nrow(model_data)
```

```{r class-imbalance}
model_data %>%
  count(is_severely_late) %>%
  mutate(pct = n / sum(n))
```

## Feature Exploration and Multicollinearity Check

Before building the model we explore relationships between features to identify
redundancy and multicollinearity.

### Numeric Feature Correlations
```{r numeric-correlations}
model_data %>%
  select(dep_hour, stop_sequence, vessel_capacity, 
         turn_slack_sched_min, avg_vessel_count, median_vessel_count) %>%
  cor(use = "complete.obs") %>%
  corrplot::corrplot(
    method  = "color",
    type    = "upper",
    addCoef.col = "black",
    tl.col  = "black",
    tl.srt  = 45,
    diag    = FALSE,
    title   = "Correlation Matrix: Numeric Features",
    mar     = c(0, 0, 1, 0)
  )
```

### Categorical Feature Overlap
```{r categorical-overlap}
# Does day_of_week add information beyond schedule_daytype?
model_data %>%
  count(day_of_week, schedule_daytype) %>%
  arrange(schedule_daytype, day_of_week) %>%
  knitr::kable(caption = "Day of Week vs Schedule Day Type")

# Does vessel_capacity vary within routes?
model_data %>%
  count(route, vessel_capacity) %>%
  arrange(route) %>%
  knitr::kable(caption = "Vessel Capacity by Route")

# Does stop_sequence vary within routes?
model_data %>%
  group_by(route) %>%
  summarise(
    min_stop = min(stop_sequence),
    max_stop = max(stop_sequence),
    .groups = "drop"
  ) %>%
  knitr::kable(caption = "Stop Sequence Range by Route")
```

## Feature Selection

Features are selected based on the multicollinearity exploration above and the
48-hour validity constraint. Comment features in or out to test different
combinations.
```{r feature-selection}
model_data <- model_data %>%
  select(
    is_severely_late,
    date,
    dep_hour,
    schedule_season,
    schedule_daytype,
    day_of_week,
    route,
    direction,
    stop_sequence,
    vessel_capacity,
    turn_slack_sched_min,
    #avg_vessel_count,  # Comment this or median_vessel_count out because of correlation
    median_vessel_count # Comment this or avg_vessel_count out because of correlation
  )

glimpse(model_data)
```

## Train/Test Split and Cross Validation

We use a temporal train/test split, holding out June 2025 as the test set.
We use monthly leave-one-out cross validation to prevent data leakage.

```{r train-test-split}
train_data <- model_data %>% filter(date < ymd("2025-06-01"))
test_data  <- model_data %>% filter(date >= ymd("2025-06-01"))

message("Training rows: ", nrow(train_data))
message("Test rows: ", nrow(test_data))

# Confirm class balance in each split
train_data %>% count(is_severely_late) %>% mutate(pct = n / sum(n))
test_data %>% count(is_severely_late) %>% mutate(pct = n / sum(n))
```

```{r cv-folds}
train_data <- train_data %>%
  mutate(cv_month = format(date, "%Y-%m"))

cv_folds <- group_vfold_cv(train_data, group = cv_month, v = 17) # NOTE: We have 17 months of data! So 17 folds are used here.
cv_folds
```

## Modeling

The recipe defines the outcome, predictors, and all preprocessing steps. 
We use `step_downsample` from the `themis` package to address class imbalance by reducing
the majority class during training. 
`date` and `cv_month` are assigned ID roles so they are carried through the pipeline 
without being used as predictors.

**NOTE**: We can tune parameters here, such as under_ratio. Important to tune performance. 

```{r recipe}
model_rec <- recipe(is_severely_late ~ ., data = train_data) %>%
  update_role(date, new_role = "id") %>%
  update_role(cv_month, new_role = "id") %>%
  step_impute_median(turn_slack_sched_min) %>%
  step_dummy(all_nominal_predictors()) %>%
  step_zv(all_predictors()) %>%
  step_normalize(all_numeric_predictors()) %>%
  themis::step_downsample(is_severely_late, under_ratio = 10, seed = 614) #NOTE: Adjust under_ratio here. 1 = 50/50. 10 is 10:1 severe:non-severe
```

## Model Specification and Workflow
```{r logistic-spec}
logistic_spec <- logistic_reg() %>%
  set_engine("glm") %>%
  set_mode("classification")

logistic_wf <- workflow() %>%
  add_recipe(model_rec) %>%
  add_model(logistic_spec)
```

## Cross Validation Fit
```{r cv-fit}
control <- control_resamples(save_pred = TRUE)
metrics <- metric_set(roc_auc, precision, recall, f_meas)

logistic_tuned <- fit_resamples(
  logistic_wf,
  resamples = cv_folds,
  control   = control,
  metrics   = metrics
)

collect_metrics(logistic_tuned)
```

## Threshold Tuning

Given that false positives are operationally costly, we prioritize precision over
recall. 
Rather than using the default 0.5 classification threshold, we evaluate
precision and recall across a range of thresholds using CV predictions.

TODO: Is this a good approach?

Note:
Sensitivity/Recall — of all trips that were actually severely late (true positive), what percentage did the model correctly flag as severely late.

Specificity — of all trips that were actually not severely late (true negative), what percentage did the model correctly identify as not severely late.

```{r threshold-tuning}
p_load(probably)

# Extract CV predictions
cv_predictions <- collect_predictions(logistic_tuned)

# Evaluate across thresholds
threshold_data <- cv_predictions %>%
  threshold_perf(
    truth      = is_severely_late,
    estimate   = .pred_yes,
    thresholds = seq(0.05, 0.95, by = 0.05),
    event_level = "first"
  )

# Plot precision and recall across thresholds
threshold_data %>%
  filter(.metric %in% c("sensitivity", "specificity")) %>%
  ggplot(aes(x = .threshold, y = .estimate, color = .metric)) +
  geom_line() +
  geom_point(size = 1.5) +
  labs(
    title = "Sensitivity and Specificity Across Classification Thresholds",
    subtitle = "Higher threshold = fewer but more reliable alerts",
    x = "Threshold",
    y = "Estimate",
    color = "Metric"
  ) +
  scale_color_viridis_d(option = "turbo") +
  theme_minimal() +
  theme(legend.position = "bottom")
```

## Validate on Test Set
```{r validate}
model_rec_final <- recipe(is_severely_late ~ ., data = train_data %>% select(-cv_month)) %>%
  update_role(date, new_role = "id") %>%
  step_impute_median(turn_slack_sched_min) %>%
  step_dummy(all_nominal_predictors()) %>%
  step_zv(all_predictors()) %>%
  step_normalize(all_numeric_predictors()) %>%
  themis::step_downsample(is_severely_late, under_ratio = 10, seed = 717) # CHANGE UNDER_RATIO HERE FOR VALIDATION AS WELL

logistic_wf_final <- workflow() %>%
  add_recipe(model_rec_final) %>%
  add_model(logistic_spec)

data_split <- make_splits(
  list(
    analysis   = which(model_data$date < ymd("2025-06-01")),
    assessment = which(model_data$date >= ymd("2025-06-01"))
  ),
  model_data
)

logistic_val_fit <- last_fit(logistic_wf_final, split = data_split, metrics = metrics)
collect_metrics(logistic_val_fit)
```

```{r validate-2}
data_split <- make_splits(
  list(
    analysis   = which(model_data$date < ymd("2025-06-01")),
    assessment = which(model_data$date >= ymd("2025-06-01"))
  ),
  model_data
)

logistic_val_fit <- last_fit(logistic_wf_final, split = data_split,
                             metrics = metric_set(roc_auc))

val_predictions <- logistic_val_fit %>%
  collect_predictions() %>%
  mutate(.pred_class = factor(
    if_else(.pred_yes >= 0.25, "yes", "no"),
    levels = c("yes", "no")
  ))

val_predictions %>%
  conf_mat(truth = is_severely_late, estimate = .pred_class)

val_predictions %>%
  precision(truth = is_severely_late, estimate = .pred_class) %>%
  bind_rows(
    val_predictions %>% recall(truth = is_severely_late, estimate = .pred_class),
    val_predictions %>% f_meas(truth = is_severely_late, estimate = .pred_class)
  ) %>%
  knitr::kable(caption = "Validation Metrics at Threshold = 0.25", digits = 3)

val_predictions %>%
  roc_auc(truth = is_severely_late, .pred_yes) %>%
  knitr::kable(caption = "ROC AUC at Validation", digits = 3)
```

## Presentation: Model Performance Summary

These metrics describe how well the model distinguishes severe delays from on-time
trips at the chosen classification threshold of 0.25.

**Sensitivity (Recall)** — Of all trips that were *actually* severely late, what
percentage did the model correctly flag? High sensitivity means fewer missed delays.
A missed delay that wasn't flagged is a false negative.

**Precision** — Of all trips the model *predicted* to be severely late, what
percentage actually were? High precision means fewer false alarms. An unnecessary
alert is a false positive.

**Specificity** — Of all trips that were *actually not* severely late, what
percentage did the model correctly leave unflagged? High specificity means the model
isn't flooding operators with noise.

There is an inherent tradeoff: raising the threshold increases precision and
specificity but reduces sensitivity (the threshold tuning plot above illustrates this).
The 0.25 threshold is chosen to favor catching more delays at the cost of more false
positives.

**ROC AUC** A value of 0.5 is no better than chance; 1.0 is perfect. 

```{r presentation-metrics}
# --- Metric table: sensitivity, precision, specificity at threshold = 0.25 ---
presentation_metrics <- bind_rows(
  val_predictions %>% sensitivity(truth = is_severely_late, estimate = .pred_class),
  val_predictions %>% precision(truth  = is_severely_late, estimate = .pred_class),
  val_predictions %>% specificity(truth = is_severely_late, estimate = .pred_class),
  val_predictions %>% f_meas(truth     = is_severely_late, estimate = .pred_class),
  val_predictions %>% roc_auc(truth    = is_severely_late, .pred_yes)
) %>%
  mutate(
    .metric = recode(.metric,
      sensitivity = "Sensitivity (Recall)",
      precision   = "Precision",
      specificity = "Specificity",
      f_meas      = "F1 Score",
      roc_auc     = "ROC AUC"
    )
  ) %>%
  select(Metric = .metric, Estimate = .estimate)

knitr::kable(presentation_metrics,
             caption = "Validation Performance at Threshold = 0.25",
             digits  = 3)

# --- ROC curve ---
val_predictions %>%
  roc_curve(truth = is_severely_late, .pred_yes, event_level = "first") %>%
  ggplot(aes(x = 1 - specificity, y = sensitivity)) +
  geom_path(color = "#440154", linewidth = 1) +
  geom_abline(lty = 2, color = "gray50") +
  labs(
    title    = "ROC Curve — Validation Set",
    subtitle = paste0("AUC = ",
      round(val_predictions %>%
        roc_auc(truth = is_severely_late, .pred_yes) %>%
        pull(.estimate), 3)),
    x = "False Positive Rate (1 - Specificity)",
    y = "True Positive Rate (Sensitivity)"
  ) +
  theme_minimal()
```
