---
title: "AIS Data Exploration #2"
format: html
author: Sujan Kakumanu
---

```{r setup}
#| message: false
#| warning: false

options(scipen = 999)

if(!require(pacman)){install.packages("pacman"); library(pacman, quietly = T)}
p_load(tidyverse, sf, viridis, lubridate)
```

## Overview

This document builds on the initial AIS exploration with one key change: 
instead of counting **pings** (raw transmissions) per grid cell, we count **unique vessels** 
(distinct MMSIs). This matters because a slow-moving tanker anchored for hours generates 
far more pings than a fast tug crossing the harbor — ping count reflects dwell time, not 
traffic volume. Vessel count is a more honest measure of how busy a given area of the 
harbor actually is.

## Data Loading

Same June 16–22 subset as before. We re-parse the timestamp and retain key columns.
```{r load-data}
ais <- read.csv("AIS_2025_06_16_filtered.csv") %>%
  mutate(
    base_date_time = ymd_hms(base_date_time),
    date = as.Date(base_date_time),
    hour = hour(base_date_time)
  )
```

## Building a Stable Fishnet

Change from previous exploration: we define the fishnet **once** and reuse 
it throughout. We want `grid_id` to refer to the same cell across all dates and time 
windows — a requirement for joining traffic features to our modeling data later.

We also project to **(EPSG:32118)**, a meter-based CRS appropriate for 
New York Harbor. This makes cell sizes interpretable: we use 200m cells for the 
harbor-wide view.

Changes can be made to the cell size for performance and info tradeoffs.

```{r build-fishnet}
ais_sf <- ais %>%
  filter(sog > 0.5) %>%
  drop_na(longitude, latitude) %>%
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326) %>%
  st_transform(32118)

fishnet <- ais_sf %>%
  st_bbox() %>%
  st_as_sfc() %>%
  st_make_grid(cellsize = 200, square = TRUE) %>%
  st_sf() %>%
  mutate(grid_id = row_number())
```

## Aggregating to Vessel Count

As in exploration 1, filter to moving vessels (SOG > 0.5 knots) to exclude anchored or docked ships.
Count **distinct MMSIs** per cell per hour — not raw pings.
```{r aggregate-vessel-count}
traffic <- fishnet %>%
  st_join(ais_sf) %>%
  filter(!is.na(mmsi)) %>%
  group_by(grid_id, date, hour) %>%
  summarise(
    vessel_count = n_distinct(mmsi),
    ping_count = n(),
    .groups = "drop"
  )
```

## Ping Count vs. Vessel Count
This is a quick sanity check to make sure we're on the right path.
Most grid cells have more pings than unique vessels, which is exactly what we expect. 
Vessels passing through a cell generate multiple pings.
```{r ping-vs-vessel}
traffic %>%
  st_drop_geometry() %>%
  group_by(grid_id) %>%
  summarise(
    total_pings = sum(ping_count),
    total_vessels = sum(vessel_count),
    .groups = "drop"
  ) %>%
  filter(total_pings > 0) %>%
  ggplot(aes(x = total_pings, y = total_vessels)) +
  geom_point(alpha = 0.3, color = "steelblue") +
  geom_abline(linetype = "dashed", color = "red") +
  labs(
    title = "Ping Count vs. Vessel Count per Grid Cell",
    x = "Total Pings",
    y = "Total Unique Vessels"
  ) +
  theme_minimal()
```

## Harbor-Wide Vessel Density

Aggregated across the full day to show overall traffic patterns in the harbor.
Note the use of a log scale — vessel counts vary dramatically across cells, and
a log transformation allows us to see variation in both busy and quiet areas.
Log10 is used for better interpretability...

```{r map-harbor-density}
traffic_day <- traffic %>%
  st_drop_geometry() %>%
  group_by(grid_id) %>%
  summarise(vessel_count = sum(vessel_count), .groups = "drop")

fishnet %>%
  left_join(traffic_day, by = "grid_id") %>%
  ggplot() +
  geom_sf(aes(fill = vessel_count), color = NA) +
  scale_fill_viridis_c(option = "inferno", trans = "log10", na.value = "grey90") +
  labs(
    title = "Harbor-Wide Vessel Density (Unique Vessels)",
    subtitle = "NY Harbor — June 16, 2025",
    fill = "Vessel Count (Log10)"
  ) +
  theme_minimal() +
  theme(axis.text = element_blank(), panel.grid = element_blank())
```
Here, we can see some clear patterns that are validated by our domain knowledge. The East River shows a lot of traffic in a tight corridor. Same with the Hudson River, flowing into the area north of the Staten Island.


## Sample Stop-Level Traffic Exploration

Just to explore, I take 3 major ferry stops (hunters point south isn't "major", but just an interesting one):
Question: **how does vessel traffic in the surrounding area vary across the day?**
NOTE: Stops are hard coded here. This is only for exploration purposes.
NOTE 2: We're using 500m buffers to understand patterns further away from stops. The fishnet is still 200m.
```{r stop-traffic-timeseries}
ferry_stops <- tibble(
  stop_name = c("Wall St / Pier 11", "East 34th St", "Hunters Point South"),
  longitude = c(-74.00625, -73.97095, -73.96137),
  latitude  = c(40.70322,  40.74409,  40.74185)
) %>%
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326) %>%
  st_transform(32118)

stop_buffers <- ferry_stops %>%
  mutate(geometry = st_buffer(geometry, dist = 500))

stop_cells <- fishnet %>%
  st_join(stop_buffers) %>%
  filter(!is.na(stop_name)) %>%
  st_drop_geometry() %>%
  select(grid_id, stop_name)

stop_traffic <- traffic %>%
  st_drop_geometry() %>%
  inner_join(stop_cells, by = "grid_id") %>%
  group_by(stop_name, date, hour) %>%
  summarise(vessel_count = sum(vessel_count), .groups = "drop") %>%
  mutate(
    date = as.Date(date),
    datetime = ymd_h(paste(date, hour))
  )

ggplot(stop_traffic, aes(x = datetime, y = vessel_count, color = stop_name)) +
  geom_line() +
  geom_point(size = 1.5) +
  labs(
    title = "Vessel Traffic Near Ferry Stops",
    subtitle = "500m buffer — June 16, 2025",
    x = NULL,
    y = "Unique Vessels per Hour",
    color = "Stop"
  ) +
  scale_color_viridis_d(option = "turbo") +
  theme_minimal() +
  theme(legend.position = "bottom")
```

## In-Water Traffic Along Route Corridors

Beyond stop-level traffic, we want to look at traffic ferries encounter **as they transit between stops**. 
This is particularly valuable for understanding in-water delays.

The route geometry below is a straight-line placeholder between East 34th St and
Hunters Point South. 
NOTE: **This will be replaced with real SWIFTLY route geometry once that data is wrangled.**

```{r route-traffic}
# Placeholder: straight line East 34th St -> Hunters Point South
# To be replaced with SWIFTLY route geometry
route_placeholder <- st_sfc(
  st_linestring(
    matrix(c(-73.97095, 40.74409,
             -73.96137, 40.74185),
           ncol = 2, byrow = TRUE)
  ),
  crs = 4326
) %>%
  st_transform(32118) %>%
  st_sf() %>%
  mutate(route_id = "east_34th_to_hunters_point")

# Buffer 250m along the route corridor
route_buffer <- route_placeholder %>%
  st_buffer(dist = 250)

# Find fishnet cells intersecting the corridor
route_cells <- fishnet %>%
  st_join(route_buffer) %>%
  filter(!is.na(route_id)) %>%
  st_drop_geometry() %>%
  select(grid_id, route_id)

# Vessel count along corridor by hour
route_traffic <- traffic %>%
  st_drop_geometry() %>%
  inner_join(route_cells, by = "grid_id") %>%
  group_by(route_id, date, hour) %>%
  summarise(vessel_count = sum(vessel_count), .groups = "drop") %>%
  mutate(datetime = ymd_h(paste(date, hour)))

ggplot(route_traffic, aes(x = datetime, y = vessel_count)) +
  geom_line(color = "steelblue") +
  geom_point(size = 1.5, color = "steelblue") +
  labs(
    title = "Vessel Traffic Along Route Segment",
    subtitle = "Segment: East 34th St to Hunters Point South — 250m buffer, June 16, 2025",
    x = NULL,
    y = "Unique Vessels per Hour"
  ) +
  theme_minimal()
```


## SAMPLE Modeling Row (Refer to TODOs at end of doc)

To make the feature concrete, the tables below show what joined observations will
look like in the final modeling dataset. The first table shows the stop-level feature,
the second shows the route corridor feature. In practice these will be joined to the
stop-level delay data by stop, route, date, and hour.
```{r sample-modeling-rows}
stop_traffic %>%
  filter(hour %in% c(8, 12, 17)) %>%
  select(stop_name, date, hour, vessel_count) %>%
  arrange(stop_name, hour) %>%
  knitr::kable(caption = "Sample: AIS Traffic Feature at Stop Level")

route_traffic %>%
  filter(hour %in% c(8, 12, 17)) %>%
  select(route_id, date, hour, vessel_count) %>%
  arrange(hour) %>%
  knitr::kable(caption = "Sample: AIS Traffic Feature Along Route Corridor")
```


```{r route-corridor-maps}
# 500m context buffer
crop_extent <- route_placeholder %>%
  st_buffer(500) %>%
  st_bbox()

corridor_traffic <- traffic %>%
  st_drop_geometry() %>%
  inner_join(route_cells, by = "grid_id") %>%
  filter(hour %in% c(8, 12, 17)) %>%
  group_by(grid_id, hour) %>%
  summarise(vessel_count = sum(vessel_count), .groups = "drop") %>%
  left_join(fishnet, by = "grid_id") %>%
  st_as_sf()

ggplot() +
  geom_sf(data = fishnet %>% st_crop(crop_extent), fill = "grey95", color = NA) +
  geom_sf(data = corridor_traffic, aes(fill = vessel_count), color = NA) +
  geom_sf(data = route_placeholder, color = "white", linewidth = 0.8) +
  scale_fill_viridis_c(option = "inferno", na.value = "grey80") +
  coord_sf(xlim = c(crop_extent["xmin"], crop_extent["xmax"]),
           ylim = c(crop_extent["ymin"], crop_extent["ymax"])) +
  facet_wrap(~hour, nrow = 1, labeller = as_labeller(c(
    "8"  = "08:00 AM",
    "12" = "12:00 PM",
    "17" = "05:00 PM"
  ))) +
  labs(
    title = "Vessel Traffic Along Route Corridor by Hour",
    subtitle = "East 34th St to Hunters Point South — 250m buffer, June 16, 2025",
    fill = "Vessel Count"
  ) +
  theme_minimal() +
  theme(axis.text = element_blank(), panel.grid = element_blank())
```
## Team Checkpoint on 3/20/26
### TODOs before Swiftly Integration:
- The second to last section above assumes a straightline path between two terminals as a POC for the feature engineering. Of course, in reality, ferries will take different paths between terminals. Do we have a general route geometry like in GTFS? Do we need to derive a median/typical path from the Swiftly data? In general, I think we need a linestring geometry to understand the traffic impacts.
- For the final traffic feature set, we need to exclude ferry AIS MMSIs. We only care about external traffic.
  - TODO: Identify and exclude NYC Ferry MMSIs from AIS Data before aggregating traffic patterns.
- pct_large_vessels could be good to add
- Explore the lag features more. In this code, we're showing typical traffic at hours of a day. This is better than raw traffic counts.
- SHIFT TO HEXBINS INSTEAD OF FISHNET

### General Notes and Thoughts:
- The above work focuses on traffic at stops and route segments (stop to stop).
- This aggregation might be fine for our predictive model becauyse we don't care about the exact minute a vessel encounters a specific source of delay. We care more about the typical traffic at the time of day, day of week, and season.


## Loading Shared Hexagonal Fishnet

Rather than building our own fishnet, we load a 100m hexbin grid derived from
the SWIFTLY data. This ensures both we can use identical cell IDs
when joining AIS and SWIFTLY features together. The grid covers the NYC Ferry service
area — a subset of the broader NY Harbor AIS extent, which is fine since ferry
operations are our area of interest.
```{r load-hexgrid}
fishnet <- st_read("../hex_grid_100m.geojson") %>%
  st_transform(32118) %>%
  mutate(grid_id = row_number())
```
## Rerunning code with Hexbins

The following code is from earlier in the file, but can be rerun here with the hexbin design.

```{r rerun-with-hexbin}
#| code-fold: true
traffic <- fishnet %>%
  st_join(ais_sf) %>%
  filter(!is.na(mmsi)) %>%
  group_by(grid_id, date, hour) %>%
  summarise(
    vessel_count = n_distinct(mmsi),
    ping_count = n(),
    .groups = "drop"
  )

traffic %>%
  st_drop_geometry() %>%
  group_by(grid_id) %>%
  summarise(
    total_pings = sum(ping_count),
    total_vessels = sum(vessel_count),
    .groups = "drop"
  ) %>%
  filter(total_pings > 0) %>%
  ggplot(aes(x = total_pings, y = total_vessels)) +
  geom_point(alpha = 0.3, color = "steelblue") +
  geom_abline(linetype = "dashed", color = "red") +
  labs(
    title = "Ping Count vs. Vessel Count per Grid Cell",
    x = "Total Pings",
    y = "Total Unique Vessels"
  ) +
  theme_minimal()

traffic_day <- traffic %>%
  st_drop_geometry() %>%
  group_by(grid_id) %>%
  summarise(vessel_count = sum(vessel_count), .groups = "drop")

fishnet %>%
  left_join(traffic_day, by = "grid_id") %>%
  ggplot() +
  geom_sf(aes(fill = vessel_count), color = NA) +
  scale_fill_viridis_c(option = "inferno", trans = "log10", na.value = "grey90") +
  labs(
    title = "Harbor-Wide Vessel Density (Unique Vessels)",
    subtitle = "NY Harbor — June 16, 2025",
    fill = "Vessel Count (Log10)"
  ) +
  theme_minimal() +
  theme(axis.text = element_blank(), panel.grid = element_blank())

ferry_stops <- tibble(
  stop_name = c("Wall St / Pier 11", "East 34th St", "Hunters Point South"),
  longitude = c(-74.00625, -73.97095, -73.96137),
  latitude  = c(40.70322,  40.74409,  40.74185)
) %>%
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326) %>%
  st_transform(32118)

stop_buffers <- ferry_stops %>%
  mutate(geometry = st_buffer(geometry, dist = 500))

stop_cells <- fishnet %>%
  st_join(stop_buffers) %>%
  filter(!is.na(stop_name)) %>%
  st_drop_geometry() %>%
  select(grid_id, stop_name)

stop_traffic <- traffic %>%
  st_drop_geometry() %>%
  inner_join(stop_cells, by = "grid_id") %>%
  group_by(stop_name, date, hour) %>%
  summarise(vessel_count = sum(vessel_count), .groups = "drop") %>%
  mutate(
    date = as.Date(date),
    datetime = ymd_h(paste(date, hour))
  )

ggplot(stop_traffic, aes(x = datetime, y = vessel_count, color = stop_name)) +
  geom_line() +
  geom_point(size = 1.5) +
  labs(
    title = "Vessel Traffic Near Ferry Stops",
    subtitle = "500m buffer — June 16, 2025",
    x = NULL,
    y = "Unique Vessels per Hour",
    color = "Stop"
  ) +
  scale_color_viridis_d(option = "turbo") +
  theme_minimal() +
  theme(legend.position = "bottom")

# Placeholder: straight line East 34th St -> Hunters Point South
# To be replaced with SWIFTLY route geometry
route_placeholder <- st_sfc(
  st_linestring(
    matrix(c(-73.97095, 40.74409,
             -73.96137, 40.74185),
           ncol = 2, byrow = TRUE)
  ),
  crs = 4326
) %>%
  st_transform(32118) %>%
  st_sf() %>%
  mutate(route_id = "east_34th_to_hunters_point")

# Buffer 250m along the route corridor
route_buffer <- route_placeholder %>%
  st_buffer(dist = 250)

# Find fishnet cells intersecting the corridor
route_cells <- fishnet %>%
  st_join(route_buffer) %>%
  filter(!is.na(route_id)) %>%
  st_drop_geometry() %>%
  select(grid_id, route_id)

# Vessel count along corridor by hour
route_traffic <- traffic %>%
  st_drop_geometry() %>%
  inner_join(route_cells, by = "grid_id") %>%
  group_by(route_id, date, hour) %>%
  summarise(vessel_count = sum(vessel_count), .groups = "drop") %>%
  mutate(datetime = ymd_h(paste(date, hour)))

ggplot(route_traffic, aes(x = datetime, y = vessel_count)) +
  geom_line(color = "steelblue") +
  geom_point(size = 1.5, color = "steelblue") +
  labs(
    title = "Vessel Traffic Along Route Segment",
    subtitle = "Segment: East 34th St to Hunters Point South — 250m buffer, June 16, 2025",
    x = NULL,
    y = "Unique Vessels per Hour"
  ) +
  theme_minimal()

# 500m context buffer
crop_extent <- route_placeholder %>%
  st_buffer(500) %>%
  st_bbox()

corridor_traffic <- traffic %>%
  st_drop_geometry() %>%
  inner_join(route_cells, by = "grid_id") %>%
  filter(hour %in% c(8, 12, 17)) %>%
  group_by(grid_id, hour) %>%
  summarise(vessel_count = sum(vessel_count), .groups = "drop") %>%
  left_join(fishnet, by = "grid_id") %>%
  st_as_sf()

ggplot() +
  geom_sf(data = fishnet %>% st_crop(crop_extent), fill = "grey95", color = NA) +
  geom_sf(data = corridor_traffic, aes(fill = vessel_count), color = NA) +
  geom_sf(data = route_placeholder, color = "white", linewidth = 0.8) +
  scale_fill_viridis_c(option = "inferno", na.value = "grey80") +
  coord_sf(xlim = c(crop_extent["xmin"], crop_extent["xmax"]),
           ylim = c(crop_extent["ymin"], crop_extent["ymax"])) +
  facet_wrap(~hour, nrow = 1, labeller = as_labeller(c(
    "8"  = "08:00 AM",
    "12" = "12:00 PM",
    "17" = "05:00 PM"
  ))) +
  labs(
    title = "Vessel Traffic Along Route Corridor by Hour",
    subtitle = "East 34th St to Hunters Point South — 250m buffer, June 16, 2025",
    fill = "Vessel Count"
  ) +
  theme_minimal() +
  theme(axis.text = element_blank(), panel.grid = element_blank())
```