---
title: "AIS Data Exploration"
format: html
author: Sujan Kakumanu
---

```{r setup}
#| message: false
#| warning: false

# Set options
options(scipen = 999)

# Packages
if(!require(pacman)){install.packages("pacman"); library(pacman, quietly = T)}
p_load(tidyverse, sf, viridis, lubridate)
```

The following quarto document begins to explore the AIS dataset. At a high level, AIS (Automatic Identification System) allows large marine vehicles to report their status (location, speed, heading, voyage data, etc). This is used to communicate with other vessels and authorities on-shore (traffic management, for example).

Depending on the size of the vessel, data can be transmitted as frequently as every 2 seconds. As a result, the provided datasets are incredibly large. The analysis code below focuses on one CSV (AIS from June 2025 in NY Harbor). This file is still too large, and will not be uploaded alongside this script.

## Sources:

AIS Data Dictionary: https://coast.noaa.gov/data/marinecadastre/ais/data-dictionary.pdf 
Vessel type codes: https://coast.noaa.gov/data/marinecadastre/ais/VesselTypeCodes2018.pdf 
US Coast Guard AIS Description: https://www.navcen.uscg.gov/automatic-identification-system-overview

## Columns of Interest (grouped by loose categories):
- Vessel features:
  - mmsi: a unique identifying integer
  - vessel_name: name of the ship, seems to include the operator name
  - vessel_type: A code that defines vessel type, cargo type, etc. See link above.
  - length; width: in meters
- Travel features:
  - sog: speed in knots
  - cog: course in degrees (course is the actual direction of the vessel, taking into account wind and current)
  - base_date_time: UTC date time
  - heading: true heading in degrees (the direction of the bow w.r.t true north)
  - latitude; longitude: in decimal degree
  
## Distributions of Select Variables (EDA Part 1)

### Data loading

For this initial exploration, I want to trim the data down to a specific time frame to help with analysis speed. I picked a Wednesday in June to hopefully get a typical working day of data in the harbor.
```{r load-data}
#| eval: false

# Loading the data. Trimming it to Wednesday June 4th for easier analysis at this point.
ais_06_2025 <- read.csv("AIS_2025.06.csv")

ais_06_16_2025_trimmed <- ais_06_2025 %>%
  mutate(base_date_time = ymd_hms(base_date_time)) %>%
  filter(date(base_date_time) == ymd("2025-06-16"))

write.csv(ais_06_16_2025_trimmed, "AIS_2025_06_16_filtered.csv", row.names = FALSE)
```

```{r load-processed-data}
ais_06_16_2025_trimmed <- read.csv("AIS_2025_06_16_filtered.csv")
```

### Frequency of Vessel Types
Below is a visual of how many pings certain vessel types are sending.
*Disclaimer*: We probably need to explore the relation between vessel type and frequency of pings (some vehicles are required to transmit more frequently).

```{r}
ggplot(ais_06_16_2025_trimmed, aes(x = as.factor(vessel_type))) +
  geom_bar(fill = "steelblue") +
  labs(title = "Frequency of Vessel Types",
       subtitle = "June 16, 2025 - NY Harbor",
       x = "Vessel Type Code",
       y = "Number of Pings") +
  theme_minimal()
```
Most Common Vessel codes: 

- 31: Tug and Tow - Towing
- 37: Pleasure craft
- 52: Tug and Tow - Tugging
- 60: All ships of passenger type

### Vessel Size
```{r}
ggplot(ais_06_16_2025_trimmed, aes(x = length)) +
  geom_histogram(binwidth = 10, fill = "steelblue", color = "white") +
  labs(title = "Distribution of Vessel Lengths",
       x = "Length (meters)",
       y = "Count") +
  theme_minimal()
```

### Traffic Volume
```{r}
# Note that I ignore docking/docked ships. Noticed a lot of zeros
ais_06_16_2025_trimmed %>%
  filter(sog > 0.5) %>% 
  ggplot(aes(x = sog)) +
  geom_histogram(binwidth = 1, fill = "darkblue", color = "white") +
  labs(title = "Speed Distribution of Moving Vessels",
       subtitle = "Excludes stationary vessels (SOG > 0.5 knots)",
       x = "Speed Over Ground (knots)",
       y = "Number of Pings") +
  theme_minimal()
```

Next steps:
- Spatial distribution in the harbor
- Fishnet grid over the course of the day?
- Vessel and speed distributions
- June 16th to 22nd

### Spatial Distribution in the Harbor

```{r}
ais_sf <- ais_06_16_2025_trimmed %>%
  drop_na(longitude, latitude) %>%
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326)

# 'cellsize' is in degrees (~0.01 is roughly 1km)
grid <- st_make_grid(ais_sf, cellsize = 0.01, square = TRUE) %>%
  st_sf() %>%
  mutate(grid_id = row_number())

grid_counts <- grid %>%
  st_join(ais_sf) %>%
  group_by(grid_id) %>%
  summarise(ping_count = n()) %>%
  filter(ping_count > 0) # Only keep cells with activity

ggplot(grid_counts) +
  geom_sf(aes(fill = ping_count), color = NA) +
  scale_fill_viridis_c(option = "inferno", trans = "log10") + # Log scale helps see quiet lanes
  labs(title = "AIS Fishnet: Vessel Density",
       subtitle = "NY Harbor - June 16, 2025",
       fill = "Ping Count (Log10)") +
  theme_minimal()
```

```{r}
library(sf)
library(tidyverse)
library(viridis)

# (Lower Manhattan & East River)
manhattan_bbox <- st_bbox(c(xmin = -74.03, xmax = -73.94, ymin = 40.69, ymax = 40.76), 
                          crs = st_crs(4326))

ais_manhattan <- ais_sf %>%
  st_crop(manhattan_bbox)

# 0.001 degrees is roughly 110 meters. 
fine_grid <- st_make_grid(ais_manhattan, cellsize = 0.001, square = TRUE) %>%
  st_sf() %>%
  mutate(grid_id = row_number())

fine_grid_counts <- fine_grid %>%
  st_join(ais_manhattan) %>%
  group_by(grid_id) %>%
  summarise(ping_count = n()) %>%
  filter(ping_count > 0)

ggplot(fine_grid_counts) +
  geom_sf(aes(fill = ping_count), color = NA) +
  scale_fill_viridis_c(option = "magma", trans = "log10") +
  labs(title = "High-Resolution Vessel Density",
       subtitle = "Lower Manhattan & East River (June 16, 2025)",
       fill = "Ping Count (Log10)") +
  theme_minimal() +
  theme(panel.grid = element_blank())
```
## Distributions and Spatial Analysis (EDA Part II)
In preparation for Presentation 1, and to do more spatial analysis, I will focus on the subset of data from June 16th to June 22nd 2025. This is during the more busy summer months, but while school is in session. It also includes a state and city holiday (Juneteenth).

### Data loading

For this initial exploration, I want to trim the data down to a specific time frame to help with analysis speed. I picked a Wednesday in June to hopefully get a typical working day of data in the harbor.
```{r load-data-2}
#| eval: false

# Loading the data. Trimming it to Wednesday June 4th for easier analysis at this point.
ais_06_2025 <- read.csv("AIS_2025.06.csv")

ais_06_16to22_2025_trimmed <- ais_06_2025 %>%
  mutate(base_date_time = ymd_hms(base_date_time)) %>%
  filter(date(base_date_time) >= ymd("2025-06-16") & date(base_date_time) <= ymd("2025-06-22"))

write.csv(ais_06_16to22_2025_trimmed, "AIS_2025_06_16to22_filtered.csv", row.names = FALSE)
```

```{r load-processed-data-2}
ais_06_16to22_2025_trimmed <- read.csv("AIS_2025_06_16to22_filtered.csv")
```

### Fishnet Visualization for Week in June
```{r}
ais_processed <- ais_06_16to22_2025_trimmed %>%
  mutate(
    base_date_time = ymd_hms(base_date_time),
    date = as.Date(base_date_time),
    hour = hour(base_date_time)
  ) %>%
  filter(
    date %in% ymd(c("2025-06-16", "2025-06-19", "2025-06-21")),
    hour %in% c(8, 12, 16, 20)
  ) %>%
  drop_na(longitude, latitude) %>%
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326)


create_daily_fishnet <- function(sf_data, target_date, title_label) {
  day_sf <- sf_data %>% filter(date == ymd(target_date))
  
  grid <- st_make_grid(day_sf, cellsize = 0.005, square = TRUE) %>%
    st_sf() %>%
    mutate(grid_id = row_number())
  
  # Aggregate pings into the grid by hour
  grid_counts <- grid %>%
    st_join(day_sf) %>%
    filter(!is.na(mmsi)) %>%
    group_by(grid_id, hour) %>%
    summarise(ping_count = n(), .groups = "drop")
  
  ggplot(grid_counts) +
    geom_sf(aes(fill = ping_count), color = NA) +
    scale_fill_viridis_c(option = "inferno", trans = "log10") +
    facet_wrap(~hour, labeller = as_labeller(c(
      "8" = "08:00 AM", 
      "12" = "12:00 PM", 
      "16" = "04:00 PM", 
      "20" = "08:00 PM"
    ))) +
    labs(
      title = title_label,
      subtitle = paste("Date:", target_date),
      fill = "Ping Count (Log Scale)",
    ) +
    theme_minimal() +
    theme(
      axis.text = element_blank(),
      panel.grid = element_blank(),
      strip.background = element_rect(fill = "grey90", color = NA)
    )
}

monday_plot <- create_daily_fishnet(ais_processed, "2025-06-16", "Monday Traffic Density")
print(monday_plot)

holiday_plot <- create_daily_fishnet(ais_processed, "2025-06-19", "Juneteenth Holiday Traffic Density")
print(holiday_plot)

saturday_plot <- create_daily_fishnet(ais_processed, "2025-06-21", "Saturday Traffic Density")
print(saturday_plot)
```

```{r}

ais_processed_zoom <- ais_06_16to22_2025_trimmed %>%
  mutate(
    base_date_time = ymd_hms(base_date_time),
    date = as.Date(base_date_time),
    hour = hour(base_date_time)
  ) %>%
  filter(
    date %in% ymd(c("2025-06-16", "2025-06-19", "2025-06-21")),
    hour %in% c(8, 12, 16, 20)
  ) %>%
  drop_na(longitude, latitude) %>%
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326) %>%
  st_crop(st_bbox(c(xmin = -74.03, xmax = -73.94, ymin = 40.69, ymax = 40.76), crs = 4326))


create_zoom_fishnet <- function(sf_data, target_date, title_label) {
  
  day_sf <- sf_data %>% filter(date == ymd(target_date))
  
  grid <- st_make_grid(day_sf, cellsize = 0.001, square = TRUE) %>%
    st_sf() %>%
    mutate(grid_id = row_number())
  
  grid_counts <- grid %>%
    st_join(day_sf) %>%
    filter(!is.na(mmsi)) %>%
    group_by(grid_id, hour) %>%
    summarise(ping_count = n(), .groups = "drop")
  
  ggplot(grid_counts) +
    geom_sf(aes(fill = ping_count), color = NA) +
    scale_fill_viridis_c(option = "inferno", trans = "log10") +
    facet_wrap(~hour, labeller = as_labeller(c(
      "8" = "08:00 AM", 
      "12" = "12:00 PM", 
      "16" = "04:00 PM", 
      "20" = "08:00 PM"
    ))) +
    labs(
      title = title_label,
      subtitle = paste("Lower Manhattan - Date:", target_date),
      fill = "Ping Count (Log Scale)"
    ) +
    theme_minimal() +
    theme(
      axis.text = element_blank(),
      panel.grid = element_blank(),
      strip.background = element_rect(fill = "grey90", color = NA)
    )
}

zoom_monday <- create_zoom_fishnet(ais_processed_zoom, "2025-06-16", "Lower Manhattan: Monday Workday")
print(zoom_monday)

zoom_holiday <- create_zoom_fishnet(ais_processed_zoom, "2025-06-19", "Lower Manhattan: Juneteenth Holiday")
print(zoom_holiday)

zoom_saturday <- create_zoom_fishnet(ais_processed_zoom, "2025-06-21", "Lower Manhattan: Saturday Weekend")
print(zoom_saturday)
```