---
title: "AIS Feature Engineering — Route Segment Corridors"
format: html
author: Sujan Kakumanu
---

Historical AIS traffic averages by route segment, direction, day of week, and hour.
Each segment is a corridor between two consecutive ferry stops along a specific GTFS
shape. Stop areas (500m radius) are subtracted from segment buffers (250m) to isolate
mid-corridor traffic and avoid overlap with the stop-level features in
`AIS_historical_averages.csv`.

```{r setup}
#| message: false
#| warning: false

options(scipen = 999)

if(!require(pacman)){install.packages("pacman"); library(pacman, quietly = T)}
p_load(tidyverse, sf, lwgeom, viridis, lubridate)
```

## File List

```{r list-files}
ais_files <- tibble(
  path = c(
    list.files("AIS/2024", pattern = "^AIS_2024\\.\\d{2}\\.csv$", full.names = TRUE),
    list.files("AIS/2025", pattern = "^AIS_2025\\.\\d{2}\\.csv$", full.names = TRUE)
  )
)

ais_files
```

## Load Shared Fishnet and Landing Buffers

```{r load-spatial}
fishnet <- st_read("../hex_grid_100m.geojson") %>%
  st_transform(32118) %>%
  mutate(grid_id = row_number())

landings <- st_read("../../GeoJSONs/landings.geojson") %>%
  st_transform(32118)

# Union of all 500m stop buffers — used to punch out stop areas from segment corridors
landing_buffers_union <- landings %>%
  distinct(stop_id, .keep_all = TRUE) %>%
  st_buffer(dist = 500) %>%
  st_union()
```

## Load and Prepare Route Segments

```{r load-segments}
# Route name lookup from GTFS shapes
route_names <- st_read("../../GeoJSONs/routes.geojson") %>%
  st_drop_geometry() %>%
  distinct(route_id, route_name = long_name)

# Stop name lookup
stop_lookup <- landings %>%
  st_drop_geometry() %>%
  distinct(stop_id, stop_name)

# Segments: filter to LineStrings only (Points in the file are stop markers)
segments_raw <- st_read("../new_routes_segmentation.geojson") %>%
  filter(st_geometry_type(.) == "LINESTRING") %>%
  st_transform(32118)

# Infer NB/SB direction per shape from the net latitude change of the full shape.
# We union all segments of a shape, transform to WGS84, then compare start vs end lat.
shape_direction <- segments_raw %>%
  group_by(shape_id) %>%
  summarise(geometry = st_union(geometry), .groups = "drop") %>%
  st_transform(4326) %>%
  rowwise() %>%
  mutate(
    coords    = list(st_coordinates(geometry)),
    start_lat = coords[1, "Y"],
    end_lat   = coords[nrow(coords), "Y"],
    direction = if_else(end_lat > start_lat, "NB", "SB")
  ) %>%
  ungroup() %>%
  st_drop_geometry() %>%
  select(shape_id, direction)

# Assemble full segment table with names and direction
segments <- segments_raw %>%
  left_join(route_names,  by = "route_id") %>%
  left_join(stop_lookup,  by = c("from_stop_id" = "stop_id")) %>%
  rename(from_stop_name = stop_name) %>%
  left_join(stop_lookup,  by = c("to_stop_id" = "stop_id")) %>%
  rename(to_stop_name = stop_name) %>%
  left_join(shape_direction, by = "shape_id") %>%
  mutate(
    segment_id = paste(route_id, shape_id, from_stop_id, to_stop_id, sep = "_")
  )

glimpse(segments)
n_distinct(segments$segment_id)
```

## Build Segment Corridor Buffers (250m, Stop Areas Excluded)

```{r segment-buffers}
# 250m buffer around each segment LineString
segment_buffers <- segments %>%
  st_buffer(dist = 250)

# Subtract 500m landing buffers element-wise.
# st_difference.sf filters to only intersecting pairs, so applying it to the whole
# sf at once drops segments that don't overlap a stop. We apply it per-geometry
# via lapply so all 113 rows are preserved — non-overlapping segments keep their
# full buffer unchanged.
masked_geoms <- lapply(st_geometry(segment_buffers), function(g) {
  result <- st_difference(st_sfc(g, crs = 32118), landing_buffers_union)
  if (length(result) == 0 || st_is_empty(result[[1]])) g else result[[1]]
})

segment_buffers_masked <- segment_buffers
st_geometry(segment_buffers_masked) <- do.call(st_sfc, c(masked_geoms, list(crs = 32118)))

# st_difference on touching edges can produce GEOMETRYCOLLECTION; extract polygons.
segment_buffers_masked <- segment_buffers_masked %>%
  filter(!st_is_empty(geometry)) %>%
  mutate(geometry = st_collection_extract(geometry, "POLYGON"))

# Map fishnet cells to segment corridors
segment_cells <- fishnet %>%
  st_join(segment_buffers_masked %>% select(segment_id)) %>%
  filter(!is.na(segment_id)) %>%
  st_drop_geometry() %>%
  select(grid_id, segment_id)

glimpse(segment_cells)
n_distinct(segment_cells$segment_id)
```

## AIS Processing Pipeline

```{r process-function}
# Mirrors the stop-level pipeline in ais-feature-engineering.qmd, but joins to
# segment_cells instead of landing_cells.
process_segments <- function(file_path) {
  message("Processing: ", file_path)

  read.csv(file_path) %>%
    rename_with(tolower) %>%
    rename(
      base_date_time = any_of("basedatetime"),
      latitude       = any_of("lat"),
      longitude      = any_of("lon")
    ) %>%
    mutate(
      base_date_time = ymd_hms(base_date_time, quiet = TRUE),
      date           = as.Date(base_date_time),
      hour           = hour(base_date_time)
    ) %>%
    filter(sog > 0.5) %>%
    drop_na(longitude, latitude, date, hour) %>%
    st_as_sf(coords = c("longitude", "latitude"), crs = 4326) %>%
    st_transform(32118) %>%
    st_join(fishnet) %>%
    st_drop_geometry() %>%
    filter(!is.na(grid_id)) %>%
    group_by(grid_id, date, hour) %>%
    summarise(vessel_count = n_distinct(mmsi), .groups = "drop") %>%
    inner_join(segment_cells, by = "grid_id") %>%
    group_by(segment_id, date, hour) %>%
    summarise(vessel_count = sum(vessel_count), .groups = "drop")
}
```

## Process All AIS Files

Appends to `AIS_segment_features.csv` one month at a time, skipping months already
on disk so the run is resumable.

```{r process-all}
output_path <- "AIS_segment_features.csv"

already_processed <- if (file.exists(output_path)) {
  read.csv(output_path) %>%
    pull(date) %>%
    as.Date() %>%
    format("%Y-%m") %>%
    unique()
} else {
  character(0)
}

ais_files %>%
  pull(path) %>%
  walk(function(f) {
    file_month <- str_extract(f, "\\d{4}\\.\\d{2}") %>%
      str_replace("\\.", "-")

    if (file_month %in% already_processed) {
      message("Skipping already processed: ", f)
      return(invisible(NULL))
    }

    result <- process_segments(f)

    if (!file.exists(output_path)) {
      write.csv(result, output_path, row.names = FALSE)
    } else {
      write.table(result, output_path, sep = ",",
                  row.names = FALSE, col.names = FALSE, append = TRUE)
    }

    message("Completed: ", f)
  })

message("All files processed. Output saved to: ", output_path)
```

```{r sanity-check}
segment_features <- read.csv("AIS_segment_features.csv") %>%
  mutate(date = as.Date(date))

glimpse(segment_features)
nrow(segment_features)
n_distinct(segment_features$segment_id)
range(segment_features$date)
```

## Zero-Fill Missing Hours

Observed-only rows leave gaps where no vessel crossed a segment in a given hour.
We complete the panel with zeros for those hours.

```{r zero-fill}
complete_panel <- segment_features %>%
  distinct(segment_id) %>%
  cross_join(
    segment_features %>%
      distinct(date) %>%
      cross_join(tibble(hour = 0:23))
  )

segment_features_complete <- complete_panel %>%
  left_join(segment_features, by = c("segment_id", "date", "hour")) %>%
  mutate(vessel_count = replace_na(vessel_count, 0))

nrow(segment_features_complete)
summary(segment_features_complete$vessel_count)

write.csv(segment_features_complete, "AIS_segment_features.csv", row.names = FALSE)
```

## Historical Averages

Averages by segment, direction, schedule season/daytype, day of week, and hour.
`direction_id` (0/1 from GTFS) and the computed `direction` (NB/SB) are both
included so this output can join to either GTFS trip keys or the direction label.

```{r historical-averages}
date_lookup <- read.csv("date_lookup.csv") %>%
  mutate(date = as.Date(date))

# Non-spatial segment metadata for joining onto the averages table
segment_meta <- segments %>%
  st_drop_geometry() %>%
  select(segment_id, route_id, route_name, shape_id,
         direction_id, direction,
         from_stop_id, from_stop_name,
         to_stop_id,   to_stop_name)

route_historical_averages <- segment_features_complete %>%
  left_join(date_lookup,    by = "date") %>%
  mutate(day_of_week = wday(date, label = TRUE)) %>%
  left_join(segment_meta,   by = "segment_id") %>%
  group_by(
    route_id, route_name, shape_id,
    direction_id, direction,
    from_stop_id, from_stop_name,
    to_stop_id,   to_stop_name,
    schedule_season, schedule_daytype, day_of_week, hour
  ) %>%
  summarise(
    avg_vessel_count    = mean(vessel_count),
    median_vessel_count = median(vessel_count),
    sd_vessel_count     = sd(vessel_count),
    .groups = "drop"
  )

glimpse(route_historical_averages)
nrow(route_historical_averages)

write.csv(route_historical_averages, "AIS_route_historical_averages.csv", row.names = FALSE)
```

## Visualization

Quick sanity check — average hourly corridor traffic for a few segments on the
East River route, separated by direction.

```{r viz}
er_segments <- route_historical_averages %>%
  filter(route_id == "ER") %>%
  mutate(
    segment_label = paste(from_stop_name, "→", to_stop_name),
    direction     = factor(direction, levels = c("NB", "SB"))
  )

er_segments %>%
  filter(schedule_season == "FALL", schedule_daytype == "WEEKDAY") %>%
  ggplot(aes(x = hour, y = avg_vessel_count, color = segment_label)) +
  geom_line() +
  facet_wrap(~direction, nrow = 1) +
  labs(
    title    = "Average Corridor Traffic — East River Route",
    subtitle = "FALL weekdays, by segment and direction",
    x        = "Hour of Day",
    y        = "Average Unique Vessels",
    color    = "Segment"
  ) +
  scale_color_viridis_d(option = "turbo") +
  theme_minimal() +
  theme(legend.position = "bottom")
```
