---
title: "AIS Feature Engineering"
format: html
author: Sujan Kakumanu
---

```{r setup}
#| message: false
#| warning: false

options(scipen = 999)

if(!require(pacman)){install.packages("pacman"); library(pacman, quietly = T)}
p_load(tidyverse, sf, viridis, lubridate)
```

## Data Loading

Rather than loading all AIS data at once, we process one monthly CSV at a time and
append results to a running panel. This keeps memory manageable given the scale of
the data.
```{r list-files}
ais_files <- tibble(
  path = c(
    list.files("AIS/2024", pattern = "^AIS_2024\\.\\d{2}\\.csv$", full.names = TRUE),
    list.files("AIS/2025", pattern = "^AIS_2025\\.\\d{2}\\.csv$", full.names = TRUE)
  )
)

# Preview file list
ais_files
```

## Load Shared Hexagonal Fishnet and Ferry Landings

```{r load-spatial}
fishnet <- st_read("../hex_grid_100m.geojson") %>%
  st_transform(32118) %>%
  mutate(grid_id = row_number())

landings <- st_read("../swiftly_eda/Shapefiles/Landings/landings.shp") %>%
  st_transform(32118)

# Preview
names(landings)
glimpse(landings)
```

## Prepare Landing Buffers

We deduplicate landings by `stop_id` since some stops serve multiple routes, then
buffer 500m around each landing to capture nearby vessel traffic.
```{r prepare-landings}
landings_clean <- landings %>%
  distinct(stop_id, .keep_all = TRUE)

landing_buffers <- landings_clean %>%
  st_buffer(dist = 500)

# Find which hexbin cells fall within each landing buffer
landing_cells <- fishnet %>%
  st_join(landing_buffers) %>%
  filter(!is.na(stop_id)) %>%
  st_drop_geometry() %>%
  select(grid_id, stop_id, stop_name)
```


## AIS Processing Pipeline: Two Approaches

We test two approaches for aggregating AIS data to the hexbin level. The first uses
`st_join` for spatial intersection.
The second uses H3 arithmetic binning.

### Method 1: Spatial Join

The spatial join approach uses `st_join` to intersect AIS points with the hexgrid.
This works with any hexgrid regardless of how it was built, but is slow on large data
due to the geometric intersection tests required for every point.
```{r method-spatial-join}
test_file <- ais_files$path[1]

process_spatial_join <- function(file_path) {
  message("Processing with spatial join: ", file_path)
  
  read.csv(file_path) %>%
    rename_with(tolower) %>%
    rename(
      base_date_time = any_of("basedatetime"),
      latitude       = any_of("lat"),
      longitude      = any_of("lon")
    ) %>%
    mutate(
      base_date_time = ymd_hms(base_date_time, quiet = TRUE),
      date = as.Date(base_date_time),
      hour = hour(base_date_time)
    ) %>%
    filter(sog > 0.5) %>%
    drop_na(longitude, latitude, date, hour) %>%
    st_as_sf(coords = c("longitude", "latitude"), crs = 4326) %>%
    st_transform(32118) %>%
    st_join(fishnet) %>%
    st_drop_geometry() %>%
    filter(!is.na(grid_id)) %>%
    group_by(grid_id, date, hour) %>%
    summarise(vessel_count = n_distinct(mmsi), .groups = "drop") %>%
    inner_join(landing_cells, by = "grid_id") %>%
    group_by(stop_id, stop_name, date, hour) %>%
    summarise(vessel_count = sum(vessel_count), .groups = "drop")
}

time_spatial <- system.time({
  result_spatial <- process_spatial_join(test_file)
})

message("Spatial join time: ", round(time_spatial["elapsed"], 1), " seconds")
glimpse(result_spatial)
```

### Method 2: H3 Arithmetic Binning

H3 is a hierarchical hexagonal grid system developed by Uber. Instead of geometric
intersection, it assigns coordinates to hex cells using arithmetic — orders of magnitude
faster than spatial joins. This approach requires the hexgrid to be built from H3 cells.
We first check whether our partner's hexgrid aligns with H3 cells at the appropriate
resolution, then use H3 binning if it does.
```{r method-h3-binning}
p_load(h3jsr)
# TODO: Add code here if processing becomes an issue!
```


## Processing All AIS Data

We process all 18 months of AIS data one file at a time, appending results to a
single CSV on disk. This avoids loading all data into memory at once. If the output
file already exists, we skip files that have already been processed — making the
pipeline resumable if it is interrupted.
```{r process-all-ais}
output_path <- "AIS_landing_features.csv"

# Get already processed months from output file if it exists
already_processed <- if (file.exists(output_path)) {
  read.csv(output_path) %>%
    pull(date) %>%
    as.Date() %>%
    format("%Y-%m") %>%
    unique()
} else {
  character(0)
}

# Process each file, skipping already completed months
ais_files %>%
  pull(path) %>%
  walk(function(f) {
    # Extract year-month from filename to check if already processed
    file_month <- str_extract(f, "\\d{4}\\.\\d{2}") %>%
      str_replace("\\.", "-")
    
    if (file_month %in% already_processed) {
      message("Skipping already processed: ", f)
      return(invisible(NULL))
    }
    
    result <- process_spatial_join(f)
    
    if (!file.exists(output_path)) {
      write.csv(result, output_path, row.names = FALSE)
    } else {
      write.table(result, output_path, sep = ",",
                  row.names = FALSE, col.names = FALSE, append = TRUE)
    }
    
    message("Completed: ", f)
  })

message("All files processed. Output saved to: ", output_path)
```

```{r sanity-check}
landing_features <- read.csv("AIS_landing_features.csv") %>%
  mutate(date = as.Date(date))

glimpse(landing_features)
nrow(landing_features)
n_distinct(landing_features$stop_id)
range(landing_features$date)
```

```{r missing-stops}
landings_clean %>%
  st_drop_geometry() %>%
  select(stop_id, stop_name) %>%
  anti_join(landing_features %>% mutate(stop_id = as.character(stop_id)), 
            by = "stop_id")
```

```{r quick-data-check}
# Check for any NAs
landing_features %>%
  summarise(across(everything(), ~sum(is.na(.))))

# Check distribution of vessel counts including zeros
summary(landing_features$vessel_count)

# Check if any zeros exist
landing_features %>%
  filter(vessel_count == 0) %>%
  nrow()
```

## Zero-Filling Missing Hours

Since the pipeline only produces rows where vessels were observed, missing hours
are absent rather than zero. We complete the panel by filling in all combinations
of stop, date, and hour with zeros where no vessels were recorded.
```{r zero-fill}
# Build complete panel of all stop/date/hour combinations
complete_panel <- landing_features %>%
  distinct(stop_id, stop_name) %>%
  cross_join(
    landing_features %>%
      distinct(date) %>%
      cross_join(tibble(hour = 0:23))
  )

landing_features_complete <- complete_panel %>%
  left_join(landing_features, by = c("stop_id", "stop_name", "date", "hour")) %>%
  mutate(vessel_count = replace_na(vessel_count, 0))

nrow(landing_features_complete)
summary(landing_features_complete$vessel_count)

write.csv(landing_features_complete, "AIS_landing_features.csv", row.names = FALSE)
```

## Historical Averages

For the predictive model, we derive historical averages by stop, day of
week, and hour — representing the typical traffic environment a ferry would encounter
at a given stop at a given time.

```{r historical-averages}
historical_averages <- landing_features_complete %>%
  mutate(day_of_week = wday(date, label = TRUE)) %>%
  group_by(stop_id, stop_name, day_of_week, hour) %>%
  summarise(
    avg_vessel_count = mean(vessel_count),
    median_vessel_count = median(vessel_count),
    sd_vessel_count = sd(vessel_count),
    .groups = "drop"
  )

glimpse(historical_averages)
nrow(historical_averages)

write.csv(historical_averages, "AIS_historical_averages.csv", row.names = FALSE)
```

## Visualizing Historical Averages
Picked 3 different landings for a quick visualization.
```{r viz-historical-averages}
stops_to_plot <- c("Hunters Point South", "Wall St/Pier 11", "East 34th Street")

plot_stop_traffic <- function(stop) {
  historical_averages %>%
    filter(stop_name == stop) %>%
    ggplot(aes(x = hour, y = avg_vessel_count, color = day_of_week)) +
    geom_line() +
    labs(
      title = paste("Average Vessel Traffic —", stop),
      subtitle = "Based on 18 months of AIS data (Jan 2024 - Jun 2025)",
      x = "Hour of Day",
      y = "Average Unique Vessels",
      color = "Day of Week"
    ) +
    scale_color_viridis_d(option = "turbo") +
    theme_minimal() +
    theme(legend.position = "bottom")
}

walk(stops_to_plot, ~print(plot_stop_traffic(.x)))
```