---
title: "Analyzing and Predicting Delays for the NYC Ferry System"
subtitle: "2026 MUSA Practicum Project Supporting the NYC Economic Development Corportation"
authors: "Yiming 'Ming' Cao, Sujan Kakumanu, Tyler Maynard, Guangze 'Simon' Sun"   
format:
  html:
    number-sections: true
    code-fold: true
execute:
  cache: true
---
```{r setup}
#| message: false
#| warning: false

# Set options
options(scipen = 999, dplyr.summarise.inform = FALSE)

# Packages
if(!require(pacman)){install.packages("pacman"); library(pacman, quietly = T)}
p_load(tidyverse, sf, viridis, lubridate, kableExtra, gridExtra, ggthemes, ggspatial, leaflet, spatstat, spdep, htmlwidgets, concaveman, patchwork, yardstick, arrow, broom, car, scales)
```

# Use Case



# Exploratory Data Analysis

## Ridership and Trip Data



## Automatic Identification System (AIS) Data
**By: Sujan Kakumanu**

### Background
At a high level, AIS (Automatic Identification System) allows large marine vehicles to report their status (location, speed, heading, voyage data, etc). This is used to communicate with other vessels and authorities on-shore (traffic management, for example).

Depending on the size of the vessel, data can be transmitted as frequently as every 2 seconds. As a result, the provided data sets are incredibly large. The analysis code below focuses on one CSV (AIS from June 2025 in NY Harbor). This file is still too large, and will not be uploaded alongside this script.

**Sources:**

AIS Data Dictionary: https://coast.noaa.gov/data/marinecadastre/ais/data-dictionary.pdf 
Vessel type codes: https://coast.noaa.gov/data/marinecadastre/ais/VesselTypeCodes2018.pdf 
US Coast Guard AIS Description: https://www.navcen.uscg.gov/automatic-identification-system-overview

**Columns of Interest (grouped by loose categories):**

- Vessel features:
  - mmsi: a unique identifying integer
  - vessel_name: name of the ship, seems to include the operator name
  - vessel_type: A code that defines vessel type, cargo type, etc. See link above.
  - length; width: in meters
- Travel features:
  - sog: speed in knots
  - cog: course in degrees (course is the actual direction of the vessel, taking into account wind and current)
  - base_date_time: UTC date time
  - heading: true heading in degrees (the direction of the bow w.r.t true north)
  - latitude; longitude: in decimal degree
  
### Distributions of Select Variables

For this initial exploration, I want to trim the data down to a specific time frame to help with analysis speed. I picked a Wednesday in June to hopefully get a typical working day of data in the harbor.
```{r load-data}
#| eval: false

# Loading the data. Trimming it to Wednesday June 4th for easier analysis at this point.
ais_06_2025 <- read.csv("AIS_2025.06.csv")

ais_06_16_2025_trimmed <- ais_06_2025 %>%
  mutate(base_date_time = ymd_hms(base_date_time)) %>%
  filter(date(base_date_time) == ymd("2025-06-16"))

write.csv(ais_06_16_2025_trimmed, "AIS_2025_06_16_filtered.csv", row.names = FALSE)
```

```{r load-processed-data}
ais_06_16_2025_trimmed <- read.csv("../exploratory_code/ais-eda/AIS_2025_06_16_filtered.csv")
```

**Frequency of Vessel Types**
Below is a visual of how many pings certain vessel types are sending.

```{r}
ggplot(ais_06_16_2025_trimmed, aes(x = as.factor(vessel_type))) +
  geom_bar(fill = "steelblue") +
  labs(title = "Frequency of Vessel Types",
       subtitle = "June 16, 2025 - NY Harbor",
       x = "Vessel Type Code",
       y = "Number of Pings") +
  theme_minimal()
```
Most Common Vessel codes: 

- 31: Tug and Tow - Towing
- 37: Pleasure craft
- 52: Tug and Tow - Tugging
- 60: All ships of passenger type

**Vessel Size**
```{r}
ggplot(ais_06_16_2025_trimmed, aes(x = length)) +
  geom_histogram(binwidth = 10, fill = "steelblue", color = "white") +
  labs(title = "Distribution of Vessel Lengths",
       x = "Length (meters)",
       y = "Count") +
  theme_minimal()
```

**Traffic Volume**
```{r}
# Note that I ignore docking/docked ships. Noticed a lot of zeros
ais_06_16_2025_trimmed %>%
  filter(sog > 0.5) %>% 
  ggplot(aes(x = sog)) +
  geom_histogram(binwidth = 1, fill = "darkblue", color = "white") +
  labs(title = "Speed Distribution of Moving Vessels",
       subtitle = "Excludes stationary vessels (SOG > 0.5 knots)",
       x = "Speed Over Ground (knots)",
       y = "Number of Pings") +
  theme_minimal()
```
### Hexbin Visualizations
We can aggregate traffic to hexbin cells to understand traffic patterns.
Instead of counting **pings** (raw transmissions) per hexbin, we count **unique vessels** 
(distinct MMSIs). This matters because a slow-moving tanker anchored for hours generates 
far more pings than a fast tug crossing the harbor — ping count reflects dwell time, not 
traffic volume. Vessel count is a more honest measure of how busy a given area of the 
harbor actually is.

**Data Loading**
Same June 16–22 subset as before.
```{r load-data-2}
ais <- read.csv("../exploratory_code/ais-eda/AIS_2025_06_16_filtered.csv") %>%
  mutate(
    base_date_time = ymd_hms(base_date_time),
    date = as.Date(base_date_time),
    hour = hour(base_date_time)
  )
```
We created a geojson containing a stable hexbin grid to be used for our project.

```{r build-fishnet}
ais_sf <- ais %>%
  filter(sog > 0.5) %>%
  drop_na(longitude, latitude) %>%
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326) %>%
  st_transform(32118)

fishnet <- st_read("../exploratory_code/hex_grid_100m.geojson") %>%
  st_transform(32118) %>%
  mutate(grid_id = row_number())
```

**Aggregating to Vessel Count**

We filter to moving vessels (SOG > 0.5 knots) to exclude anchored or docked ships.
Count **distinct MMSIs** per cell per hour — not raw pings.
```{r aggregate-vessel-count}
traffic <- fishnet %>%
  st_join(ais_sf) %>%
  filter(!is.na(mmsi)) %>%
  group_by(grid_id, date, hour) %>%
  summarise(
    vessel_count = n_distinct(mmsi),
    ping_count = n(),
    .groups = "drop"
  )
```

**Ping Count vs. Vessel Count**
This is a quick sanity check to make sure we're on the right path.
Most hexbin cells have more pings than unique vessels, which is exactly what we expect. 
Vessels passing through a cell generate multiple pings.
```{r ping-vs-vessel}
traffic %>%
  st_drop_geometry() %>%
  group_by(grid_id) %>%
  summarise(
    total_pings = sum(ping_count),
    total_vessels = sum(vessel_count),
    .groups = "drop"
  ) %>%
  filter(total_pings > 0) %>%
  ggplot(aes(x = total_pings, y = total_vessels)) +
  geom_point(alpha = 0.3, color = "steelblue") +
  geom_abline(linetype = "dashed", color = "red") +
  labs(
    title = "Ping Count vs. Vessel Count per Hexbin Cell",
    x = "Total Pings",
    y = "Total Unique Vessels"
  ) +
  theme_minimal()
```

#### Harbor-Wide Vessel Density
Aggregated across the full day to show overall traffic patterns in the harbor.
Note the use of a log scale — vessel counts vary dramatically across cells, and
a log transformation allows us to see variation in both busy and quiet areas.
Log10 is used for better interpretability.

```{r map-harbor-density}
traffic_day <- traffic %>%
  st_drop_geometry() %>%
  group_by(grid_id) %>%
  summarise(vessel_count = sum(vessel_count), .groups = "drop")

fishnet %>%
  left_join(traffic_day, by = "grid_id") %>%
  ggplot() +
  geom_sf(aes(fill = vessel_count), color = NA) +
  scale_fill_viridis_c(option = "inferno", trans = "log10", na.value = "grey90") +
  labs(
    title = "Harbor-Wide Vessel Density (Unique Vessels)",
    subtitle = "NY Harbor — June 16, 2025",
    fill = "Vessel Count (Log10)"
  ) +
  theme_minimal() +
  theme(axis.text = element_blank(), panel.grid = element_blank())
```
Here, we can see some clear patterns that are validated by our domain knowledge. The East River shows a lot of traffic in a tight corridor. Same with the Hudson River, flowing into the area north of the Staten Island.


#### Sample Stop-Level Traffic Exploration
Just to explore, I take 3 major ferry stops (hunters point south isn't "major", but just an interesting one):
Question: **how does vessel traffic in the surrounding area vary across the day?**
NOTE: Stops are hard coded here. This is only for exploration purposes.
NOTE 2: We're using 500m buffers to understand patterns further away from stops. The fishnet is still 200m.
```{r stop-traffic-timeseries}
ferry_stops <- tibble(
  stop_name = c("Wall St / Pier 11", "East 34th St", "Hunters Point South"),
  longitude = c(-74.00625, -73.97095, -73.96137),
  latitude  = c(40.70322,  40.74409,  40.74185)
) %>%
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326) %>%
  st_transform(32118)

stop_buffers <- ferry_stops %>%
  mutate(geometry = st_buffer(geometry, dist = 500))

stop_cells <- fishnet %>%
  st_join(stop_buffers) %>%
  filter(!is.na(stop_name)) %>%
  st_drop_geometry() %>%
  select(grid_id, stop_name)

stop_traffic <- traffic %>%
  st_drop_geometry() %>%
  inner_join(stop_cells, by = "grid_id") %>%
  group_by(stop_name, date, hour) %>%
  summarise(vessel_count = sum(vessel_count), .groups = "drop") %>%
  mutate(
    date = as.Date(date),
    datetime = ymd_h(paste(date, hour))
  )

ggplot(stop_traffic, aes(x = datetime, y = vessel_count, color = stop_name)) +
  geom_line() +
  geom_point(size = 1.5) +
  labs(
    title = "Vessel Traffic Near Ferry Stops",
    subtitle = "500m buffer — June 16, 2025",
    x = NULL,
    y = "Unique Vessels per Hour",
    color = "Stop"
  ) +
  scale_color_viridis_d(option = "turbo") +
  theme_minimal() +
  theme(legend.position = "bottom")
```

#### In-Water Traffic Along Route Corridors

Beyond stop-level traffic, we want to look at traffic ferries encounter **as they transit between stops**. 
This is particularly valuable for understanding in-water delays.

The route geometry below is a straight-line placeholder between East 34th St and
Hunters Point South. 
NOTE: **This will be replaced with real SWIFTLY route geometry once that data is wrangled.**

```{r route-traffic}
# Placeholder: straight line East 34th St -> Hunters Point South
# To be replaced with SWIFTLY route geometry
route_placeholder <- st_sfc(
  st_linestring(
    matrix(c(-73.97095, 40.74409,
             -73.96137, 40.74185),
           ncol = 2, byrow = TRUE)
  ),
  crs = 4326
) %>%
  st_transform(32118) %>%
  st_sf() %>%
  mutate(route_id = "east_34th_to_hunters_point")

# Buffer 250m along the route corridor
route_buffer <- route_placeholder %>%
  st_buffer(dist = 250)

# Find fishnet cells intersecting the corridor
route_cells <- fishnet %>%
  st_join(route_buffer) %>%
  filter(!is.na(route_id)) %>%
  st_drop_geometry() %>%
  select(grid_id, route_id)

# Vessel count along corridor by hour
route_traffic <- traffic %>%
  st_drop_geometry() %>%
  inner_join(route_cells, by = "grid_id") %>%
  group_by(route_id, date, hour) %>%
  summarise(vessel_count = sum(vessel_count), .groups = "drop") %>%
  mutate(datetime = ymd_h(paste(date, hour)))

ggplot(route_traffic, aes(x = datetime, y = vessel_count)) +
  geom_line(color = "steelblue") +
  geom_point(size = 1.5, color = "steelblue") +
  labs(
    title = "Vessel Traffic Along Route Segment",
    subtitle = "Segment: East 34th St to Hunters Point South — 250m buffer, June 16, 2025",
    x = NULL,
    y = "Unique Vessels per Hour"
  ) +
  theme_minimal()
```


#### Sample Modeling Row

To make the feature concrete, the tables below show what joined observations will
look like in the final modeling dataset. The first table shows the stop-level feature,
the second shows the route corridor feature. In practice these will be joined to the
stop-level delay data by stop, route, date, and hour.
```{r sample-modeling-rows}
stop_traffic %>%
  filter(hour %in% c(8, 12, 17)) %>%
  select(stop_name, date, hour, vessel_count) %>%
  arrange(stop_name, hour) %>%
  knitr::kable(caption = "Sample: AIS Traffic Feature at Stop Level")

route_traffic %>%
  filter(hour %in% c(8, 12, 17)) %>%
  select(route_id, date, hour, vessel_count) %>%
  arrange(hour) %>%
  knitr::kable(caption = "Sample: AIS Traffic Feature Along Route Corridor")
```


```{r route-corridor-maps}
# 500m context buffer
crop_extent <- route_placeholder %>%
  st_buffer(500) %>%
  st_bbox()

corridor_traffic <- traffic %>%
  st_drop_geometry() %>%
  inner_join(route_cells, by = "grid_id") %>%
  filter(hour %in% c(8, 12, 17)) %>%
  group_by(grid_id, hour) %>%
  summarise(vessel_count = sum(vessel_count), .groups = "drop") %>%
  left_join(fishnet, by = "grid_id") %>%
  st_as_sf()

ggplot() +
  geom_sf(data = fishnet %>% st_crop(crop_extent), fill = "grey95", color = NA) +
  geom_sf(data = corridor_traffic, aes(fill = vessel_count), color = NA) +
  geom_sf(data = route_placeholder, color = "white", linewidth = 0.8) +
  scale_fill_viridis_c(option = "inferno", na.value = "grey80") +
  coord_sf(xlim = c(crop_extent["xmin"], crop_extent["xmax"]),
           ylim = c(crop_extent["ymin"], crop_extent["ymax"])) +
  facet_wrap(~hour, nrow = 1, labeller = as_labeller(c(
    "8"  = "08:00 AM",
    "12" = "12:00 PM",
    "17" = "05:00 PM"
  ))) +
  labs(
    title = "Vessel Traffic Along Route Corridor by Hour",
    subtitle = "East 34th St to Hunters Point South — 250m buffer, June 16, 2025",
    fill = "Vessel Count"
  ) +
  theme_minimal() +
  theme(axis.text = element_blank(), panel.grid = element_blank())
```

## Swiftly
**By: Tyler Maynard**

### Background

The Swiftly GPS dataset contains location data every 10-60 seconds from the NYC Ferry AVL system. Columns include vessel, route, trip, block, and coordinates. This will be especially useful in calculating the speed of the vehicle at various points in time, helping identify locations, routes, and trips where the system struggles with on-time performance.

**Reference Week: June 16-22, 2025**

For the sake of processing power, we selected a "sample week" to conduct EDA for Swiftly GPS and AIS data. Based on conversations with NYCEDC, ridership is much higher in the summer months, especially on weekends. Thus, we chose a week during the summer, also including a holiday to evaluate any unique characteristics (if any) during special service days.

```{r swiftly-special-service}
#| warning: false
#| message: false

special_service <- data.frame(
  dates = as.Date(c("2025-06-19")))
```

### Data Import

The Swiftly GPS data set is available by month in CSV files. The following chunk reads the June 2025 CSV file, filters to the dates included in the sample week, and parses out each record's date and time into separate columns.

```{r swiftly-data-import}
#| warning: false
#| message: false

# Original csv
jun2025_original <- read.csv("../exploratory_code/swiftly_eda/swiftly_gps.2025_06.csv")

# Removing extra 'data.' from column names, removing unneeded columns
jun2025 <- jun2025_original %>%
  rename_with(~ gsub("data.", "", .x)) %>%
  select(-(1:3), -(7))

# Converting to datetime
jun2025$time <- ymd_hms(jun2025$time)
jun2025$timeProcessed <- ymd_hms(jun2025$timeProcessed)

# Seeing how many unique values each column has
jun2025 %>%
  summarise(across(everything(), n_distinct)) %>%
  pivot_longer(cols = everything()) %>%
  kable()

# Parse dates and times
jun2025 <- jun2025 %>%
  mutate(
    date_time = ymd_hms(time, tz = "America/New_York"),
    processed_date_time = ymd_hms(timeProcessed, tz = "America/New_York"),

    date = as.Date(date_time),
    time = format(date_time, "%H:%M:%S"),
    hour = substr(time,1,2),

    dateProcessed = as.Date(processed_date_time),
    timeProcessed = format(processed_date_time, "%H:%M:%S"),

    dayOfWeek = if_else(wday(date, week_start = 1) %in% c(6,7), "Weekend", "Weekday"),
    dayOfWeek = case_when(
      date %in% special_service$dates ~ "Holiday",
      TRUE ~ dayOfWeek)) %>%
  select(date, time, dateProcessed, timeProcessed, dayOfWeek, headsign, tripShortName, everything(), date_time, processed_date_time)

jun2025_reference_week <- jun2025 %>%
  filter(date %in% c("2025-06-16","2025-06-17","2025-06-18","2025-06-19","2025-06-20","2025-06-21","2025-06-22")) %>%
  filter(!is.na(tripId)) %>%
  arrange(vehicleId, date_time)
str(jun2025_reference_week$time)
```

### Shapefiles

As a way to add more spatial context to this analysis, the landings, routes, hydrology, and boroughs were extracted from the NYC Ferry and NYC OpenData websites. The hydrology object was trimmed to only include hydrology contained within the bounding box created around the study area.

```{r swiftly-sf-objects}
#| warning: false
#| message: false

landings <- read_sf("../exploratory_code/swiftly_eda/Shapefiles/Landings/landings.shp") %>%
  st_transform(32118)

landings_east_river <- landings %>%
  filter(!stop_id %in% c("37","39","43","45","46","48","49","50",
    "51","52","55","56","57","58","61","62",
    "16","88","103","104","105","142",
    "145","149","150","151")) %>%
  st_transform(32118)

routes <- read_sf("../exploratory_code/swiftly_eda/Shapefiles/Routes/routes.shp") %>%
  group_by(route_id) %>%
  summarise() %>%
  st_transform(32118)

boroughs <- read_sf("../exploratory_code/swiftly_eda/Shapefiles/Boroughs/boroughs.shp") %>%
  st_transform(32118)

bbox_east_river <- st_bbox(landings_east_river)
bbox_expanded_east_river <- bbox_east_river
bbox_expanded_east_river["xmin"] <- bbox_east_river["xmin"] - 15000
bbox_expanded_east_river["ymin"] <- bbox_east_river["ymin"] - 15000
bbox_expanded_east_river["xmax"] <- bbox_east_river["xmax"] + 15000
bbox_expanded_east_river["ymax"] <- bbox_east_river["ymax"] + 15000
bbox_poly_east_river <- st_as_sfc(bbox_expanded_east_river)

water <- read_sf("../exploratory_code/swiftly_eda/Shapefiles/Hydrology/WBDHU2.shp") %>%
  st_transform(32118) %>%
  summarise(geometry = st_union(geometry)) %>%
  st_intersection(st_transform(st_as_sfc(st_bbox(bbox_expanded_east_river)), 32118))
```

```{r swiftly-bounding-boxes}
#| warning: false
#| message: false

bbox <- st_bbox(c(
  xmin = -74.30603027343751,
  ymin = 40.492915026895815,
  xmax = -73.64959716796876,
  ymax = 40.919739051062216
), crs = 4326)

bbox_area <- st_as_sfc(bbox) %>%
  st_transform(32118)

bbox_service_area <- boroughs %>%
  st_transform(32118) %>%
  st_bbox()
```

### Data Cleaning

The following chunk creates sf objects for each data point and computes new fields for feature engineering. Features created include:

- **Knots**: based on the time and distance of the previous node of the same trip.
- **Key identifiers**: for different analyses, generally including date, trip, and route.
- **OTP**: classifies on-time performance at each node based on the schedule adherence column (`schedAdhMsec`).
- **Excessively Late**: a binary field indicating whether a record is more than 10 minutes past the expected time.
- **First Record Late**: indicates if the trip starts five minutes past its scheduled time.
- **OTP Transition**: identifies trips where the OTP changes from one status to another (e.g., on-time to late).

Of the 851,238 data points within this sample week, 25,558 were identified as excessively late — approximately three percent of all records.

```{r swiftly-cleaning}
#| warning: false
#| message: false

pts <- st_as_sf(jun2025_reference_week, coords = c("lon", "lat"), crs = 4326, remove = FALSE) %>%
  st_transform(32118)

xy <- st_coordinates(pts)

stopifnot(nrow(jun2025_reference_week) == nrow(xy))

jun2025_reference_week <- jun2025_reference_week %>%
  mutate(x = xy[, 1], y = xy[, 2]) %>%
  group_by(date, blockId, routeId, tripId, directionId, dayOfWeek) %>%
  arrange(date_time, .by_group = TRUE) %>%
  mutate(
    directionId = case_when(
      directionId == 1 ~ "NB",
      TRUE ~ "SB"
    ),
    isFirst = row_number() == 1,
    isSecond = row_number() == 2,
    dist_to_prev_m = round(sqrt((x - lag(x))^2 + (y - lag(y))^2), 2),
    time_diff_sec  = as.numeric(difftime(date_time, lag(date_time), units = "secs")),
    m_per_sec      = if_else(is.na(time_diff_sec) | time_diff_sec <= 0,
                             0,
                             round(dist_to_prev_m / time_diff_sec, 2)),
    knots = m_per_sec * 1.94384,
    date_route_trip_direction_vehicle_dow = paste(date, routeId, tripId, directionId, vehicleId, dayOfWeek, sep = " "),
    key_identifier = paste(date, blockId, routeId, tripId, directionId, dayOfWeek, sep = " "),
    dailyTrip = paste(dayOfWeek, routeId, tripId, sep = " "),
    countDailyTrip = n(),
    OTP = case_when(
      schedAdhMsec < -60000 ~ "Early",
      schedAdhMsec > 600000 ~ "Excessively Late",
      schedAdhMsec > 300000 ~ "Late",
      TRUE ~ "Ontime"
    ),
    excessivelyLate = case_when(
      OTP == "Excessively Late" ~ 1,
      TRUE ~ 0
    ),
    timeOfDay = as.integer(hour(date_time)),
    firstRecordLate = if_else(isSecond == 1 & (OTP == "Late" | OTP == "Excessively Late"), "Yes", "No"),
    OTP_prev = lag(OTP),
    OTP_changed = coalesce(OTP != OTP_prev, FALSE),
    OTP_transition = case_when(
      isFirst ~ "Start",
      is.na(OTP_prev) ~ "Start",
      OTP == OTP_prev ~ "No Change",
      TRUE ~ paste0(OTP_prev, " -> ", OTP)
    )
  ) %>%
  ungroup() %>%
  select(-x, -y, -m_per_sec) %>%
  select(dailyTrip, key_identifier, date_route_trip_direction_vehicle_dow, date, time, dateProcessed, timeProcessed, OTP, OTP_transition, excessivelyLate, timeOfDay, schedAdh, dayOfWeek,
         dist_to_prev_m, time_diff_sec, knots, headsign, tripShortName,
         everything(), date_time, processed_date_time)

summary_table <- data.frame(
  Statistic = names(summary(jun2025_reference_week$knots)),
  Value     = as.numeric(summary(jun2025_reference_week$knots)),
  row.names = NULL
)

jun2025_reference_week_sf <- st_as_sf(jun2025_reference_week, coords = c("lon", "lat"), crs = 4326, remove = FALSE) %>%
  st_transform(32118)

kable(summary_table, digits = 2, caption = "Summary of Knots") %>%
  kable_styling(
    bootstrap_options = c("striped", "hover"),
    full_width = FALSE
  ) %>%
  column_spec(1:2, width = "150px")

jun2025_reference_week %>%
  count(excessivelyLate, sort = TRUE)

jun2025_reference_week_excessively_lates <- jun2025_reference_week %>%
  filter(OTP == "Excessively Late")

jun2025_reference_week_excessively_lates_sf <- st_as_sf(jun2025_reference_week_excessively_lates, coords = c("lon", "lat"), crs = 4326, remove = FALSE) %>%
  st_transform(32118)

hull <- concaveman(jun2025_reference_week_excessively_lates_sf)
bbox_hull <- st_bbox(hull)

ggplot() +
  geom_sf(data = boroughs, color = "white") +
  geom_sf(data = hull, color = "red", fill = NA, size = 3) +
  coord_sf(
    xlim = c(bbox_east_river["xmin"] - 10000, bbox_east_river["xmax"] + 10000),
    ylim = c(bbox_east_river["ymin"] - 10000, bbox_east_river["ymax"] + 10000),
    expand = FALSE
  ) +
  labs(title = "Excessively Late 'Zone' - Concave Polygon") +
  theme_solid() +
  theme(
    plot.title = element_text(size = 20),
    plot.title.position = "plot"
  )

jun2025_reference_week %>%
  count(routeId)
```

### OTP by Direction

This chunk computes on-time performance by direction (NB or SB), filtering to excessively late records. Northbound and southbound trips comprise roughly the same share of all trips; however, the higher number of excessively late southbound data points could be indicative of other issues.

```{r swiftly-otp-direction}
#| warning: false
#| message: false

jun2025_reference_week_dir <- jun2025_reference_week %>%
  select(dailyTrip, directionId, OTP, routeId) %>%
  filter(OTP == "Excessively Late")

jun2025_reference_week_dir %>%
  count(directionId)

jun2025_reference_week_dir %>%
  count(directionId, routeId)
```

### Bar Charts

The bar charts below visualize key engineered features. The first chart shows the distribution of excessively late trips by the time spent excessively late (as a percentage of total data points per trip).

Three key takeaways from the second plot:

1. Even though weekdays account for the majority of data points, they account for under half of excessively late data points. Weekend service is the only category where excessively late counts surpass non-excessively late counts.
2. Nearly 80 percent of excessively late data points are attributed to the East River and Rockaway routes.
3. The PM peak period (approximately 1pm to 10pm) accounts for the overwhelming majority of problematic data points.

```{r swiftly-bar-charts}
#| warning: false
#| message: false

excessivelyLatePlotVariables <-
  jun2025_reference_week %>%
  as.data.frame() %>%
  select(excessivelyLate, dayOfWeek, routeId, timeOfDay, blockId, directionId, firstRecordLate) %>%
  gather(variable, value, -excessivelyLate) %>%
  filter(!is.na(value))

jun2025_reference_week_wide <- jun2025_reference_week %>%
  count(key_identifier, OTP) %>%
  pivot_wider(
    names_from = OTP,
    values_from = n,
    values_fill = 0
  ) %>%
  mutate(Total = rowSums(across(-key_identifier)))

jun2025_reference_week_wide

jun2025_reference_week_percents <- jun2025_reference_week_wide %>%
  mutate(round(across(-c(key_identifier, Total), ~ . / Total * 100), 2)) %>%
  pivot_longer(
    cols = -c(key_identifier, Total),
    names_to = "OTP",
    values_to = "Percent"
  ) %>%
  filter(OTP == "Excessively Late", Percent > 0) %>%
  mutate(Percent_Rounded = round(Percent, 0))

mean_val <- mean(jun2025_reference_week_percents$Percent_Rounded)
median_val <- median(jun2025_reference_week_percents$Percent_Rounded)

ggplot() +
  geom_histogram(data = jun2025_reference_week_percents, color = "white", fill = "#006bb6", linewidth = 0.3, aes(x = Percent_Rounded), binwidth = 10) +
  geom_vline(xintercept = mean_val, color = "red2", linetype = "dashed", linewidth = 1) +
  geom_vline(xintercept = median_val, color = "darkgreen", linetype = "dashed", linewidth = 1) +
  annotate("text", x = mean_val, y = 15, label = "Mean", color = "red2", angle = 90, vjust = -0.5) +
  annotate("text", x = median_val, y = 15, label = "Median", color = "darkgreen", angle = 90, vjust = -0.5) +
  labs(
    title = "Distribution of Excessively Late Trips by Time Spent Excessively Late",
    x = "Percent",
    y = "Count"
  ) +
  theme_excel_new() +
  theme(
    axis.title = element_text(size = 14),
    plot.title = element_text(size = 20),
    plot.subtitle = element_text(size = 16),
    plot.title.position = "plot"
  )

plot_df <- excessivelyLatePlotVariables %>%
  group_by(excessivelyLate, variable, value) %>%
  summarise(count = n(), .groups = "drop") %>%
  group_by(excessivelyLate, variable) %>%
  mutate(prop = count / sum(count)) %>%
  ungroup()

ggplot(plot_df, aes(x = ifelse(variable == "timeOfDay", as.numeric(value), value),
                    y = prop, fill = as.factor(excessivelyLate))) +
  geom_col(position = "dodge") +
  facet_wrap(~ variable, scales = "free_x") +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_manual(values = c("0" = "limegreen", "1" = "darkred"),
                    labels = c("Not Excessively Late", "Excessively Late"),
                    name = "") +
  labs(x = NULL, y = "Percent within group") +
  theme_excel_new()
```

```{r swiftly-bar-chart-by-route}
#| warning: false
#| message: false

jun2025_reference_week <- jun2025_reference_week %>%
  mutate(day_route = paste(dayOfWeek, routeId, sep = " "))

speed_by_route <- jun2025_reference_week %>%
  group_by(day_route, dayOfWeek, routeId) %>%
  summarise(
    avgKnots = mean(knots, na.rm = TRUE),
    medKnots = median(knots, na.rm = TRUE),
    n = dplyr::n(),
    .groups = "drop"
  )

ggplot() +
  geom_col(data = speed_by_route, aes(x = day_route, y = avgKnots, fill = dayOfWeek)) +
  labs(
    title = "Average Speed by Route and Service Level",
    x = "Service Level",
    y = "Average Knots"
  ) +
  scale_fill_manual(
    values = c(
      "Holiday" = "limegreen",
      "Weekday" = "#006bb6",
      "Weekend" = "darkred"
    )
  ) +
  theme_excel_new() +
  theme(
    plot.title = element_text(size = 20),
    plot.subtitle = element_text(size = 16),
    plot.title.position = "plot",
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```

### Landings Map

```{r swiftly-landings-map}
#| warning: false
#| message: false

ggplot() +
  geom_sf(data = boroughs, color = "white") +
  geom_sf(data = landings, color = "black", size = 3) +
  labs(title = "NYC Ferry Landings") +
  theme_solid() +
  theme(
    plot.title = element_text(size = 16),
    plot.title.position = "plot"
  )
```

### Fishnet and Kernel Density

Since these datasets contain thousands of points per day, aggregation is the most efficient way to process and observe the data. A fishnet was created using a grid cell side of 250 meters. The figure represents the data for the sample week, excluding any grid cell with a density of 0.

```{r swiftly-fishnet}
#| dev: "ragg_png"
#| dpi: 300
#| warning: false
#| message: false

fishnet_swiftly <-
  st_make_grid(water, cellsize = 250) %>%
  st_sf() %>%
  mutate(uniqueID = rownames(.))

jun2025_net <- st_join(
  fishnet_swiftly, jun2025_reference_week_excessively_lates_sf, join = st_intersects) %>%
  group_by(uniqueID) %>%
  summarise(mean_speed = mean(knots, na.rm = TRUE),
            n = n(), .groups = "drop") %>%
  filter(!is.na(mean_speed))

jun2025_ppp <- as.ppp(st_coordinates(jun2025_reference_week_excessively_lates_sf), W = st_bbox(jun2025_net))
jun2025_KD <- density.ppp(jun2025_ppp, sigma = 1000)

as.data.frame(jun2025_KD) %>%
  st_as_sf(coords = c("x", "y"), crs = st_crs(jun2025_net)) %>%
  aggregate(., jun2025_net, mean) %>%
  ggplot() +
    geom_sf(aes(fill = value), color = NA, linewidth = 0) +
    geom_sf(data = landings, color = "black", size = 1, alpha = 0.5) +
    scale_fill_viridis(name = "Density") +
    labs(title = "Kernel Density of Excessively Late Data Points, 6/16/25 to 6/22/25") +
    theme_solid() +
    theme(
      plot.title = element_text(size = 16),
      plot.title.position = "plot"
    )
```

### Mean Speed

The following figure shows the average speed of each data point in knots per grid cell, regardless of on-time performance designation.

```{r swiftly-mean-speed}
#| warning: false
#| message: false

ggplot(jun2025_net) +
  geom_sf(data = boroughs, color = "white") +
  geom_sf(data = landings, color = "black", size = 2, alpha = 0.5) +
  geom_sf(aes(fill = mean_speed), color = NA) +
  coord_sf(
    xlim = c(bbox_hull["xmin"] - 100, bbox_hull["xmax"] + 100),
    ylim = c(bbox_hull["ymin"] - 200, bbox_hull["ymax"] + 200),
    expand = FALSE
  ) +
  scale_fill_viridis_c(option = "D", name = "Knots") +
  labs(title = "Mean Vessel Speed by Cell") +
  theme_solid() +
  theme(
    plot.title = element_text(size = 16),
    plot.title.position = "plot"
  )
```

### Kernel Density Plots — Hourly

To analyze how OTP changes spatially throughout the PM peak, kernel densities of excessively late data points were mapped for each hour between 1pm and 10pm.

```{r swiftly-kds}
#| dev: "ragg_png"
#| dpi: 300
#| warning: false
#| message: false

make_hourly_kd <- function(hour_val) {
  net <- st_join(fishnet_swiftly, jun2025_reference_week_excessively_lates_sf, join = st_intersects) %>%
    group_by(uniqueID) %>%
    filter(hour == hour_val) %>%
    summarise(mean_speed = mean(knots, na.rm = TRUE), n = n(), .groups = "drop")
  kd <- density.ppp(as.ppp(st_coordinates(net), W = st_bbox(jun2025_net)), sigma = 1000)
  kd
}

kd_list <- list(
  '1PM'  = make_hourly_kd("13"),
  '2PM'  = make_hourly_kd("14"),
  '3PM'  = make_hourly_kd("15"),
  '4PM'  = make_hourly_kd("16"),
  '5PM'  = make_hourly_kd("17"),
  '6PM'  = make_hourly_kd("18"),
  '7PM'  = make_hourly_kd("19"),
  '8PM'  = make_hourly_kd("20"),
  '9PM'  = make_hourly_kd("21"),
  '10PM' = make_hourly_kd("22")
)

all_values <- unlist(lapply(kd_list, \(kd) as.data.frame(kd)$value))
global_limits <- range(all_values, na.rm = TRUE)

plots <- imap(kd_list, ~{
  sf_pts <- st_as_sf(as.data.frame(.x), coords = c("x", "y"), crs = st_crs(jun2025_net))
  agg <- aggregate(sf_pts["value"], jun2025_net, mean, na.rm = TRUE)

  ggplot() +
    geom_sf(data = boroughs, color = "white") +
    geom_sf(data = agg, aes(fill = value), color = NA, linewidth = 0) +
    coord_sf(
      xlim = c(bbox_hull["xmin"] - 100, bbox_hull["xmax"] + 100),
      ylim = c(bbox_hull["ymin"] - 100, bbox_hull["ymax"] + 100),
      expand = FALSE
    ) +
    scale_fill_viridis_c(name = "Kernel Density", limits = global_limits, oob = scales::squish) +
    labs(title = paste("KD -", .y)) +
    theme_solid() +
    theme(
      plot.title = element_text(size = 12),
      plot.title.position = "plot"
    )
})

wrap_plots(plots, ncol = 5) +
  plot_layout(guides = "collect") +
  plot_annotation(
    title = "Excessively Late Across TOD",
    caption = "Unified KD Scale, 250m Grid Cell Size"
  ) &
  theme(
    legend.position = "right",
    legend.box = "horizontal",
    plot.margin = margin(1, 1, 1, 1)
  )
```

One potentially useful consideration is to observe locations where trips change OTP status. The following figure includes only instances where OTP status changes to either late or excessively late.

```{r swiftly-otp-transitions}
#| warning: false
#| message: false

jun2025_reference_week_late_sf <- jun2025_reference_week_sf %>%
  filter(OTP_transition %in% c("Early -> Late", "Late -> Excessively Late",
                                "Ontime -> Late", "Ontime -> Excessively Late"))

ggplot() +
  geom_sf(data = boroughs, color = "white") +
  geom_sf(data = landings, color = "black", size = 2, alpha = 0.2) +
  geom_sf(data = jun2025_reference_week_late_sf,
          aes(color = OTP_transition), size = 2, alpha = 0.2) +
  coord_sf(
    xlim = c(bbox_east_river["xmin"] - 10000, bbox_east_river["xmax"] + 10000),
    ylim = c(bbox_east_river["ymin"] - 10000, bbox_east_river["ymax"] + 10000),
    expand = FALSE
  ) +
  labs(title = "OTP Transition Locations") +
  scale_color_manual(
    name = "Transition Type",
    values = c(
      "Early -> Late"               = "#1e3888",
      "Ontime -> Late"              = "#47a8bd",
      "Ontime -> Excessively Late"  = "#ffad69",
      "Late -> Excessively Late"    = "#9c3848"
    ),
    labels = c(
      "Early -> Late"               = "Early to Late",
      "Ontime -> Late"              = "Ontime to Late",
      "Ontime -> Excessively Late"  = "Ontime to Excessively Late",
      "Late -> Excessively Late"    = "Late to Excessively Late"
    )
  ) +
  facet_wrap(~OTP_transition) +
  theme_solid() +
  theme(
    plot.title = element_text(size = 20),
    plot.title.position = "plot"
  )
```

### Leaflet Distribution

```{r swiftly-leaflet}
#| warning: false
#| message: false
#| eval: false

landings_buffers_100ft <- st_buffer(landings, dist = 30.48)
counts <- lengths(st_intersects(landings_buffers_100ft, jun2025_reference_week_excessively_lates_sf))
landings_buffers_100ft$record_count <- counts

points_plot  <- st_transform(landings, 4326)
buffers_plot <- st_transform(landings_buffers_100ft, 4326)
routes_plot  <- st_transform(routes, 4326)
boroughs_plot <- st_transform(boroughs, 4326)

jun2025_reference_week_elate_sf <- jun2025_reference_week_sf %>%
  filter(OTP == "Excessively Late") %>%
  st_transform(4326)

jun2025_reference_week_not_elate_sf <- jun2025_reference_week_sf %>%
  filter(OTP != "Excessively Late") %>%
  st_transform(4326)

leaflet() %>%
  addProviderTiles("Esri.WorldImagery", group = "Imagery",
                   options = tileOptions(opacity = 0.6)) %>%
  addPolygons(data = boroughs_plot, color = "#828282", weight = 1, fillOpacity = 0.75, group = "Boroughs") %>%
  addPolylines(data = routes_plot, color = "#006bb6", weight = 2, opacity = 0.2, group = "Routes") %>%
  addPolygons(data = buffers_plot, color = "#f58426", weight = 10, fillOpacity = 0.75, group = "Landings",
              label = ~paste0("Records: ", record_count)) %>%
  addCircleMarkers(data = points_plot, radius = 3, color = "#f58426", opacity = 0.6, group = "Landings") %>%
  addCircleMarkers(data = jun2025_reference_week_elate_sf, radius = 0.5, color = "#47a8bd",
    opacity = 0.6, fillOpacity = 0.4, group = "Excessively Late",
    label = ~paste0("Speed: ", round(knots, 2)),
    popup = ~paste0("<b>Speed:</b> ", round(knots, 2), " knots",
                    "<br><b>Route:</b> ", routeId,
                    "<br><b>Trip:</b> ", tripId,
                    "<br><b>Vessel:</b> ", vehicleId,
                    "<br><b>Time:</b> ", time)) %>%
  addCircleMarkers(data = jun2025_reference_week_not_elate_sf, radius = 0.5, color = "#9c3848",
    opacity = 0.6, fillOpacity = 0.4, group = "Not Excessively Late",
    label = ~paste0("Speed: ", round(knots, 2)),
    popup = ~paste0("<b>Speed:</b> ", round(knots, 2), " knots",
                    "<br><b>Route:</b> ", routeId,
                    "<br><b>Trip:</b> ", tripId,
                    "<br><b>Vessel:</b> ", vehicleId,
                    "<br><b>Time:</b> ", time)) %>%
  addLayersControl(
    overlayGroups = c("Boroughs", "Routes", "Landings", "Excessively Late", "Not Excessively Late"),
    baseGroups = c("Imagery"),
    options = layersControlOptions(collapsed = FALSE)
  )
```

This Leaflet map shows the data with an imagery layer for context. For the sake of processing power, the results are not shown by default — remove the `eval: false` chunk option to render it.



# Modeling

## Associative Model
**By: Yiming 'Ming' Cao**

### Objective and Design

This section presents an associative model for explaining stop-level arrival lateness in the NYC Ferry system. The goal is interpretation rather than advance prediction. Instead of asking only whether a future trip will be delayed, this model asks which historical factors are associated with observed delay patterns.

The unit of analysis is a stop-level observation. The dependent variable is arrival lateness in minutes relative to the original schedule. Predictors are organized into five explanatory domains: structural context, environmental conditions, temporal patterns, operational propagation, and vessel characteristics.

This grouped design allows us to compare how different types of factors contribute to delay variation while keeping the final model interpretable for operational use.

```{r ming-settings}
#| message: false
#| warning: false

set.seed(7900)

input_parquet <- "../exploratory_code/modeling/associative/baseline_modeling_table_final_direct.parquet"
outcome_cont  <- "lateness_vs_original_schedule_min"

week_window_days <- 3

candidate_predictors <- c(
  "route_id", "direction", "stop_sequence", "prev_stop_sched_hour", "prev_stop_sched_hour_group",
  "schedule_season", "schedule_daytype", "vessel_trip_seq_in_chain", "vessel_capacity",
  "is_terminal_stop", "prev_stop_shared_stop_flag", "curr_stop_shared_stop_flag",
  "planned_turnaround_gap_min", "scheduled_headway_min", "expected_water_min",
  "prev_stop_sequence", "prev_stop_boardings", "prev_stop_alightings", "prev_stop_onboard",
  "prev_stop_load_factor", "env_temperature_c", "env_wind_speed_mps", "env_visibility_m",
  "env_precipitation_mm", "env_battery_water_level_anomaly", "env_narrows_current_residual",
  "env_battery_water_level_obs", "env_narrows_current_obs"
)
prior_delay_predictors <- c("inherited_prev_trip_delay", "carried_within_trip_delay")
```

### Data Scope

The model uses the stop-level baseline modeling table as its input dataset. After loading the data, we confirmed the number of records and the availability of candidate predictors before fitting the associative models.

```{r ming-load}
#| message: false
#| warning: false

if (!file.exists(input_parquet)) stop("Missing input parquet: ", input_parquet)
df_raw <- read_parquet(input_parquet) %>% as_tibble()

df <- df_raw %>%
  mutate(
    trip_date = as.Date(substr(as.character(trip_key), 1, 8), format = "%Y%m%d")
  )

predictors_no_prior   <- intersect(candidate_predictors, names(df))
predictors_with_prior <- unique(c(predictors_no_prior, intersect(prior_delay_predictors, names(df))))

tibble(
  input_rows              = nrow(df_raw),
  rows_after_load         = nrow(df),
  predictors_no_prior_n   = length(predictors_no_prior),
  predictors_with_prior_n = length(predictors_with_prior)
) %>% kable(caption = "Data Scope and Predictor Availability")
```

### Predictor Domains

To support interpretation, predictors in the associative model were organized into five explanatory domains. This grouped structure separates baseline route and stop context from environmental conditions, temporal patterns, operational propagation, and vessel-related characteristics. The goal is to interpret delay association at the domain level rather than focusing only on isolated individual coefficients.

#### Domain A: Structural Context

The structural domain captures baseline differences in how delay varies across the ferry network. These predictors describe route, stop position, and stop-context features that reflect how service is organized spatially and operationally. Variables in this domain include route and direction identifiers, stop sequence, terminal status, and shared-stop context. We interpret this domain as representing the structural baseline of delay rather than short-term changing conditions.

#### Domain B: Environmental Conditions

The environmental domain captures external natural conditions that may affect vessel movement and docking performance. These predictors include weather and water-related variables such as temperature, wind speed, visibility, precipitation, water-level, and current-related measures. This domain is especially important for the client because it helps distinguish external operating conditions from delay that is simply inherited from earlier parts of the trip.

#### Domain C: Temporal Patterns

The temporal domain captures recurring calendar and time-of-day structure in delay. These predictors describe when service occurs rather than where it occurs or what the vessel has already experienced. Variables in this domain include scheduled hour, scheduled hour group, day type, and season. We use this domain to represent regular service rhythms such as weekday versus weekend patterns, seasonal demand differences, and time-of-day effects.

#### Domain D: Operational Propagation

The operational propagation domain captures delay carried forward through the operating chain. These predictors reflect whether lateness has already accumulated earlier in the trip or in a prior linked trip on the same vessel. Variables in this domain include inherited previous-trip delay and carried within-trip delay. This domain is operationally meaningful, but it is interpreted as one component of the associative model rather than the sole basis for explaining delay.

#### Domain E: Vessel Characteristics

The vessel domain captures systematic differences associated with vessel capacity and operating role. These predictors help assess whether some delay patterns are associated with the characteristics of the vessel being used rather than only with route structure or environmental conditions. Variables in this domain include vessel capacity and related vessel-level operational identifiers retained in the final specification.

### Data Preparation

Before model estimation, the stop-level modeling table was prepared to support a consistent and interpretable associative model. This section summarizes the key preparation steps: variable construction, missing-value treatment for environmental fields, and helper functions used for model fitting and summary extraction.

#### Data Cleaning and Variable Construction

The input table was converted into a stop-level analytic dataset. We created a trip date field from the trip identifier, checked which candidate predictors were available, and retained variables needed for the outcome and domain-based model structure.

#### Environmental Variable Imputation

Missing environmental values were imputed using a local time-slot rolling rule. The primary fill used observations from the same scheduled hour within a centered seven-day window around the trip date. If no value was available, the method fell back to the same scheduled hour group within the same centered window.

This approach keeps environmental predictors in the model without relying only on complete-case filtering. It also keeps the imputed values tied to nearby dates and comparable operating periods.

```{r ming-impute}
#| message: false
#| warning: false

build_week_fill_map <- function(day_slot_tbl, group_col, date_col = "trip_date", value_col = "slot_day_median", window_days = 3) {
  if (nrow(day_slot_tbl) == 0) return(tibble())
  split_tbl <- split(day_slot_tbl, day_slot_tbl[[group_col]], drop = TRUE)
  out <- lapply(split_tbl, function(gdf) {
    gdf <- gdf[order(gdf[[date_col]]), , drop = FALSE]
    d <- gdf[[date_col]]
    v <- gdf[[value_col]]
    week_fill <- vapply(seq_along(v), function(i) {
      idx <- abs(as.integer(d - d[i])) <= window_days
      val <- suppressWarnings(median(v[idx], na.rm = TRUE))
      ifelse(is.finite(val), val, NA_real_)
    }, numeric(1))
    gdf$week_fill <- week_fill
    gdf[, c(group_col, date_col, "week_fill"), drop = FALSE]
  })
  bind_rows(out)
}

impute_by_local_slot_week <- function(data, var_name, window_days = 3) {
  x <- data
  miss_before <- sum(is.na(x[[var_name]]))
  if (miss_before == 0) {
    return(list(data = x, summary = tibble(variable = var_name, n_missing_before = 0L, n_filled_primary = 0L, n_filled_fallback = 0L, n_missing_after = 0L)))
  }

  primary_daily <- x %>%
    filter(!is.na(.data[[var_name]]), !is.na(trip_date), !is.na(prev_stop_sched_hour)) %>%
    group_by(prev_stop_sched_hour, trip_date) %>%
    summarise(slot_day_median = median(.data[[var_name]], na.rm = TRUE), .groups = "drop")
  primary_fill <- build_week_fill_map(primary_daily, group_col = "prev_stop_sched_hour", value_col = "slot_day_median", window_days = window_days) %>%
    rename(fill_primary = week_fill)

  fallback_daily <- x %>%
    filter(!is.na(.data[[var_name]]), !is.na(trip_date), !is.na(prev_stop_sched_hour_group)) %>%
    group_by(prev_stop_sched_hour_group, trip_date) %>%
    summarise(slot_day_median = median(.data[[var_name]], na.rm = TRUE), .groups = "drop")
  fallback_fill <- build_week_fill_map(fallback_daily, group_col = "prev_stop_sched_hour_group", value_col = "slot_day_median", window_days = window_days) %>%
    rename(fill_fallback = week_fill)

  x <- x %>%
    left_join(primary_fill, by = c("prev_stop_sched_hour", "trip_date")) %>%
    left_join(fallback_fill, by = c("prev_stop_sched_hour_group", "trip_date"))

  was_missing      <- is.na(x[[var_name]])
  fill_primary_idx  <- was_missing & !is.na(x$fill_primary)
  fill_fallback_idx <- was_missing & is.na(x$fill_primary) & !is.na(x$fill_fallback)

  x[[var_name]][fill_primary_idx]  <- x$fill_primary[fill_primary_idx]
  x[[var_name]][fill_fallback_idx] <- x$fill_fallback[fill_fallback_idx]
  x <- x %>% select(-fill_primary, -fill_fallback)

  list(
    data = x,
    summary = tibble(
      variable          = var_name,
      n_missing_before  = miss_before,
      n_filled_primary  = sum(fill_primary_idx, na.rm = TRUE),
      n_filled_fallback = sum(fill_fallback_idx, na.rm = TRUE),
      n_missing_after   = sum(is.na(x[[var_name]]))
    )
  )
}

impute_vars <- intersect(
  c("env_temperature_c", "env_wind_speed_mps", "env_visibility_m",
    "env_precipitation_mm", "env_narrows_current_residual",
    "env_battery_water_level_obs", "env_narrows_current_obs"),
  names(df)
)

impute_summary_list <- list()
df_imp <- df
for (v in impute_vars) {
  res <- impute_by_local_slot_week(df_imp, v, window_days = week_window_days)
  df_imp <- res$data
  impute_summary_list[[length(impute_summary_list) + 1]] <- res$summary
}
imputation_summary <- bind_rows(impute_summary_list)
```

```{r ming-imputation-summary}
imputation_summary %>%
  mutate(
    pct_missing_before = round(n_missing_before / nrow(df) * 100, 2),
    pct_missing_after  = round(n_missing_after  / nrow(df_imp) * 100, 2)
  ) %>%
  kable(caption = "Summary of Environmental Variable Imputation", digits = 2)
```

#### Modeling Utilities

A small set of reusable helper functions was created to prepare model inputs, fit linear models, extract model metrics, and generate standardized coefficient summaries.

```{r ming-helpers}
#| message: false
#| warning: false

prepare_model_df <- function(data, outcome, predictors) {
  data %>% select(all_of(c(outcome, predictors))) %>% drop_na()
}

fit_linear <- function(data, model_name, outcome, predictors) {
  d   <- prepare_model_df(data, outcome, predictors)
  fml <- as.formula(paste(outcome, "~", paste(predictors, collapse = " + ")))
  fit <- lm(fml, data = d)
  pred <- predict(fit, newdata = d)
  rmse <- sqrt(mean((d[[outcome]] - pred)^2))
  g    <- glance(fit)

  list(
    metrics = tibble(
      model_name  = model_name, model_type = "linear",
      n = nrow(d), aic = g$AIC, bic = g$BIC,
      rsq = g$r.squared, adj_rsq = g$adj.r.squared, rmse = rmse,
      pseudo_rsq = NA_real_, accuracy = NA_real_,
      sensitivity = NA_real_, specificity = NA_real_, auc = NA_real_
    ),
    coefs        = tidy(fit) %>% mutate(model_name = model_name, model_type = "linear"),
    formula_text = paste(deparse(fml), collapse = "")
  )
}

standardize_for_lm <- function(data, outcome, predictors) {
  d       <- data %>% select(all_of(c(outcome, predictors))) %>% drop_na()
  num_vars <- names(d)[sapply(d, is.numeric)]
  d %>% mutate(across(all_of(num_vars), ~ as.numeric(scale(.x))))
}

fit_standardized_linear <- function(data, model_name, outcome, predictors) {
  d_std <- standardize_for_lm(data, outcome, predictors)
  fml   <- as.formula(paste(outcome, "~", paste(predictors, collapse = " + ")))
  fit   <- lm(fml, data = d_std)
  broom::tidy(fit) %>%
    filter(term != "(Intercept)") %>%
    mutate(model_name = model_name, abs_estimate = abs(estimate)) %>%
    arrange(desc(abs_estimate))
}
```

### Variable Screening and Final Specification

This section explains how the final associative-model specification was selected from the broader predictor pool. The goal was not to include every available variable, but to keep interpretable coverage across the five explanatory domains while reducing redundancy among closely related predictors.

#### Initial Candidate Variables

The initial predictor pool was intentionally broad so that each explanatory domain was represented. Candidate variables included route and stop structure, environmental conditions, temporal indicators, operational propagation terms, and vessel/load-related characteristics. This broader pool provided the starting point for a smaller, more interpretable final specification.

```{r ming-domain-predictors}
#| message: false
#| warning: false

domain_a_predictors <- intersect(c(
  "route_id", "direction", "stop_sequence", "is_terminal_stop",
  "prev_stop_shared_stop_flag", "curr_stop_shared_stop_flag"
), names(df_imp))

domain_b_predictors <- intersect(c(
  "env_temperature_c", "env_wind_speed_mps", "env_visibility_m",
  "env_precipitation_mm", "env_narrows_current_residual",
  "env_battery_water_level_obs", "env_narrows_current_obs"
), names(df_imp))

domain_c_predictors <- intersect(c(
  "prev_stop_sched_hour", "schedule_season", "schedule_daytype"
), names(df_imp))

domain_d_predictors <- intersect(c(
  "inherited_prev_trip_delay", "carried_within_trip_delay"
), names(df_imp))

domain_e_predictors <- intersect(c(
  "vessel_capacity", "vessel_trip_seq_in_chain",
  "prev_stop_boardings", "prev_stop_alightings",
  "prev_stop_onboard", "prev_stop_load_factor"
), names(df_imp))

final_predictors <- unique(c(
  domain_a_predictors, domain_b_predictors,
  domain_c_predictors, domain_d_predictors, domain_e_predictors
))
```

#### Redundancy and Multicollinearity Checks

Because several candidate predictors measured closely related aspects of stop position, ridership, and operating context, we reviewed pairwise correlations among numeric predictors before fitting the final associative model. The goal was not to eliminate all correlation, but to identify clusters of variables that conveyed largely overlapping information and would make coefficient interpretation less clear.

```{r ming-corr-setup}
#| message: false
#| warning: false

numeric_candidate_predictors <- intersect(
  candidate_predictors,
  names(df_imp)[sapply(df_imp, is.numeric)]
)

corr_vars   <- df_imp %>%
  select(all_of(numeric_candidate_predictors)) %>%
  select(where(~ !all(is.na(.))))

corr_matrix <- cor(corr_vars, use = "pairwise.complete.obs")
```

The figure below highlights selected variables reviewed during specification design. These variables are related to stop position and previous-stop load conditions, so they were checked carefully before finalizing the model.

```{r ming-corr-plot}
#| fig-width: 6
#| fig-height: 5

focus_vars <- c("stop_sequence", "prev_stop_sequence", "prev_stop_onboard", "prev_stop_load_factor")
focus_corr <- corr_matrix[focus_vars, focus_vars]

as.data.frame(as.table(focus_corr)) %>%
  rename(var1 = Var1, var2 = Var2, corr = Freq) %>%
  mutate(label = ifelse(var1 == var2, "", sprintf("%.2f", corr))) %>%
  ggplot(aes(x = var1, y = var2, fill = corr)) +
  geom_tile(color = "white", linewidth = 0.5) +
  geom_text(aes(label = label), size = 4) +
  scale_fill_gradient2(low = "#1A9988", mid = "white", high = "#EB5600", midpoint = 0, limits = c(-1, 1)) +
  coord_fixed() +
  theme_minimal() +
  theme(
    panel.grid = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(size = 13, face = "bold", hjust = 0.5)
  ) +
  labs(title = "Correlation Among Selected Candidate Predictors", x = NULL, y = NULL, fill = "Correlation")
```

To support specification decisions more systematically, we also identified highly correlated numeric predictor pairs using an absolute correlation threshold of 0.7.

```{r ming-corr-pairs}
as.data.frame(as.table(corr_matrix)) %>%
  rename(var1 = Var1, var2 = Var2, corr = Freq) %>%
  filter(var1 != var2) %>%
  mutate(abs_corr = abs(corr)) %>%
  filter(abs_corr >= 0.7) %>%
  rowwise() %>%
  mutate(pair_id = paste(sort(c(var1, var2)), collapse = " || ")) %>%
  ungroup() %>%
  distinct(pair_id, .keep_all = TRUE) %>%
  arrange(desc(abs_corr)) %>%
  select(var1, var2, corr, abs_corr) %>%
  head(10) %>%
  kable(caption = "Highly correlated candidate predictor pairs (|r| >= 0.7)")
```

#### Final Variable Set

The final variable set was selected to preserve coverage across the five explanatory domains while avoiding unnecessary duplication. When multiple variables represented similar information, we prioritized predictors that were easier to interpret operationally and more directly connected to the domain being represented.

After reviewing correlation patterns, we removed redundant candidate variables and retained a reduced final specification for the combined associative model.

```{r ming-vif-setup}
#| message: false
#| warning: false

final_predictors_trimmed <- setdiff(
  final_predictors,
  c("prev_stop_sequence", "prev_stop_sched_hour_group")
)

tibble(
  model        = "Final combined associative model",
  n_predictors = length(final_predictors_trimmed)
) %>% kable(caption = "Predictor count in the final combined associative specification")
```

```{r ming-vif-run}
#| message: false
#| warning: false

fit_vif_model <- function(data, outcome, predictors) {
  model_df <- data %>% select(all_of(unique(c(outcome, predictors)))) %>% drop_na()
  lm(as.formula(paste(outcome, "~", paste(predictors, collapse = " + "))), data = model_df)
}

extract_vif_table <- function(model, model_name) {
  vif_vals <- car::vif(model)
  if (is.matrix(vif_vals)) {
    out <- as.data.frame(vif_vals)
    out$term <- rownames(out)
    rownames(out) <- NULL
    out$model <- model_name
    out %>% rename(gvif = GVIF) %>%
      mutate(adjusted_vif = gvif^(1 / (2 * Df))) %>%
      select(model, term, Df, gvif, adjusted_vif)
  } else {
    tibble(model = model_name, term = names(vif_vals), vif = as.numeric(vif_vals)) %>%
      arrange(desc(vif))
  }
}

vif_model_final <- fit_vif_model(df_imp, outcome_cont, final_predictors_trimmed)
vif_table_final <- extract_vif_table(vif_model_final, "Final combined associative model")
```

We then checked variance inflation factors for the final combined specification. For multi-level categorical predictors, we report the adjusted GVIF, which is more comparable across terms with different degrees of freedom.

```{r ming-vif-table}
if ("adjusted_vif" %in% names(vif_table_final)) {
  vif_table_final %>%
    arrange(desc(adjusted_vif)) %>%
    head(10) %>%
    mutate(gvif = round(gvif, 2), adjusted_vif = round(adjusted_vif, 2)) %>%
    kable(caption = "Top VIF results for the final combined associative specification")
} else {
  vif_table_final %>%
    arrange(desc(vif)) %>%
    head(10) %>%
    mutate(vif = round(vif, 2)) %>%
    kable(caption = "Top VIF results for the final combined associative specification")
}
```

### Estimation Strategy

The associative model was estimated in two steps. First, we fit separate domain-specific linear models to summarize how much stop-level lateness was associated with each explanatory domain on its own. Second, we fit a combined associative model that included predictors from all five domains.

To make model comparisons more consistent, all models were estimated on the same analytic sample used for the final combined specification.

#### Domain-Specific Models

We estimated separate linear models for the five explanatory domains: structural context, environmental conditions, temporal patterns, operational propagation, and vessel/load characteristics. Each model used the same dependent variable: stop-level arrival lateness in minutes.

These models were not treated as competing final models. Instead, they were used to compare the relative explanatory signal of each domain when considered separately.

```{r ming-fit}
#| message: false
#| warning: false

df_model <- df_imp %>%
  drop_na(all_of(c(outcome_cont, final_predictors_trimmed)))

model_results <- list(
  domain_a_structural   = fit_linear(df_model, "domain_a_structural",   outcome_cont, domain_a_predictors),
  domain_b_environment  = fit_linear(df_model, "domain_b_environment",  outcome_cont, domain_b_predictors),
  domain_c_temporal     = fit_linear(df_model, "domain_c_temporal",     outcome_cont, domain_c_predictors),
  domain_d_operational  = fit_linear(df_model, "domain_d_operational",  outcome_cont, domain_d_predictors),
  domain_e_vessel       = fit_linear(df_model, "domain_e_vessel",       outcome_cont, domain_e_predictors),
  combined_associative  = fit_linear(df_model, "combined_associative",  outcome_cont, final_predictors_trimmed)
)
```

#### Combined Associative Model

The combined associative model uses the final predictor set across all five domains. This model serves as the main explanatory specification. It allows us to interpret individual coefficients while also discussing how different types of factors contribute to observed delay patterns within one framework.

#### Comparison Logic

Model comparison focused on interpretation rather than predictive model selection. We compared adjusted R-squared and RMSE across the domain-specific models and the combined model. The domain-specific models show how much delay variation is associated with each factor group on its own, while the combined model shows the explanatory power of the full specification.

### Results: Explanatory Contribution by Domain

This section compares how much stop-level delay variation is explained by each explanatory domain. The goal is not to choose a predictive model, but to understand which types of factors are most strongly associated with observed lateness.

The domain-specific models show a clear hierarchy. Operational propagation explains the largest share of delay variation, followed by structural context. Environmental conditions, temporal patterns, and vessel/load characteristics explain smaller shares on their own, but they remain useful for interpretation because they describe monitorable operating conditions.

```{r ming-model-comparison}
model_comparison_summary <- bind_rows(lapply(model_results, `[[`, "metrics")) %>%
  mutate(
    domain = dplyr::recode(
      model_name,
      combined_associative  = "Combined Associative Model",
      domain_a_structural   = "Domain A: Structural Context",
      domain_b_environment  = "Domain B: Environmental Conditions",
      domain_c_temporal     = "Domain C: Temporal Patterns",
      domain_d_operational  = "Domain D: Operational Propagation",
      domain_e_vessel       = "Domain E: Vessel / Load Characteristics"
    )
  ) %>%
  arrange(match(domain, c(
    "Domain A: Structural Context", "Domain B: Environmental Conditions",
    "Domain C: Temporal Patterns",  "Domain D: Operational Propagation",
    "Domain E: Vessel / Load Characteristics", "Combined Associative Model"
  )))

model_comparison_summary %>%
  select(domain, n, rsq, adj_rsq, rmse, aic, bic) %>%
  mutate(across(c(rsq, adj_rsq, rmse), ~ round(.x, 4)),
         across(c(aic, bic), ~ round(.x, 0))) %>%
  kable(caption = "Explained Variation by Domain and Final Combined Associative Model")
```

```{r ming-domain-rsq-plot}
#| fig-width: 8
#| fig-height: 4.5

model_comparison_summary %>%
  filter(domain != "Combined Associative Model") %>%
  mutate(domain = factor(domain, levels = domain), adj_rsq = round(adj_rsq, 4)) %>%
  ggplot(aes(x = domain, y = adj_rsq)) +
  geom_col(width = 0.65) +
  geom_text(aes(label = round(adj_rsq, 3)), vjust = -0.3, size = 3.5) +
  labs(title = "Explanatory Contribution by Domain", x = NULL, y = "Adjusted R-squared") +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 25, hjust = 1), plot.title = element_text(face = "bold"))
```

#### Structural Context

The structural-context model explains a meaningful share of stop-level delay variation on its own. This indicates that route and stop configuration establish an important baseline pattern of delay across the system. In substantive terms, some portions of the ferry network appear structurally more delay-prone than others even before environmental or operational carry-forward effects are considered.

#### Environmental Conditions

The environmental model explains a smaller share of total variation on its own. This does not mean environmental signals are unimportant. Instead, it suggests that weather and water conditions are secondary to propagated delay and structural context, while still offering useful monitorable signals for operational awareness.

#### Temporal Patterns

The temporal model explains a modest share of variation, indicating that day type, scheduled hour, and season capture recurring service rhythms but do not dominate the delay process on their own. These results suggest that delay varies systematically across time, but those temporal patterns are weaker than operational propagation and structural context.

#### Operational Propagation

The operational-propagation model explains the largest share of variation among the individual domains. This confirms that delay accumulation within and across trips is the strongest single explanatory component in the associative model. In other words, stop-level delay is highly path-dependent: once lateness has entered the operating chain, a substantial portion of subsequent delay can be understood as propagated rather than newly generated at each stop.

#### Vessel / Load Characteristics

The vessel/load model explains the smallest share of variation among the five domains. This suggests that vessel capacity and load-related differences are present but relatively limited when considered in isolation. These factors are still useful as operational context, but they are secondary to route structure, external conditions, and propagated delay in explaining stop-level lateness.

### Focused Interpretation of Key Predictors

The previous section compared explanatory contribution across domains. This section focuses on selected predictors used for coefficient interpretation and final presentation findings. These results are intended to make the model findings easier to interpret operationally, not to replace the full combined associative model.

#### Focused Interpretive Specification

In addition to the full combined associative model, we estimated a smaller interpretive specification using predictors selected for coefficient interpretation. This focused model includes prior delay, route context, seasonal and day-type patterns, time bins, selected environmental conditions, and stop sequence.

This model is used to support coefficient interpretation in the final presentation. It should be read as a simplified interpretive view of the larger associative framework.

```{r ming-focused-model}
selected_predictors <- intersect(c(
  "inherited_prev_trip_delay", "route_id", "schedule_season",
  "prev_stop_sched_hour_group", "schedule_daytype",
  "env_temperature_c", "env_wind_speed_mps", "env_precipitation_mm",
  "env_battery_water_level_obs", "stop_sequence"
), names(df_model))

selected_model_df <- prepare_model_df(df_model, outcome_cont, selected_predictors)

selected_formula <- as.formula(paste(
  outcome_cont, "~",
  paste(selected_predictors, collapse = " + "),
  "+ prev_stop_sched_hour_group:schedule_daytype"
))

selected_fit <- lm(selected_formula, data = selected_model_df)

broom::glance(selected_fit) %>%
  transmute(
    model   = "Focused interpretive specification",
    n       = nobs(selected_fit),
    rsq     = round(r.squared, 4),
    adj_rsq = round(adj.r.squared, 4),
    rmse    = round(sqrt(mean(residuals(selected_fit)^2)), 4),
    aic     = round(AIC(selected_fit), 0),
    bic     = round(BIC(selected_fit), 0)
  ) %>%
  kable(caption = "Focused Interpretive Specification Fit")
```

#### Raw Coefficient Interpretation

To support operational interpretation, we report raw coefficients from the focused interpretive specification. These coefficients show the estimated change in arrival lateness, in minutes, associated with each predictor while holding other variables in the model constant.

For numeric predictors, the coefficient is interpreted per one-unit increase. For categorical predictors, coefficients are interpreted relative to the reference category.

```{r ming-raw-coefficients}
broom::tidy(selected_fit) %>%
  filter(term != "(Intercept)") %>%
  mutate(abs_estimate = abs(estimate)) %>%
  arrange(desc(abs_estimate)) %>%
  select(term, estimate, std.error, statistic, p.value) %>%
  mutate(across(c(estimate, std.error), ~ round(.x, 4)),
         statistic = round(statistic, 2),
         p.value   = round(p.value, 4)) %>%
  kable(caption = "Raw Coefficients for Focused Interpretive Specification")
```

#### Supporting Exploratory Patterns

The following exploratory plots provide visual context for the coefficient interpretation. They show how delay accumulates, how lateness varies across time bins and day type, and how selected environmental and route-level patterns relate to average lateness.

```{r ming-delay-components}
#| fig-width: 8
#| fig-height: 4

tibble::tibble(
  component = factor(
    c("Inherited previous delay", "Carried within-trip delay", "New delay at current stop"),
    levels = c("Inherited previous delay", "Carried within-trip delay", "New delay at current stop")
  ),
  minutes = c(6.24, 1.85, 1.68)
) %>%
  ggplot(aes(x = "Excessive delay example", y = minutes, fill = component)) +
  geom_col(width = 0.45) +
  geom_text(aes(label = paste0(minutes, " min")),
            position = position_stack(vjust = 0.5),
            color = "white", fontface = "bold", size = 4) +
  scale_fill_manual(values = c(
    "Inherited previous delay"   = "#EB5600",
    "Carried within-trip delay"  = "#1A9988",
    "New delay at current stop"  = "#C7C7C7"
  )) +
  labs(title    = "Where Does Excessive Delay Come From?",
       subtitle = "Most excessive delay is inherited or carried forward, not newly created at the current stop",
       x = NULL, y = "Minutes of delay", fill = NULL) +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        panel.grid.major.x = element_blank(), panel.grid.minor = element_blank(),
        legend.position = "bottom", plot.title = element_text(face = "bold"))
```

```{r ming-delay-accumulation}
#| fig-width: 8
#| fig-height: 5

df_model %>%
  group_by(route_id, stop_sequence) %>%
  summarise(avg_lateness = mean(.data[[outcome_cont]], na.rm = TRUE), n = n(), .groups = "drop") %>%
  filter(n >= 100) %>%
  ggplot(aes(x = stop_sequence, y = avg_lateness, color = route_id)) +
  geom_line(linewidth = 0.9) +
  geom_point(size = 1.8) +
  labs(title    = "Delay Accumulation Along Route",
       subtitle = "Average lateness tends to change as trips move through stop sequence",
       x = "Stop sequence", y = "Average lateness (minutes)", color = "Route") +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold"), legend.position = "bottom")
```

```{r ming-time-of-day-setup}
#| message: false
#| warning: false

lateness_col <- c(outcome_cont, "arr_lateness", "arr_late_min", "arrival_lateness",
                  "delay_min", "lateness")[
  c(outcome_cont, "arr_lateness", "arr_late_min", "arrival_lateness",
    "delay_min", "lateness") %in% names(df_model)][1]

if (is.na(lateness_col)) stop("Could not find the lateness column.")

time_delay_summary <- df_model %>%
  mutate(lateness_for_plot = .data[[lateness_col]]) %>%
  filter(!is.na(prev_stop_sched_hour_group), !is.na(schedule_daytype), !is.na(lateness_for_plot)) %>%
  group_by(prev_stop_sched_hour_group, schedule_daytype) %>%
  summarise(avg_lateness = mean(lateness_for_plot, na.rm = TRUE), n = n(), .groups = "drop") %>%
  mutate(prev_stop_sched_hour_group = factor(
    prev_stop_sched_hour_group, levels = c("5_8", "9_11", "12_14", "15_17", "18_22")
  ))
```

```{r ming-time-of-day-plot}
#| fig-width: 8
#| fig-height: 5

ggplot(time_delay_summary, aes(x = prev_stop_sched_hour_group, y = avg_lateness, fill = schedule_daytype)) +
  geom_col(position = position_dodge(width = 0.75), width = 0.6) +
  scale_fill_manual(values = c("WEEKDAY" = "#EB5600", "WEEKEND" = "#1A9988")) +
  labs(title = "Average Lateness by Time Bin and Day Type",
       x = "Scheduled time bin", y = "Average lateness (minutes)", fill = "Day type") +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold"), legend.position = "bottom")
```

```{r ming-temperature-setup}
#| message: false
#| warning: false

temp_delay_summary <- df_model %>%
  filter(!is.na(env_temperature_c), !is.na(.data[[outcome_cont]])) %>%
  mutate(temp_bin = cut(env_temperature_c,
                        breaks = quantile(env_temperature_c, probs = seq(0, 1, 0.1), na.rm = TRUE),
                        include.lowest = TRUE)) %>%
  group_by(temp_bin) %>%
  summarise(avg_temperature = mean(env_temperature_c, na.rm = TRUE),
            avg_lateness    = mean(.data[[outcome_cont]], na.rm = TRUE),
            n = n(), .groups = "drop")
```

```{r ming-temperature-plot}
#| fig-width: 8
#| fig-height: 5

ggplot(temp_delay_summary, aes(x = avg_temperature, y = avg_lateness)) +
  geom_point(color = "#1A9988", size = 2.5) +
  geom_smooth(method = "lm", se = FALSE, color = "#EB5600") +
  labs(title = "Temperature and Average Lateness",
       x = "Average temperature (°C)", y = "Average lateness (minutes)") +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold"))
```

```{r ming-route-travel-time-plot}
#| fig-width: 8
#| fig-height: 5

if ("expected_water_min" %in% names(df_model)) {
  df_model %>%
    group_by(route_id) %>%
    summarise(avg_delay              = mean(.data[[outcome_cont]], na.rm = TRUE),
              avg_expected_water_min = mean(expected_water_min, na.rm = TRUE),
              n = n(), .groups = "drop") %>%
    ggplot(aes(x = avg_expected_water_min, y = avg_delay, label = route_id)) +
    geom_point(size = 3) +
    geom_text(nudge_y = 0.15) +
    geom_smooth(method = "lm", se = FALSE) +
    labs(title = "Route Travel Time vs Average Lateness",
         x = "Average expected water travel time (minutes)\n(proxy for route length / exposure)",
         y = "Average lateness (minutes)") +
    theme_minimal()
}
```

#### Monitorable Signals Beyond Prior Delay

Although operational propagation explains the largest share of delay variation, smaller signals are still useful when they are monitorable. Environmental conditions, time patterns, and vessel/load context may not dominate the model, but they describe conditions that operators can check before or during service.

```{r ming-monitorable-signals}
#| fig-width: 8
#| fig-height: 5

domain_e_plot_predictors <- intersect(c(
  domain_e_predictors, "prev_stop_boardings", "prev_stop_alightings",
  "prev_stop_onboard", "prev_stop_load_factor"
), names(df_model))

bce_predictors       <- unique(c(domain_b_predictors, domain_c_predictors, domain_e_plot_predictors))
plot_model_predictors <- unique(c(domain_a_predictors, domain_b_predictors, domain_c_predictors,
                                   domain_d_predictors, domain_e_plot_predictors))
numeric_predictors   <- plot_model_predictors[sapply(df_model[plot_model_predictors], is.numeric)]

df_model_std <- df_model %>%
  mutate(across(all_of(numeric_predictors), ~ as.numeric(scale(.x))))

combined_std_fit <- lm(
  as.formula(paste(outcome_cont, "~", paste(plot_model_predictors, collapse = " + "))),
  data = df_model_std
)

label_lookup <- c(
  env_temperature_c              = "Temperature",
  env_wind_speed_mps             = "Wind speed",
  env_visibility_m               = "Visibility",
  env_precipitation_mm           = "Precipitation",
  env_battery_water_level_obs    = "Water level",
  env_battery_water_level_anomaly = "Water-level anomaly",
  env_narrows_current_obs        = "Current",
  env_narrows_current_residual   = "Current residual",
  prev_stop_sched_hour           = "Previous scheduled hour",
  prev_stop_sched_hour_group     = "Previous scheduled hour group",
  schedule_season                = "Schedule season",
  schedule_daytype               = "Schedule day type",
  vessel_capacity                = "Vessel capacity",
  vessel_trip_seq_in_chain       = "Vessel trip sequence",
  prev_stop_boardings            = "Boardings",
  prev_stop_alightings           = "Alightings",
  prev_stop_onboard              = "Onboard count",
  prev_stop_load_factor          = "Load factor"
)

get_base_var <- function(term, vars) {
  hits <- vars[vapply(vars, function(v) startsWith(term, v), logical(1))]
  if (length(hits) == 0) return(NA_character_)
  hits[which.max(nchar(hits))]
}

coef_plot_df <- broom::tidy(combined_std_fit) %>%
  filter(term != "(Intercept)") %>%
  rowwise() %>%
  mutate(variable = get_base_var(term, bce_predictors)) %>%
  ungroup() %>%
  filter(!is.na(variable)) %>%
  mutate(
    domain = case_when(
      variable %in% domain_b_predictors       ~ "Environmental Conditions",
      variable %in% domain_c_predictors       ~ "Temporal Patterns",
      variable %in% domain_e_plot_predictors  ~ "Demand and Vessel Context",
      TRUE ~ "Other"
    )
  ) %>%
  filter(domain != "Other",
         !str_detect(term, "^schedule_season"),
         !variable %in% c("prev_stop_load_factor", "prev_stop_onboard", "vessel_trip_seq_in_chain")) %>%
  mutate(
    clean_label = if_else(
      variable %in% names(label_lookup),
      unname(label_lookup[variable]),
      variable
    ),
    clean_label = case_when(
      variable == "schedule_daytype"           ~ str_replace(term, variable, "Schedule day type "),
      variable == "prev_stop_sched_hour_group" ~ str_replace(term, variable, "Previous scheduled hour "),
      TRUE ~ clean_label
    ),
    abs_estimate = abs(estimate)
  )

coef_plot_top <- bind_rows(
  coef_plot_df %>% filter(domain == "Environmental Conditions") %>% slice_max(abs_estimate, n = 4, with_ties = FALSE),
  coef_plot_df %>% filter(domain == "Temporal Patterns")        %>% slice_max(abs_estimate, n = 3, with_ties = FALSE),
  coef_plot_df %>% filter(domain == "Demand and Vessel Context") %>% slice_max(abs_estimate, n = 4, with_ties = FALSE)
) %>%
  distinct(term, .keep_all = TRUE) %>%
  arrange(abs_estimate)

ggplot(coef_plot_top, aes(x = reorder(clean_label, abs_estimate), y = abs_estimate, fill = domain)) +
  geom_col(width = 0.72) +
  coord_flip() +
  scale_fill_manual(values = c(
    "Environmental Conditions"  = "#1A9988",
    "Temporal Patterns"         = "#7A7A7A",
    "Demand and Vessel Context" = "#EB5600"
  )) +
  labs(title = "Monitorable Signals Beyond Prior Delay",
       x = NULL, y = "Absolute standardized coefficient", fill = NULL) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom", panel.grid.major.y = element_blank(),
        plot.title = element_text(face = "bold"))
```

#### Operational Interpretation

The focused coefficient results and supporting plots reinforce three operational takeaways.

First, prior delay remains the strongest signal. This means delay management needs to account for accumulation across stops and linked trips, not only conditions at the current stop.

Second, route and stop structure matter. Some parts of the system appear to have a less favorable baseline, meaning they may be more vulnerable to delay even before short-run conditions are considered.

Third, weather, water, time, and vessel/load patterns are smaller signals, but they are monitorable. They can help operators understand when conditions may be less favorable and where extra attention may be useful.

## Predictive Model
**By: Guangze 'Simon' Sun**

### Feature Table Construction

The prediction feature table consolidates schedule structure, stop context, lagged demand, vessel traffic (AIS), and environmental conditions into a single modeling dataset. Each row represents one stop-level trip observation. The table below lists the required predictors; the build chunk is set to `eval: false` since it writes output files and depends on the full source parquet.

```{r simon-feature-paths}
#| eval: false

source_parquet <- "../exploratory_code/modeling/baseline_prediction_table_stop_level_v2.parquet"
out_csv     <- "../exploratory_code/modeling/final_predictive_feature_table.csv"
out_parquet <- "../exploratory_code/modeling/final_predictive_feature_table.parquet"
```

```{r simon-build-feature-table}
#| eval: false

feature_table <- arrow::read_parquet(source_parquet) %>% as_tibble()

required_predictors <- c(
  "alightings_lag_7d", "boardings_lag_7d", "curr_stop_shared_stop_flag",
  "curr_vessel_count_lag_7d", "direction", "expected_water_min",
  "is_terminal_stop", "planned_turnaround_gap_min", "prev_stop_sched_hour",
  "prev_stop_shared_stop_flag", "prev_vessel_count_lag_7d", "schedule_daytype",
  "schedule_season", "scheduled_headway_min", "stop_sequence", "vessel_capacity",
  "vessel_trip_seq_in_chain", "env_battery_water_level_obs", "env_narrows_current_obs",
  "env_precipitation_mm", "env_temperature_c", "env_visibility_m",
  "env_wind_speed_mps", "from_stop_id", "to_stop_id", "segment_hourly_vessel_count_all"
)

required_columns <- c("trip_date", "lateness_vs_original_schedule_min", required_predictors)
missing_columns  <- setdiff(required_columns, names(feature_table))
if (length(missing_columns) > 0) stop("Missing required columns: ", paste(missing_columns, collapse = ", "))
stopifnot(nrow(feature_table) > 0)
stopifnot(anyDuplicated(names(feature_table)) == 0)
stopifnot("segment_hourly_vessel_count_all" %in% names(feature_table))
stopifnot("prev_stop_sched_hour" %in% names(feature_table))

readr::write_csv(feature_table, out_csv)
arrow::write_parquet(feature_table, out_parquet)
stopifnot(file.exists(out_csv), file.exists(out_parquet))
```

### Baseline Model-Family Comparison

We used a chronological train / validation / test split. The most recent 30 days were held out as the test set, the previous 30 days were used as validation, and earlier observations were used for training. This matches the operational setting where future trips are predicted from historical data.

The baseline screen compared five model families: Logistic Regression, SVM, DNN, Random Forest, and XGBoost. SVM used a capped 20,000-row stratified training sample, so it should be interpreted as lightweight screening evidence. The baseline screen supported carrying forward Random Forest and XGBoost for focused finalist comparison.

```{r simon-paths}
#| message: false
#| warning: false

model_dir <- "../exploratory_code/modeling"

baseline_file   <- file.path(model_dir, "baseline_model_family_comparison.csv")
prediction_file <- file.path(model_dir, "final_rf_xgb_test_predictions.csv")
importance_file <- file.path(model_dir, "final_rf_top10_feature_importance.csv")

threshold_source_note <- "Test predictions were used for the recall >= 0.20 operating-point table because validation prediction probabilities were not available in the final delivery package."
rf_xgb_highlight_note <- "Yellow shading marks recall ranges where RF has higher precision than XGBoost and at least one model has precision >= 0.10."
```

```{r simon-baseline-plot}
#| warning: false
#| message: false
#| fig-cap: "Validation ROC-AUC from the baseline screen. SVM used a capped training sample and is included as lightweight screening evidence."

if (!file.exists(baseline_file)) {
  warning("Missing baseline comparison file: ", baseline_file)
} else {
  baseline_tbl <- read_csv(baseline_file, show_col_types = FALSE) %>%
    mutate(model_family = fct_reorder(model_family, validation_roc_auc))

  ggplot(baseline_tbl, aes(x = validation_roc_auc, y = model_family)) +
    geom_col(fill = "#1A9988", width = 0.68) +
    scale_x_continuous(labels = percent_format(accuracy = 1), limits = c(0, NA)) +
    labs(title = "Baseline Model-Family Screen", x = "Validation ROC-AUC", y = NULL) +
    theme_minimal(base_size = 15) +
    theme(plot.title = element_text(face = "bold", size = 18), panel.grid.minor = element_blank())
}
```

### Final RF vs XGBoost Comparison

After the baseline screen, we focused on Random Forest and XGBoost. Both model families can capture nonlinear relationships and interactions among schedule structure, stop context, lagged demand, vessel traffic, and environmental conditions.

The final tuning process used `trees = 1500` as the practical central setting. Around that tree count, RF tuning mainly adjusted `mtry` and `min_n`. XGBoost tuning adjusted tree depth, `min_n`, learning rate, loss reduction, and feature sampling. The final comparison retained one RF operational candidate and one XGBoost benchmark candidate.

XGBoost remains a strong predictive benchmark, but the dashboard use case emphasizes conservative alerting rather than global ranking alone. In the higher-precision region, Random Forest provides a more useful recall balance, so we select RF as the operational model and retain XGBoost as the benchmark.

```{r simon-pr-curve}
#| warning: false
#| message: false
#| fig-cap: "Yellow shading marks recall ranges where RF has higher precision than XGBoost."

if (!file.exists(prediction_file)) {
  warning("Missing final RF/XGB prediction file: ", prediction_file)
} else {
  pred_tbl <- read_csv(prediction_file, show_col_types = FALSE) %>%
    mutate(delay_gt_10 = factor(delay_gt_10, levels = c("no", "yes")))

  pr_tbl <- bind_rows(
    pred_tbl %>% transmute(model = "Final RF",  delay_gt_10, .pred_yes = rf_pred_yes)  %>%
      group_by(model) %>% pr_curve(truth = delay_gt_10, .pred_yes, event_level = "second"),
    pred_tbl %>% transmute(model = "Final XGB", delay_gt_10, .pred_yes = xgb_pred_yes) %>%
      group_by(model) %>% pr_curve(truth = delay_gt_10, .pred_yes, event_level = "second")
  )

  prep_curve <- function(dat, model_name) {
    dat %>%
      filter(model == model_name, is.finite(recall), is.finite(precision)) %>%
      group_by(recall) %>%
      summarise(precision = max(precision, na.rm = TRUE), .groups = "drop") %>%
      arrange(recall)
  }

  interp_precision <- function(curve, recall_grid) {
    approx(curve$recall, curve$precision, xout = recall_grid, rule = 2, ties = "ordered")$y
  }

  rf_curve  <- prep_curve(pr_tbl, "Final RF")
  xgb_curve <- prep_curve(pr_tbl, "Final XGB")
  max_shared_recall <- min(max(rf_curve$recall, na.rm = TRUE), max(xgb_curve$recall, na.rm = TRUE))
  recall_grid <- seq(0, max_shared_recall, length.out = 600)

  highlight_grid <- tibble(
    recall        = recall_grid,
    rf_precision  = interp_precision(rf_curve,  recall_grid),
    xgb_precision = interp_precision(xgb_curve, recall_grid)
  ) %>%
    mutate(
      high_precision_region = pmax(rf_precision, xgb_precision, na.rm = TRUE) >= 0.10,
      rf_above_xgb = rf_precision > xgb_precision,
      highlight    = rf_above_xgb & high_precision_region
    )

  if (!any(highlight_grid$highlight, na.rm = TRUE)) {
    highlight_grid <- highlight_grid %>% mutate(highlight = rf_above_xgb)
    rf_xgb_highlight_note <- "No RF-above-XGB intervals met the precision >= 0.10 rule, so the plot highlights all recall ranges where RF precision is above XGBoost."
  }

  highlight_intervals <- highlight_grid %>%
    mutate(run_id = cumsum(highlight != lag(highlight, default = first(highlight)))) %>%
    filter(highlight) %>%
    group_by(run_id) %>%
    summarise(xmin = min(recall), xmax = max(recall), .groups = "drop") %>%
    filter(xmax > xmin)

  ggplot(pr_tbl, aes(x = recall, y = precision, color = model)) +
    geom_rect(data = highlight_intervals,
              aes(xmin = xmin, xmax = xmax, ymin = 0, ymax = 1),
              inherit.aes = FALSE, fill = "#F2E8CF", alpha = 0.55) +
    geom_path(linewidth = 1.15) +
    scale_color_manual(values = c("Final RF" = "#D95F59", "Final XGB" = "#1A9988")) +
    scale_x_continuous(labels = percent_format(accuracy = 1), limits = c(0, 1)) +
    scale_y_continuous(labels = percent_format(accuracy = 1), limits = c(0, NA)) +
    labs(title = "Final RF vs XGBoost Precision-Recall Curve",
         x = "Recall", y = "Precision", color = NULL) +
    theme_minimal(base_size = 15) +
    theme(plot.title = element_text(face = "bold", size = 18),
          panel.grid.minor = element_blank(), legend.position = "bottom")
}
```

### Final RF Feature Importance

We inspected feature importance from the final Random Forest model to understand what types of advance-available information the operational model relies on. This importance is model-based and descriptive; it is not a causal estimate.

The top features combine environmental / water-condition variables and recurring operational patterns.

```{r simon-feature-importance}
#| warning: false
#| message: false

if (!file.exists(importance_file)) {
  warning("Missing RF feature importance file: ", importance_file)
} else {
  importance_tbl <- read_csv(importance_file, show_col_types = FALSE) %>%
    arrange(desc(importance)) %>%
    slice_head(n = 10) %>%
    mutate(
      display_label = dplyr::recode(
        feature,
        segment_hourly_vessel_count_all  = "Segment vessel count, 7-day lag",
        boardings_lag_7d                 = "Boardings, 7-day lag",
        alightings_lag_7d                = "Alightings, 7-day lag",
        curr_vessel_count_lag_7d         = "Current-stop vessel count, 7-day lag",
        prev_vessel_count_lag_7d         = "Previous-stop vessel count, 7-day lag",
        env_temperature_c                = "Temperature",
        env_battery_water_level_obs      = "Battery water level",
        env_narrows_current_obs          = "Narrows current",
        env_precipitation_mm             = "Precipitation",
        env_visibility_m                 = "Visibility",
        env_wind_speed_mps               = "Wind speed",
        expected_water_min               = "Expected water time",
        planned_turnaround_gap_min       = "Planned turnaround gap",
        prev_stop_sched_hour             = "Scheduled hour",
        .default = str_to_sentence(str_replace_all(feature, "_", " "))
      )
    )

  ggplot(importance_tbl, aes(x = importance, y = fct_reorder(display_label, importance))) +
    geom_col(fill = "#1A9988", width = 0.72) +
    labs(title = "Final RF Feature Importance", x = "Importance", y = NULL) +
    theme_minimal(base_size = 15) +
    theme(plot.title = element_text(face = "bold", size = 18), panel.grid.minor = element_blank())
}
```

The final RF relies on a mix of environmental conditions, water-current context, and recurring operational demand/vessel-pattern features.

# Interpretation and Implications




# Acknowledgements




# Code Appendix
Through this project process, we wrote a lot of code that is not meaningful toward our final results. You can view that below.



# References