---
title: "Associative Baseline Models (Simplified)"
output: html_document
author: "Project Team"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
options(scipen = 999, dplyr.summarise.inform = FALSE)
```

```{r packages}
required_pkgs <- c("dplyr", "tidyr", "arrow", "broom", "knitr", "readr")
missing_pkgs <- required_pkgs[!vapply(required_pkgs, requireNamespace, logical(1), quietly = TRUE)]
if (length(missing_pkgs) > 0) stop("Missing packages: ", paste(missing_pkgs, collapse = ", "))

library(dplyr)
library(tidyr)
library(arrow)
library(broom)
library(knitr)
library(readr)
```

## 1) Purpose and Setup

This section uses associative linear models to examine which factors are associated with ferry delay.

Because the goal is interpretation rather than forecasting, the models are estimated in-sample rather than in a train/test framework.

1. `baseline_linear_no_prior`
2. `baseline_linear_with_prior`

```{r settings}
set.seed(7900)

input_parquet <- "E:/MUSA/8010practicum/outputs/direct_pipeline_final_20260327/baseline_modeling_table_final_direct.parquet"
output_dir <- "E:/MUSA/8010practicum/outputs/direct_pipeline_final_20260327/associative_baseline_models_20260404"
dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)

outcome_cont <- "lateness_vs_original_schedule_min"


# Weekly local-time-slot imputation:
# interpreted as a 7-day centered window (±3 days) by local trip date.
week_window_days <- 3

base_predictors <- c(
  "route_id", "direction", "stop_sequence", "prev_stop_sched_hour", "prev_stop_sched_hour_group",
  "schedule_season", "schedule_daytype", "vessel_trip_seq_in_chain", "vessel_capacity",
  "is_terminal_stop", "prev_stop_shared_stop_flag", "curr_stop_shared_stop_flag",
  "planned_turnaround_gap_min", "scheduled_headway_min", "expected_water_min",
  "prev_stop_sequence", "prev_stop_boardings", "prev_stop_alightings", "prev_stop_onboard",
  "prev_stop_load_factor", "env_temperature_c", "env_wind_speed_mps", "env_visibility_m",
  "env_precipitation_mm", "env_battery_water_level_anomaly", "env_narrows_current_residual",
  "env_battery_water_level_obs", "env_narrows_current_obs",
  "prev_stop_hourly_vessel_count_all", "curr_stop_hourly_vessel_count_all"
)
prior_delay_predictors <- c("inherited_prev_trip_delay", "carried_within_trip_delay")
```

## 2) Data Loading and Variable Setup

The modeling table is loaded from the final baseline dataset, and trip date is parsed from the trip key for downstream processing.

```{r load, include=FALSE}
if (!file.exists(input_parquet)) stop("Missing input parquet: ", input_parquet)
df_raw <- read_parquet(input_parquet) %>% as_tibble()

df <- df_raw %>%
  mutate(
    trip_date = as.Date(substr(as.character(trip_key), 1, 8), format = "%Y%m%d")
  )

predictors_no_prior <- intersect(base_predictors, names(df))
predictors_with_prior <- unique(c(predictors_no_prior, intersect(prior_delay_predictors, names(df))))

tibble(
  input_rows = nrow(df_raw),
  rows_after_load = nrow(df),
  predictors_no_prior_n = length(predictors_no_prior),
  predictors_with_prior_n = length(predictors_with_prior)
) %>% kable(caption = "Data and Predictor Setup")
```

## 3) AIS/Environment Missing-Value Imputation Rule
Missing values in AIS and environmental variables are imputed using a local time-slot based rolling rule. We first use the same scheduled hour within a centered 7-day window, and then fall back to the same hour group when needed.

```{r impute, include=FALSE}
build_week_fill_map <- function(day_slot_tbl, group_col, date_col = "trip_date", value_col = "slot_day_median", window_days = 3) {
  if (nrow(day_slot_tbl) == 0) return(tibble())
  split_tbl <- split(day_slot_tbl, day_slot_tbl[[group_col]], drop = TRUE)
  out <- lapply(split_tbl, function(gdf) {
    gdf <- gdf[order(gdf[[date_col]]), , drop = FALSE]
    d <- gdf[[date_col]]
    v <- gdf[[value_col]]
    week_fill <- vapply(seq_along(v), function(i) {
      idx <- abs(as.integer(d - d[i])) <= window_days
      val <- suppressWarnings(median(v[idx], na.rm = TRUE))
      ifelse(is.finite(val), val, NA_real_)
    }, numeric(1))
    gdf$week_fill <- week_fill
    gdf[, c(group_col, date_col, "week_fill"), drop = FALSE]
  })
  bind_rows(out)
}

impute_by_local_slot_week <- function(data, var_name, window_days = 3) {
  x <- data
  miss_before <- sum(is.na(x[[var_name]]))
  if (miss_before == 0) {
    return(list(data = x, summary = tibble(variable = var_name, n_missing_before = 0L, n_filled_primary = 0L, n_filled_fallback = 0L, n_missing_after = 0L)))
  }

  # Primary: same local hour within centered 7-day window (±3 days by default).
  primary_daily <- x %>%
    filter(!is.na(.data[[var_name]]), !is.na(trip_date), !is.na(prev_stop_sched_hour)) %>%
    group_by(prev_stop_sched_hour, trip_date) %>%
    summarise(slot_day_median = median(.data[[var_name]], na.rm = TRUE), .groups = "drop")
  primary_fill <- build_week_fill_map(primary_daily, group_col = "prev_stop_sched_hour", value_col = "slot_day_median", window_days = window_days) %>%
    rename(fill_primary = week_fill)

  # Fallback: same local hour-group within centered 7-day window.
  fallback_daily <- x %>%
    filter(!is.na(.data[[var_name]]), !is.na(trip_date), !is.na(prev_stop_sched_hour_group)) %>%
    group_by(prev_stop_sched_hour_group, trip_date) %>%
    summarise(slot_day_median = median(.data[[var_name]], na.rm = TRUE), .groups = "drop")
  fallback_fill <- build_week_fill_map(fallback_daily, group_col = "prev_stop_sched_hour_group", value_col = "slot_day_median", window_days = window_days) %>%
    rename(fill_fallback = week_fill)

  x <- x %>%
    left_join(primary_fill, by = c("prev_stop_sched_hour", "trip_date")) %>%
    left_join(fallback_fill, by = c("prev_stop_sched_hour_group", "trip_date"))

  was_missing <- is.na(x[[var_name]])
  fill_primary_idx <- was_missing & !is.na(x$fill_primary)
  fill_fallback_idx <- was_missing & is.na(x$fill_primary) & !is.na(x$fill_fallback)

  x[[var_name]][fill_primary_idx] <- x$fill_primary[fill_primary_idx]
  x[[var_name]][fill_fallback_idx] <- x$fill_fallback[fill_fallback_idx]
  x <- x %>% select(-fill_primary, -fill_fallback)

  list(
    data = x,
    summary = tibble(
      variable = var_name,
      n_missing_before = miss_before,
      n_filled_primary = sum(fill_primary_idx, na.rm = TRUE),
      n_filled_fallback = sum(fill_fallback_idx, na.rm = TRUE),
      n_missing_after = sum(is.na(x[[var_name]]))
    )
  )
}

impute_vars <- intersect(
  c(
    names(df)[grepl("^env_", names(df))],
    "prev_stop_hourly_vessel_count_all",
    "curr_stop_hourly_vessel_count_all"
  ),
  names(df)
)

impute_summary_list <- list()
df_imp <- df
for (v in impute_vars) {
  res <- impute_by_local_slot_week(df_imp, v, window_days = week_window_days)
  df_imp <- res$data
  impute_summary_list[[length(impute_summary_list) + 1]] <- res$summary
}
imputation_summary <- bind_rows(impute_summary_list)
```

## 4) Helper Functions

```{r helpers, include=FALSE}
prepare_model_df <- function(data, outcome, predictors) {
  data %>% select(all_of(c(outcome, predictors))) %>% drop_na()
}

fit_linear <- function(data, model_name, outcome, predictors) {
  d <- prepare_model_df(data, outcome, predictors)
  fml <- as.formula(paste(outcome, "~", paste(predictors, collapse = " + ")))
  fit <- lm(fml, data = d)
  pred <- predict(fit, newdata = d)
  rmse <- sqrt(mean((d[[outcome]] - pred)^2))
  g <- glance(fit)
  list(
    metrics = tibble(
      model_name = model_name, model_type = "linear", prior_delay_included = ifelse(grepl("with_prior", model_name), "yes", "no"),
      n = nrow(d), aic = g$AIC, bic = g$BIC, rsq = g$r.squared, adj_rsq = g$adj.r.squared, rmse = rmse,
      pseudo_rsq = NA_real_, accuracy = NA_real_, sensitivity = NA_real_, specificity = NA_real_, auc = NA_real_
    ),
    coefs = tidy(fit) %>% mutate(model_name = model_name, model_type = "linear"),
    formula_text = paste(deparse(fml), collapse = "")
  )
}

```

## 5) Fit Two Baseline Linear Models

```{r fit}
df_model <- df_imp 

model_results <- list(
  baseline_linear_no_prior = fit_linear(df_model, "baseline_linear_no_prior", outcome_cont, predictors_no_prior),
  baseline_linear_with_prior = fit_linear(df_model, "baseline_linear_with_prior", outcome_cont, predictors_with_prior)
)
```

## 6) Main Metrics Table

```{r metrics}
model_comparison_summary <- bind_rows(lapply(model_results, `[[`, "metrics")) %>%
  arrange(prior_delay_included, model_name)
model_comparison_summary %>% kable(digits = 4, caption = "Main Metrics for Baseline Linear Models")
```

## 7) Interpretable Coefficients from the Associative Model

```{r coefs}
model_coefficients_all <- bind_rows(lapply(model_results, `[[`, "coefs"))

coef_table_display <- model_coefficients_all %>%
  filter(
    model_name == "baseline_linear_with_prior",
    term != "(Intercept)"
  ) %>%
  mutate(abs_estimate = abs(estimate)) %>%
  arrange(desc(abs_estimate)) %>%
  slice_head(n = 12) %>%
  select(term, estimate, std.error, statistic, p.value)

coef_table_display %>%
  kable(digits = 4, caption = "Selected Coefficients: Baseline Linear Model with Prior Delay")
```

## 8) Save Outputs

```{r save}
write_csv(model_comparison_summary, file.path(output_dir, "model_comparison_summary.csv"))
write_csv(model_coefficients_all, file.path(output_dir, "model_coefficients_all.csv"))
write_csv(coef_table_display, file.path(output_dir, "selected_coefficients_with_prior.csv"))
write_csv(imputation_summary, file.path(output_dir, "ais_env_imputation_summary.csv"))

tibble(
  file = c(
    "model_comparison_summary.csv",
    "model_coefficients_all.csv",
    "selected_coefficients_with_prior.csv",
    "ais_env_imputation_summary.csv"
  )
) %>% kable(caption = "Saved Outputs")
```
