---
title: "NYC Ferry Direct Pipeline to Final Tables"
output:
  html_document:
    toc: true
    toc_depth: 3
---

```{r setup, message=FALSE, warning=FALSE}
suppressPackageStartupMessages({
  library(dplyr)
  library(tidyr)
  library(readr)
  library(stringr)
  library(lubridate)
  library(purrr)
  library(arrow)
  library(knitr)
})
options(scipen = 999)

TZ_NY <- "America/New_York"
TZ_UTC <- "UTC"
trip_data_dir <- "C:/Data/Spring 2026/CPLN 7900/trip_data"
out_dir <- "C:/Data/Spring 2026/CPLN 7900/explore/outputs/direct_pipeline_final_20260327"
dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)

path_trip_2024 <- file.path(trip_data_dir, "NYCF_trip_data_2024.csv")
path_trip_2025 <- file.path(trip_data_dir, "NYCF_trip_data_2025.01_06.csv")
path_stop_2024 <- file.path(trip_data_dir, "NYCF_stop_data_2024.csv")
path_stop_2025 <- file.path(trip_data_dir, "NYCF_stop_data_2025.01_06.csv")
path_env <- file.path(trip_data_dir, "environment_utc_hourly_2024_2025.csv")
path_ais <- file.path(trip_data_dir, "AIS_landing_features.csv")

out_baseline <- file.path(out_dir, "baseline_modeling_table_final_direct.csv")
out_decomp <- file.path(out_dir, "stop_level_arrival_decomposition_table_final_direct.csv")
out_lookup <- file.path(out_dir, "lookup.csv")

out_baseline_parquet <- file.path(out_dir, "baseline_modeling_table_final_direct.parquet")
out_decomp_parquet <- file.path(out_dir, "stop_level_arrival_decomposition_table_final_direct.parquet")
out_lookup_parquet <- file.path(out_dir, "lookup.parquet")
```

## Purpose

This notebook rebuilds the pipeline directly from `trip_data` to:

1. Final baseline modeling table  
2. Final lean stop-level arrival decomposition table  
3. One merged lookup table (optional but kept for auditability)  

No extra intermediate output tables are written.

## Source Files

```{r}
source_files <- tibble(
  file = c(path_trip_2024, path_trip_2025, path_stop_2024, path_stop_2025, path_env, path_ais),
  exists = file.exists(c(path_trip_2024, path_trip_2025, path_stop_2024, path_stop_2025, path_env, path_ais))
)
print(source_files)
stopifnot(all(source_files$exists))
```

## Helper Functions

```{r}
clean_names_simple <- function(nm) {
  nm %>%
    str_replace_all("\ufeff", "") %>% str_trim() %>% str_to_lower() %>%
    str_replace_all("[^a-z0-9]+", "_") %>% str_replace_all("_+", "_") %>% str_replace_all("^_|_$", "")
}

parse_trip_date <- function(x) as.Date(parse_date_time(as.character(x), orders = c("mdY", "Ymd"), quiet = TRUE))
normalize_trip_no <- function(x) suppressWarnings(as.numeric(as.character(x)))

parse_datetime_with_default_tz <- function(x, default_tz) {
  x_chr <- if (inherits(x, "POSIXt")) {
    # Preserve source clock time and then assign the intended source timezone explicitly.
    format(x, format = "%Y-%m-%d %H:%M:%S", tz = "UTC")
  } else {
    as.character(x)
  }
  x_chr <- str_trim(x_chr)
  x_chr[x_chr %in% c("", "NA", "NaN", "NAT")] <- NA_character_
  has_tz <- str_detect(x_chr, "(Z$|[+-][0-9]{2}:?[0-9]{2}$|\\sUTC$|\\sEST$|\\sEDT$)")
  has_tz[is.na(has_tz)] <- FALSE
  out <- as.POSIXct(rep(NA, length(x_chr)), origin = "1970-01-01", tz = default_tz)
  if (any(!has_tz, na.rm = TRUE)) {
    out[!has_tz] <- suppressWarnings(parse_date_time(
      x_chr[!has_tz], orders = c("mdY HMS", "mdY HM", "mdY IMSp", "mdY IMp", "Ymd HMS", "Ymd HM"), tz = default_tz, quiet = TRUE
    ))
  }
  if (any(has_tz, na.rm = TRUE)) {
    dt_utc <- suppressWarnings(parse_date_time(
      x_chr[has_tz], orders = c("mdY HMS z", "mdY HM z", "Ymd HMS z", "Ymd HM z", "Ymd HMS", "mdY HMS"), tz = "UTC", quiet = TRUE
    ))
    out[has_tz] <- with_tz(dt_utc, tzone = default_tz)
  }
  out
}

parse_ny_datetime <- function(x) parse_datetime_with_default_tz(x, TZ_NY)
parse_utc_datetime <- function(x) parse_datetime_with_default_tz(x, TZ_UTC)

add_trip_keys <- function(df, date_col = "trip_date", trip_no_col = "trip_no", route_col = "route_id", direction_col = "direction") {
  date_vec <- df[[date_col]]
  trip_no_vec <- df[[trip_no_col]]
  route_vec <- df[[route_col]]
  direction_vec <- df[[direction_col]]
  trip_no_valid <- !is.na(trip_no_vec) & trip_no_vec >= 0
  trip_key <- ifelse(
    trip_no_valid & !is.na(date_vec),
    paste0(format(date_vec, "%Y%m%d"), as.character(as.integer(trip_no_vec))),
    NA_character_
  )
  full_trip_key <- ifelse(
    trip_no_valid & !is.na(date_vec),
    paste(format(date_vec, "%Y-%m-%d"), as.character(as.integer(trip_no_vec)), route_vec, direction_vec, sep = "|"),
    NA_character_
  )
  mutate(df, trip_no_valid = trip_no_valid, trip_key = trip_key, full_trip_key = full_trip_key)
}

apply_lookup_group_rules <- function(df) {
  df %>%
    mutate(
      lk_schedule_season = if_else(route_id == "GI" & lk_schedule_season == "SPRING", "SUMMER_1", lk_schedule_season),
      lk_schedule_season = if_else(route_id == "GI" & lk_schedule_season == "FALL", "SUMMER_2", lk_schedule_season),
      lk_dep_hour_group = if_else(lk_schedule_daytype == "WEEKEND" & lk_dep_hour_group == "5_8", "9_11", lk_dep_hour_group),
      lk_schedule_season = if_else(route_id == "ER" & lk_stop_id == 90L, "ALL", lk_schedule_season),
      lk_dep_hour_group = if_else(route_id == "ER" & lk_stop_id == 90L, "ALL", lk_dep_hour_group),
      lk_stop_id = if_else(route_id == "ER" & lk_stop_id == 90L, -1L, lk_stop_id),
      lk_schedule_season = if_else(route_id == "RR", "ALL", lk_schedule_season),
      lk_schedule_daytype = if_else(route_id == "RR", "ALL", lk_schedule_daytype),
      lk_dep_hour_group = if_else(route_id == "RR", "ALL", lk_dep_hour_group),
      lk_stop_id = if_else(route_id == "RR", -1L, lk_stop_id)
    )
}

build_hourly_ais_utc <- function(path_ais) {
  read_csv(path_ais, show_col_types = FALSE) %>%
    transmute(
      stop_id = suppressWarnings(as.integer(stop_id)),
      ais_date = as.Date(date),
      ais_hour = suppressWarnings(as.integer(hour)),
      vessel_count = suppressWarnings(as.numeric(vessel_count))
    ) %>%
    mutate(
      ais_hour_start_utc = if_else(
        !is.na(ais_date) & !is.na(ais_hour),
        as.POSIXct(ais_date, tz = TZ_UTC) + hours(ais_hour),
        as.POSIXct(NA, origin = "1970-01-01", tz = TZ_UTC)
      ),
      ais_hour_start_utc = floor_date(ais_hour_start_utc, unit = "hour")
    ) %>%
    group_by(stop_id, ais_hour_start_utc) %>%
    summarise(vessel_count = sum(vessel_count, na.rm = TRUE), .groups = "drop")
}

build_hourly_env_utc <- function(path_env) {
  env_raw <- read_csv(path_env, show_col_types = FALSE)
  names(env_raw) <- clean_names_simple(names(env_raw))

  hour_from_hour_col_utc <- if ("hour_utc" %in% names(env_raw)) parse_utc_datetime(env_raw$hour_utc) else rep(as.POSIXct(NA, origin = "1970-01-01", tz = TZ_UTC), nrow(env_raw))
  hour_from_obs_col_utc <- if ("obs_hour_utc" %in% names(env_raw)) parse_utc_datetime(env_raw$obs_hour_utc) else rep(as.POSIXct(NA, origin = "1970-01-01", tz = TZ_UTC), nrow(env_raw))
  hour_raw <- coalesce(hour_from_hour_col_utc, hour_from_obs_col_utc)

  # Midnight repair: when daily 00:00 UTC is missing in source hour fields, recover it from row order
  # using neighboring timestamps (prev + 1 hour when prev=23:00 and next=01:00 pattern).
  prev_hour <- lag(hour_raw)
  next_hour <- lead(hour_raw)
  recover_midnight <- is.na(hour_raw) & !is.na(prev_hour) & !is.na(next_hour) &
    hour(prev_hour) == 23 & hour(next_hour) == 1 &
    as.numeric(difftime(next_hour, prev_hour, units = "hours")) == 2
  hour_fixed <- if_else(recover_midnight, prev_hour + hours(1), hour_raw)

  env_utc <- env_raw %>%
    mutate(
      env_hour_start_utc = floor_date(hour_fixed, unit = "hour"),
      env_temperature_c = suppressWarnings(as.numeric(temperature_c)),
      env_wind_speed_mps = suppressWarnings(as.numeric(wind_speed_mps)),
      env_visibility_m = suppressWarnings(as.numeric(visibility_m)),
      env_precipitation_mm = suppressWarnings(as.numeric(precipitation_mm)),
      env_battery_water_level_anomaly = suppressWarnings(as.numeric(battery_water_level_anomaly)),
      env_narrows_current_residual = suppressWarnings(as.numeric(narrows_current_residual)),
      env_battery_water_level_obs = suppressWarnings(as.numeric(battery_water_level_obs)),
      env_narrows_current_obs = suppressWarnings(as.numeric(narrows_current_obs))
    ) %>%
    select(env_hour_start_utc, starts_with("env_"))

  env_sentinel_cols <- setdiff(names(env_utc)[grepl("^env_", names(env_utc))], "env_hour_start_utc")
  weather_clean <- replace_weather_sentinels(env_utc, env_sentinel_cols)
  env_value_cols <- setdiff(names(weather_clean$data)[grepl("^env_", names(weather_clean$data))], "env_hour_start_utc")
  weather_clean$data %>%
    group_by(env_hour_start_utc) %>%
    summarise(across(all_of(env_value_cols), ~ mean(.x, na.rm = TRUE)), .groups = "drop")
}

hour_group_5bin <- function(h) {
  h_num <- suppressWarnings(as.numeric(h))
  out <- rep(NA_character_, length(h_num))
  out[h_num >= 5 & h_num <= 8] <- "5_8"
  out[h_num >= 9 & h_num <= 11] <- "9_11"
  out[h_num >= 12 & h_num <= 14] <- "12_14"
  out[h_num >= 15 & h_num <= 17] <- "15_17"
  out[h_num >= 18 & h_num <= 22] <- "18_22"
  out
}

segment_id_3digit <- function(from_stop_id, to_stop_id) {
  a <- suppressWarnings(as.integer(from_stop_id))
  b <- suppressWarnings(as.integer(to_stop_id))
  ifelse(is.na(a) | is.na(b), NA_character_, paste0(sprintf("%03d", a), "_", sprintf("%03d", b)))
}

safe_mode <- function(x) {
  x <- x[!is.na(x)]
  if (length(x) == 0) return(NA)
  ux <- unique(x)
  ux[which.max(tabulate(match(x, ux)))]
}

parse_variance_seconds <- function(x) {
  parse_one <- function(s) {
    s <- str_trim(as.character(s))
    if (is.na(s) || s %in% c("", "NA", "NaN", "NAT", "No_Data", "NO_DATA", "No Data", "NULL", "null")) return(NA_real_)
    sign <- 1
    if (str_starts(s, "-")) {
      sign <- -1
      s <- str_sub(s, 2)
    } else if (str_starts(s, "\\+")) {
      s <- str_sub(s, 2)
    }
    if (str_detect(s, "^\\d+(\\.\\d+)?$")) return(sign * as.numeric(s))
    if (str_detect(s, "^\\d{1,3}:\\d{1,2}(:\\d{1,2})?$")) {
      parts <- str_split(s, ":", simplify = TRUE)
      if (ncol(parts) == 2) {
        mm <- suppressWarnings(as.numeric(parts[1]))
        ss <- suppressWarnings(as.numeric(parts[2]))
        if (!is.na(mm) && !is.na(ss)) return(sign * (mm * 60 + ss))
      }
      if (ncol(parts) == 3) {
        hh <- suppressWarnings(as.numeric(parts[1]))
        mm <- suppressWarnings(as.numeric(parts[2]))
        ss <- suppressWarnings(as.numeric(parts[3]))
        if (!is.na(hh) && !is.na(mm) && !is.na(ss)) return(sign * (hh * 3600 + mm * 60 + ss))
      }
    }
    NA_real_
  }
  vapply(x, parse_one, numeric(1))
}

reconstruct_with_variance_sign <- function(reference_time, actual_truncated, variance_seconds) {
  n <- length(reference_time)
  out_time <- as.POSIXct(rep(NA_real_, n), origin = "1970-01-01", tz = TZ_NY)
  sign_rule <- rep(NA_character_, n)

  mag <- abs(variance_seconds)
  plus <- reference_time + as.difftime(mag, units = "secs")
  minus <- reference_time - as.difftime(mag, units = "secs")

  trunc_floor <- floor_date(actual_truncated, unit = "minute")
  plus_match <- !is.na(plus) & !is.na(trunc_floor) & floor_date(plus, unit = "minute") == trunc_floor
  minus_match <- !is.na(minus) & !is.na(trunc_floor) & floor_date(minus, unit = "minute") == trunc_floor

  plus_only <- plus_match & !minus_match
  minus_only <- minus_match & !plus_match
  out_time[plus_only] <- plus[plus_only]
  out_time[minus_only] <- minus[minus_only]
  sign_rule[plus_only] <- "plus_match_trunc"
  sign_rule[minus_only] <- "minus_match_trunc"

  unresolved <- !is.na(reference_time) & !is.na(mag) & is.na(out_time)
  if (any(unresolved)) {
    plus_diff <- abs(as.numeric(difftime(plus[unresolved], actual_truncated[unresolved], units = "secs")))
    minus_diff <- abs(as.numeric(difftime(minus[unresolved], actual_truncated[unresolved], units = "secs")))
    choose_plus <- plus_diff <= minus_diff
    idx <- which(unresolved)
    out_time[idx[choose_plus]] <- plus[idx[choose_plus]]
    out_time[idx[!choose_plus]] <- minus[idx[!choose_plus]]
    sign_rule[idx[choose_plus]] <- "plus_nearest_trunc"
    sign_rule[idx[!choose_plus]] <- "minus_nearest_trunc"
  }

  # If truncated raw time is unavailable, fall back to signed variance convention in source.
  no_trunc_fallback <- !is.na(reference_time) & !is.na(variance_seconds) & is.na(actual_truncated) & is.na(out_time)
  if (any(no_trunc_fallback)) {
    out_time[no_trunc_fallback] <- reference_time[no_trunc_fallback] + as.difftime(variance_seconds[no_trunc_fallback], units = "secs")
    sign_rule[no_trunc_fallback] <- ifelse(variance_seconds[no_trunc_fallback] >= 0, "plus_from_signed_variance_no_trunc", "minus_from_signed_variance_no_trunc")
  }

  list(
    time = out_time,
    sign_rule = sign_rule,
    sign_from_trunc = sign_rule %in% c("plus_match_trunc", "minus_match_trunc", "plus_nearest_trunc", "minus_nearest_trunc")
  )
}

replace_weather_sentinels <- function(df, weather_cols) {
  sentinel <- c(999, 999.0, 999.9, 9999, 99999, 999999, -999, -999.0, -999.9, -9999, -99999, -999999)
  out <- df
  log_tbl <- list()
  for (col in intersect(weather_cols, names(out))) {
    v <- suppressWarnings(as.numeric(out[[col]]))
    hit <- !is.na(v) & v %in% sentinel
    n_hit <- sum(hit, na.rm = TRUE)
    if (n_hit > 0) v[hit] <- NA_real_
    out[[col]] <- v
    log_tbl[[length(log_tbl) + 1]] <- tibble(column_name = col, sentinel_replacements = n_hit)
  }
  list(data = out, log = bind_rows(log_tbl))
}
```

## Ingest and Harmonize Trip/Stop Inputs

```{r}
trip_raw <- bind_rows(
  read_csv(path_trip_2024, show_col_types = FALSE),
  read_csv(path_trip_2025, show_col_types = FALSE)
)
names(trip_raw) <- clean_names_simple(names(trip_raw))

schedule_regime_vec <- if ("trip_schedule_regime" %in% names(trip_raw)) {
  trip_raw$trip_schedule_regime
} else if ("trip_schedule" %in% names(trip_raw)) {
  trip_raw$trip_schedule
} else if ("schedule" %in% names(trip_raw)) {
  trip_raw$schedule
} else {
  rep(NA_character_, nrow(trip_raw))
}

stop_raw <- bind_rows(
  read_csv(path_stop_2024, show_col_types = FALSE),
  read_csv(path_stop_2025, show_col_types = FALSE)
)
names(stop_raw) <- clean_names_simple(names(stop_raw))

stop_tbl <- stop_raw %>%
  transmute(
    trip_date = parse_trip_date(trip_date),
    trip_no = normalize_trip_no(trip_no),
    route_id = toupper(str_trim(as.character(route_no))),
    direction = toupper(str_trim(as.character(route_direction))),
    stop_id = suppressWarnings(as.integer(stop_id)),
    stop_name = as.character(stop_name),
    stop_sequence = suppressWarnings(as.numeric(stop_sequence)),
    scheduled_arrival_time = parse_ny_datetime(scheduled_arrival_dt),
    scheduled_departure_time = parse_ny_datetime(scheduled_departure_dt),
    actual_arrival_time_trunc = parse_ny_datetime(actual_arrival_dt),
    actual_departure_time_trunc = parse_ny_datetime(actual_departure_dt),
    arrival_variance_raw = as.character(arrivaltimevariance),
    departure_variance_raw = as.character(departuretimevariance),
    boarding_wake = suppressWarnings(as.numeric(boarding_wake)),
    alightings_wake = suppressWarnings(as.numeric(alightings_wake)),
    vehicle_id = as.character(vehicle_id)
  ) %>%
  mutate(
    scheduled_time_original = coalesce(scheduled_departure_time, scheduled_arrival_time)
  ) %>%
  arrange(trip_date, trip_no, route_id, direction, stop_sequence) %>%
  group_by(trip_date, trip_no, route_id, direction) %>%
  mutate(
    onboard_raw_unclipped = cumsum(coalesce(boarding_wake, 0)) - cumsum(coalesce(alightings_wake, 0)),
    onboard_raw = pmax(onboard_raw_unclipped, 0)
  ) %>%
  ungroup()

trip_tbl <- trip_raw %>%
  transmute(
    trip_date = parse_trip_date(trip_date),
    trip_no = normalize_trip_no(trip_no),
    route_id = toupper(str_trim(as.character(route_no))),
    direction = toupper(str_trim(as.character(route_direction))),
    vessel_operation_class = as.character(vessel_operation_class),
    trip_schedule_regime = as.character(schedule_regime_vec),
    vessel_id_trip = as.character(vehicle_id)
  )
```

## Trip Key, Invalid Trip Filtering, and Schedule Mapping

```{r}
stop_tbl <- stop_tbl %>%
  add_trip_keys()

trip_tbl <- trip_tbl %>%
  add_trip_keys()

stop_tbl <- stop_tbl %>% filter(trip_no_valid, !is.na(trip_key), !is.na(full_trip_key))
trip_tbl <- trip_tbl %>% filter(trip_no_valid, !is.na(trip_key), !is.na(full_trip_key))

trip_meta <- trip_tbl %>%
  group_by(full_trip_key, trip_key) %>%
  summarise(
    vessel_operation_class = safe_mode(vessel_operation_class),
    trip_schedule_regime = safe_mode(trip_schedule_regime),
    vessel_id_trip = safe_mode(vessel_id_trip),
    vessel_capacity_from_class = suppressWarnings(as.numeric(str_extract(safe_mode(vessel_operation_class), "\\d+"))),
    .groups = "drop"
  )

stop_tbl <- stop_tbl %>%
  left_join(trip_meta %>% select(full_trip_key, vessel_operation_class, vessel_capacity_from_class, trip_schedule_regime, vessel_id_trip), by = "full_trip_key") %>%
  mutate(
    vessel_id = coalesce(vehicle_id, vessel_id_trip),
    vessel_capacity = vessel_capacity_from_class,
    regime_l = str_to_lower(coalesce(trip_schedule_regime, "")),
    dow = wday(trip_date, week_start = 1),
    is_weekend_date = dow %in% c(6, 7),
    # Holiday is folded into WEEKEND (no separate HOLIDAY category).
    schedule_daytype = case_when(
      str_detect(regime_l, "weekend") ~ "WEEKEND",
      str_detect(regime_l, "modified|holiday|passover|sukkot|memorial") ~ "WEEKEND",
      is_weekend_date ~ "WEEKEND",
      TRUE ~ "WEEKDAY"
    ),
    season_tmp = case_when(
      str_detect(regime_l, "winter") ~ "WINTER",
      str_detect(regime_l, "spring|passover") ~ "SPRING",
      str_detect(regime_l, "fall|sukkot") ~ "FALL",
      str_detect(regime_l, "summer|early summer|late summer|memorial") ~ "SUMMER",
      TRUE ~ NA_character_
    ),
    mm = month(trip_date),
    summer_like = season_tmp == "SUMMER" | str_detect(regime_l, "summer|early summer|late summer|memorial"),
    schedule_season = case_when(
      !summer_like ~ season_tmp,
      summer_like & mm <= 6 ~ "SUMMER_1",
      summer_like & mm >= 7 ~ "SUMMER_2",
      TRUE ~ season_tmp
    )
  ) %>%
  select(-vessel_id_trip)

stop_tbl <- stop_tbl %>%
  filter(!(trip_date == as.Date("2024-04-25") & route_id == "SB"))
```

## Time Refinement (2024 Variance Reconstruction) and Origin Hour

```{r}
# Reconstruct second-level actual times using variance fields + truncated displayed raw times.
# Sign is not hard-coded: we test +/- variance and choose the option that matches minute truncation of
# the displayed raw timestamp. This preserves the intended full-precision timestamp reconstruction.
arrival_var_sec <- parse_variance_seconds(stop_tbl$arrival_variance_raw)
departure_var_sec <- parse_variance_seconds(stop_tbl$departure_variance_raw)

arr_recon <- reconstruct_with_variance_sign(
  reference_time = stop_tbl$scheduled_arrival_time,
  actual_truncated = stop_tbl$actual_arrival_time_trunc,
  variance_seconds = arrival_var_sec
)
dep_recon <- reconstruct_with_variance_sign(
  reference_time = stop_tbl$scheduled_departure_time,
  actual_truncated = stop_tbl$actual_departure_time_trunc,
  variance_seconds = departure_var_sec
)

stop_tbl <- stop_tbl %>%
  mutate(
    actual_arrival_time_from_variance = arr_recon$time,
    actual_departure_time_from_variance = dep_recon$time,
    actual_arrival_time = coalesce(actual_arrival_time_from_variance, actual_arrival_time_trunc),
    actual_departure_time = coalesce(actual_departure_time_from_variance, actual_departure_time_trunc)
  )

stop_tbl <- stop_tbl %>%
  arrange(trip_key, stop_sequence) %>%
  group_by(trip_key) %>%
  mutate(
    origin_stop_sequence = min(stop_sequence, na.rm = TRUE),
    terminal_stop_sequence = max(stop_sequence, na.rm = TRUE),
    stop_rank = row_number() - 1L
  ) %>%
  ungroup()

origin_hour_map <- stop_tbl %>%
  group_by(trip_key) %>%
  arrange(stop_sequence, .by_group = TRUE) %>%
  summarise(dep_hour_origin = hour(first(scheduled_time_original)), .groups = "drop")

stop_tbl <- stop_tbl %>%
  left_join(origin_hour_map, by = "trip_key") %>%
  mutate(dep_hour_origin = as.numeric(dep_hour_origin))
```

## One Merged Lookup (Median Offsets on Merged Groups)

```{r}
stop_tbl <- stop_tbl %>%
  mutate(
    arr_offset_min = as.numeric(difftime(actual_arrival_time, scheduled_time_original, units = "mins")),
    dep_offset_min = as.numeric(difftime(actual_departure_time, scheduled_time_original, units = "mins"))
  )

lookup_base <- stop_tbl %>%
  mutate(
    lk_schedule_season = coalesce(schedule_season, "UNK"),
    lk_schedule_daytype = coalesce(schedule_daytype, "UNK"),
    lk_dep_hour_group = case_when(
      dep_hour_origin >= 5 & dep_hour_origin <= 8 ~ "5_8",
      dep_hour_origin >= 9 & dep_hour_origin <= 11 ~ "9_11",
      dep_hour_origin >= 12 & dep_hour_origin <= 14 ~ "12_14",
      dep_hour_origin >= 15 & dep_hour_origin <= 17 ~ "15_17",
      dep_hour_origin >= 18 & dep_hour_origin <= 22 ~ "18_22",
      !is.na(dep_hour_origin) ~ as.character(as.integer(dep_hour_origin)),
      TRUE ~ "UNK"
    ),
    lk_stop_id = as.integer(stop_id)
  ) %>%
  apply_lookup_group_rules()

lookup_output <- lookup_base %>%
  group_by(route_id, direction, lk_schedule_season, lk_schedule_daytype, lk_dep_hour_group, lk_stop_id) %>%
  summarise(
    n_obs = n(),
    median_arr_offset_min = ifelse(sum(!is.na(arr_offset_min)) > 0, median(arr_offset_min, na.rm = TRUE), NA_real_),
    median_dep_offset_min = ifelse(sum(!is.na(dep_offset_min)) > 0, median(dep_offset_min, na.rm = TRUE), NA_real_),
    .groups = "drop"
  ) %>%
  arrange(route_id, direction, lk_schedule_daytype, lk_schedule_season, lk_dep_hour_group, lk_stop_id)

expected_tbl <- lookup_base %>%
  left_join(
    lookup_output,
    by = c("route_id", "direction", "lk_schedule_season", "lk_schedule_daytype", "lk_dep_hour_group", "lk_stop_id")
  ) %>%
  mutate(
    lookup_level_used = "merged_rule_v1",
    lookup_n_obs = n_obs
  )

expected_tbl <- expected_tbl %>%
  mutate(
    expected_arrival_time = scheduled_time_original + as.difftime(median_arr_offset_min, units = "mins"),
    expected_departure_time = scheduled_time_original + as.difftime(median_dep_offset_min, units = "mins"),
    lateness_vs_original_schedule_min = as.numeric(difftime(actual_arrival_time, scheduled_time_original, units = "mins")),
    arrival_delay_min = as.numeric(difftime(actual_arrival_time, expected_arrival_time, units = "mins")),
    departure_delay_min = as.numeric(difftime(actual_departure_time, expected_departure_time, units = "mins"))
  )

lookup_output <- lookup_output %>%
  mutate(lookup_level_used = "merged_rule_v1", lookup_n_obs = n_obs) %>%
  select(lookup_level_used, route_id, direction, lk_schedule_season, lk_schedule_daytype, lk_dep_hour_group, lk_stop_id,
         lookup_n_obs, median_arr_offset_min, median_dep_offset_min)
```

## Trip Sequence and Delay Decomposition

```{r}
trip_summary <- expected_tbl %>%
  group_by(trip_key) %>%
  summarise(
    service_date = safe_mode(trip_date),
    route_id = safe_mode(route_id),
    direction = safe_mode(direction),
    vessel_id = safe_mode(vessel_id),
    sched_start = min(scheduled_time_original, na.rm = TRUE),
    sched_end = max(scheduled_time_original, na.rm = TRUE),
    terminal_arrival_delay_min = arrival_delay_min[which.max(stop_sequence)][1],
    .groups = "drop"
  ) %>%
  arrange(vessel_id, sched_start, trip_key) %>%
  group_by(vessel_id) %>%
  mutate(
    previous_trip_key = lag(trip_key),
    previous_sched_end = lag(sched_end),
    previous_trip_terminal_arrival_delay = lag(terminal_arrival_delay_min),
    gap_from_previous_sched_end_min = as.numeric(difftime(sched_start, previous_sched_end, units = "mins")),
    eligible_inherited_pair = !is.na(previous_trip_key) & !is.na(vessel_id) & !is.na(gap_from_previous_sched_end_min) &
      gap_from_previous_sched_end_min >= 0 & gap_from_previous_sched_end_min <= 30,
    inherited_prev_trip_delay = if_else(eligible_inherited_pair, pmax(coalesce(previous_trip_terminal_arrival_delay, 0), 0), 0)
  ) %>%
  mutate(
    vessel_trip_seq_in_chain = {
      x <- rep(1, n())
      if (n() > 1) for (j in 2:n()) x[j] <- ifelse(eligible_inherited_pair[j], x[j - 1] + 1, 1)
      as.numeric(x)
    },
    # Chain starts should have no prior turnaround gap (NA), then downstream cap/fill applies.
    planned_turnaround_gap_min = if_else(vessel_trip_seq_in_chain == 1, NA_real_, gap_from_previous_sched_end_min)
  ) %>%
  ungroup()

decomp_tbl <- expected_tbl %>%
  left_join(trip_summary %>% select(trip_key, inherited_prev_trip_delay, vessel_trip_seq_in_chain, planned_turnaround_gap_min), by = "trip_key") %>%
  arrange(trip_key, stop_sequence) %>%
  group_by(trip_key) %>%
  mutate(
    prev_stop_id = lag(stop_id),
    prev_stop_name = lag(stop_name),
    prev_stop_sequence = lag(stop_sequence),
    prev_stop_scheduled_time_original = lag(scheduled_time_original),
    prev_stop_sched_hour = as.numeric(hour(prev_stop_scheduled_time_original)),
    prev_stop_sched_hour_group = hour_group_5bin(prev_stop_sched_hour),
    from_stop_id = prev_stop_id,
    to_stop_id = stop_id,
    segment_id = segment_id_3digit(from_stop_id, to_stop_id),
    is_terminal_stop = stop_sequence == max(stop_sequence, na.rm = TRUE),
    prev_actual_arrival_time = lag(actual_arrival_time),
    prev_actual_departure_time = lag(actual_departure_time),
    prev_expected_arrival_time = lag(expected_arrival_time),
    prev_expected_departure_time = lag(expected_departure_time),
    prev_arrival_delay_min = lag(arrival_delay_min),
    origin_departure_delay_total = departure_delay_min[which.min(stop_sequence)][1],
    upstream_wait = pmax(0, as.numeric(difftime(prev_expected_arrival_time, prev_actual_arrival_time, units = "mins"))),
    upstream_landside_delay = as.numeric(difftime(prev_actual_departure_time, pmax(prev_actual_arrival_time, prev_expected_arrival_time), units = "mins")) -
      as.numeric(difftime(prev_expected_departure_time, prev_expected_arrival_time, units = "mins")),
    upstream_waterside_delay = as.numeric(difftime(actual_arrival_time, prev_actual_departure_time, units = "mins")) -
      as.numeric(difftime(expected_arrival_time, prev_expected_departure_time, units = "mins")),
    carried_within_trip_delay = case_when(
      stop_rank == 1 ~ 0,
      stop_rank >= 2 ~ prev_arrival_delay_min - inherited_prev_trip_delay,
      TRUE ~ NA_real_
    ),
    # Explicit business rule: second stop upstream_wait is always zero.
    upstream_wait = if_else(stop_rank == 1, 0, upstream_wait),
    # Unified upstream landside definition: second stop absorbs origin excess delay
    # (origin departure delay minus inherited previous-trip delay).
    upstream_landside_delay = if_else(stop_rank == 1, origin_departure_delay_total - inherited_prev_trip_delay, upstream_landside_delay),
    expected_water_min = as.numeric(difftime(expected_arrival_time, prev_expected_departure_time, units = "mins"))
  ) %>%
  ungroup()
```

## Completeness Flags, Environment/AIS Merge, and Final Tables

```{r}
decomp_tbl <- decomp_tbl %>%
  group_by(trip_key) %>%
  arrange(stop_sequence, .by_group = TRUE) %>%
  mutate(
    origin_stop_sequence = min(stop_sequence, na.rm = TRUE),
    terminal_stop_sequence = max(stop_sequence, na.rm = TRUE),
    bad_arrival = is.na(actual_arrival_time) & stop_sequence != origin_stop_sequence,
    bad_departure = is.na(actual_departure_time) & stop_sequence != terminal_stop_sequence,
    bad_required_time = bad_arrival | bad_departure,
    first_bad_stop_sequence = {
      idx_bad <- which(bad_required_time)
      if (length(idx_bad) > 0) min(stop_sequence[idx_bad], na.rm = TRUE) else NA_real_
    },
    trip_incomplete = any(bad_required_time),
    model_ineligible_from_incomplete = if_else(trip_incomplete & !is.na(first_bad_stop_sequence), stop_sequence >= first_bad_stop_sequence, FALSE)
  ) %>%
  ungroup() %>%
  mutate(
    boarding_wake = coalesce(boarding_wake, 0),
    alightings_wake = coalesce(alightings_wake, 0)
  ) %>%
  group_by(trip_key) %>%
  arrange(stop_sequence, .by_group = TRUE) %>%
  mutate(
    prev_stop_boardings = lag(boarding_wake),
    prev_stop_alightings = lag(alightings_wake),
    prev_stop_onboard = lag(onboard_raw)
  ) %>%
  ungroup() %>%
  mutate(
    prev_stop_boardings = coalesce(prev_stop_boardings, 0),
    prev_stop_alightings = coalesce(prev_stop_alightings, 0),
    prev_stop_onboard = pmax(coalesce(prev_stop_onboard, 0), 0),
    vessel_capacity = suppressWarnings(as.numeric(vessel_capacity)),
    prev_stop_load_factor = if_else(!is.na(vessel_capacity) & vessel_capacity > 0, prev_stop_onboard / vessel_capacity, NA_real_)
  )

shared_map <- decomp_tbl %>%
  filter(!is.na(trip_date), !is.na(stop_id), !is.na(route_id)) %>%
  group_by(trip_date, stop_id) %>%
  summarise(shared_flag = n_distinct(route_id) >= 2, .groups = "drop")

decomp_tbl <- decomp_tbl %>%
  left_join(shared_map %>% rename(from_stop_id = stop_id, prev_stop_shared_stop_flag = shared_flag), by = c("trip_date", "from_stop_id")) %>%
  left_join(shared_map %>% rename(to_stop_id = stop_id, curr_stop_shared_stop_flag = shared_flag), by = c("trip_date", "to_stop_id"))

decomp_tbl <- decomp_tbl %>%
  arrange(trip_date, route_id, direction, from_stop_id, prev_stop_scheduled_time_original, trip_key) %>%
  group_by(trip_date, route_id, direction, from_stop_id) %>%
  mutate(scheduled_headway_min = as.numeric(difftime(prev_stop_scheduled_time_original, lag(prev_stop_scheduled_time_original), units = "mins"))) %>%
  ungroup() %>%
  mutate(
    planned_turnaround_gap_min = if_else(is.na(planned_turnaround_gap_min) | planned_turnaround_gap_min > 30, 31, planned_turnaround_gap_min),
    scheduled_headway_min = if_else(is.na(scheduled_headway_min) | scheduled_headway_min > 90, 91, scheduled_headway_min)
  )

env_hourly_utc <- build_hourly_env_utc(path_env)

decomp_tbl <- decomp_tbl %>%
  mutate(
    prev_stop_sched_time_utc = with_tz(prev_stop_scheduled_time_original, tzone = TZ_UTC),
    prev_stop_sched_hour_start_utc = floor_date(prev_stop_sched_time_utc, unit = "hour")
  )

decomp_tbl <- decomp_tbl %>%
  left_join(env_hourly_utc, by = c("prev_stop_sched_hour_start_utc" = "env_hour_start_utc"))

ais_hourly_utc <- build_hourly_ais_utc(path_ais)

decomp_tbl <- decomp_tbl %>%
  left_join(
    ais_hourly_utc %>% rename(from_stop_id = stop_id, prev_stop_hourly_vessel_count_all = vessel_count),
    by = c("from_stop_id", "prev_stop_sched_hour_start_utc" = "ais_hour_start_utc")
  ) %>%
  left_join(
    ais_hourly_utc %>% rename(to_stop_id = stop_id, curr_stop_hourly_vessel_count_all = vessel_count),
    by = c("to_stop_id", "prev_stop_sched_hour_start_utc" = "ais_hour_start_utc")
  )

# Convert display hour fields back to New York for final outputs.
decomp_tbl <- decomp_tbl %>%
  mutate(
    prev_stop_sched_time_ny_display = with_tz(prev_stop_scheduled_time_original, tzone = TZ_NY),
    prev_stop_sched_hour = as.numeric(hour(prev_stop_sched_time_ny_display)),
    prev_stop_sched_hour_group = hour_group_5bin(prev_stop_sched_hour)
  ) %>%
  select(-prev_stop_sched_time_ny_display, -prev_stop_sched_time_utc, -prev_stop_sched_hour_start_utc)

decomp_tbl <- decomp_tbl %>%
  mutate(
    row_id = paste0(as.character(trip_key), as.character(as.integer(stop_sequence))),
    stop_sequence = as.numeric(stop_sequence),
    prev_stop_sequence = as.numeric(prev_stop_sequence),
    vessel_trip_seq_in_chain = as.numeric(vessel_trip_seq_in_chain)
  ) %>%
  relocate(row_id, .before = 1)

output_tbl <- decomp_tbl %>%
  filter(stop_sequence > 0)

baseline_final <- output_tbl %>%
  select(
    row_id, trip_key, route_id, direction,
    stop_sequence, prev_stop_sched_hour, prev_stop_sched_hour_group,
    upstream_landside_delay, upstream_waterside_delay, lateness_vs_original_schedule_min,
    inherited_prev_trip_delay, carried_within_trip_delay,
    schedule_season, schedule_daytype,
    vessel_trip_seq_in_chain, vessel_capacity,
    from_stop_id, to_stop_id, segment_id,
    is_terminal_stop, prev_stop_shared_stop_flag, curr_stop_shared_stop_flag,
    planned_turnaround_gap_min, scheduled_headway_min, expected_water_min,
    prev_stop_sequence, prev_stop_boardings, prev_stop_alightings, prev_stop_onboard,
    prev_stop_load_factor,
    env_temperature_c, env_wind_speed_mps, env_visibility_m, env_precipitation_mm,
    env_battery_water_level_anomaly, env_narrows_current_residual, env_battery_water_level_obs, env_narrows_current_obs,
    prev_stop_hourly_vessel_count_all, curr_stop_hourly_vessel_count_all,
    trip_incomplete, model_ineligible_from_incomplete
  ) %>%
  arrange(trip_key, stop_sequence, row_id)

decomp_final <- output_tbl %>%
  mutate(
    scheduled_time_original = with_tz(scheduled_time_original, tzone = TZ_NY),
    prev_stop_scheduled_time_original = with_tz(prev_stop_scheduled_time_original, tzone = TZ_NY),
    actual_arrival_time = with_tz(actual_arrival_time, tzone = TZ_NY),
    actual_departure_time = with_tz(actual_departure_time, tzone = TZ_NY),
    expected_arrival_time = with_tz(expected_arrival_time, tzone = TZ_NY),
    expected_departure_time = with_tz(expected_departure_time, tzone = TZ_NY)
  ) %>%
  select(
    row_id, trip_key, service_date = trip_date, route_id, direction,
    stop_sequence, stop_id, stop_name, prev_stop_id, prev_stop_name,
    schedule_season, schedule_daytype, prev_stop_sched_hour,
    scheduled_time_original, prev_stop_scheduled_time_original,
    actual_arrival_time, actual_departure_time, expected_arrival_time, expected_departure_time,
    lateness_vs_original_schedule_min, arrival_delay_min, departure_delay_min,
    inherited_prev_trip_delay, carried_within_trip_delay,
    upstream_wait, upstream_landside_delay, upstream_waterside_delay
  ) %>%
  arrange(trip_key, stop_sequence, row_id)
```

## Write Outputs

```{r}
write_csv(lookup_output, out_lookup, na = "")
write_csv(baseline_final, out_baseline, na = "")
write_csv(decomp_final, out_decomp, na = "")

write_parquet(lookup_output, out_lookup_parquet)
write_parquet(baseline_final, out_baseline_parquet)
write_parquet(decomp_final, out_decomp_parquet)

sanity_first_row <- baseline_final %>%
  slice(1) %>%
  select(
    row_id, trip_key, prev_stop_sched_hour,
    prev_stop_hourly_vessel_count_all, curr_stop_hourly_vessel_count_all,
    env_narrows_current_obs, env_temperature_c
  )
print(sanity_first_row)

cat("Core internal time arithmetic and joins run in UTC.\n")
cat("Output directory:", out_dir, "\n")
cat("Written files:\n")
cat("-", out_lookup, "\n")
cat("-", out_lookup_parquet, "\n")
cat("-", out_baseline, "\n")
cat("-", out_baseline_parquet, "\n")
cat("-", out_decomp, "\n")
cat("-", out_decomp_parquet, "\n")
```
