---
title: "Associative Model: Grouped Explanatory Framework"
output: html_document
author: "Project Team"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
options(scipen = 999, dplyr.summarise.inform = FALSE)
```

```{r packages, include=FALSE}
required_pkgs <- c("dplyr", "tidyr", "arrow", "broom", "knitr", "readr",
  "car", "ggplot2", "stringr", "forcats")
missing_pkgs <- required_pkgs[!vapply(required_pkgs, requireNamespace, logical(1), quietly = TRUE)]
if (length(missing_pkgs) > 0) stop("Missing packages: ", paste(missing_pkgs, collapse = ", "))

library(dplyr)
library(tidyr)
library(arrow)
library(broom)
library(knitr)
library(readr)
library(car)
library(ggplot2)
library(stringr)
library(broom)
library(forcats)
```

## 1) Associative Model Objective and Design

This section presents an associative model for explaining stop-level arrival lateness in the NYC Ferry system. The goal is interpretation rather than advance prediction. Instead of asking only whether a future trip will be delayed, this model asks which historical factors are associated with observed delay patterns.

The unit of analysis is a stop-level observation. The dependent variable is arrival lateness in minutes relative to the original schedule. Predictors are organized into five explanatory domains: structural context, environmental conditions, temporal patterns, operational propagation, and vessel characteristics.

This grouped design allows us to compare how different types of factors contribute to delay variation while keeping the final model interpretable for operational use.

```{r settings, include=FALSE}
set.seed(7900)

input_parquet <- "E:/MUSA/8010practicum/outputs/direct_pipeline_final_20260327/baseline_modeling_table_final_direct.parquet"
output_dir <- "E:/MUSA/8010practicum/outputs/direct_pipeline_final_20260327/associative_baseline_models_20260404"
dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)

outcome_cont <- "lateness_vs_original_schedule_min"


# Weekly local-time-slot imputation:
# interpreted as a 7-day centered window (±3 days) by local trip date.
week_window_days <- 3

candidate_predictors <- c(
  "route_id", "direction", "stop_sequence", "prev_stop_sched_hour", "prev_stop_sched_hour_group",
  "schedule_season", "schedule_daytype", "vessel_trip_seq_in_chain", "vessel_capacity",
  "is_terminal_stop", "prev_stop_shared_stop_flag", "curr_stop_shared_stop_flag",
  "planned_turnaround_gap_min", "scheduled_headway_min", "expected_water_min",
  "prev_stop_sequence", "prev_stop_boardings", "prev_stop_alightings", "prev_stop_onboard",
  "prev_stop_load_factor", "env_temperature_c", "env_wind_speed_mps", "env_visibility_m",
  "env_precipitation_mm", "env_battery_water_level_anomaly", "env_narrows_current_residual",
  "env_battery_water_level_obs", "env_narrows_current_obs"
)
prior_delay_predictors <- c("inherited_prev_trip_delay", "carried_within_trip_delay")
```

## 2) Data Scope

The model uses the stop-level baseline modeling table as its input dataset. After loading the data, we confirmed the number of records and the availability of candidate predictors before fitting the associative models.

```{r load, include=FALSE}
if (!file.exists(input_parquet)) stop("Missing input parquet: ", input_parquet)
df_raw <- read_parquet(input_parquet) %>% as_tibble()

df <- df_raw %>%
  mutate(
    trip_date = as.Date(substr(as.character(trip_key), 1, 8), format = "%Y%m%d")
  )

predictors_no_prior <- intersect(candidate_predictors, names(df))
predictors_with_prior <- unique(c(predictors_no_prior, intersect(prior_delay_predictors, names(df))))

tibble(
  input_rows = nrow(df_raw),
  rows_after_load = nrow(df),
  predictors_no_prior_n = length(predictors_no_prior),
  predictors_with_prior_n = length(predictors_with_prior)
) %>% kable(caption = "Data Scope and Predictor Availability")
```


## 3) Predictor Domains

To support interpretation, predictors in the associative model were organized into five explanatory domains. This grouped structure separates baseline route and stop context from environmental conditions, temporal patterns, operational propagation, and vessel-related characteristics. The goal is to interpret delay association at the domain level rather than focusing only on isolated individual coefficients.

### 3.1 Domain A: Structural Context

The structural domain captures baseline differences in how delay varies across the ferry network. These predictors describe route, stop position, and stop-context features that reflect how service is organized spatially and operationally. Variables in this domain include route and direction identifiers, stop sequence, terminal status, and shared-stop context. We interpret this domain as representing the structural baseline of delay rather than short-term changing conditions.

### 3.2 Domain B: Environmental Conditions

The environmental domain captures external natural conditions that may affect vessel movement and docking performance. These predictors include weather and water-related variables such as temperature, wind speed, visibility, precipitation, water-level, and current-related measures. This domain is especially important for the client because it helps distinguish external operating conditions from delay that is simply inherited from earlier parts of the trip.

### 3.3 Domain C: Temporal Patterns

The temporal domain captures recurring calendar and time-of-day structure in delay. These predictors describe when service occurs rather than where it occurs or what the vessel has already experienced. Variables in this domain include scheduled hour, scheduled hour group, day type, and season. We use this domain to represent regular service rhythms such as weekday versus weekend patterns, seasonal demand differences, and time-of-day effects.

### 3.4 Domain D: Operational Propagation

The operational propagation domain captures delay carried forward through the operating chain. These predictors reflect whether lateness has already accumulated earlier in the trip or in a prior linked trip on the same vessel. Variables in this domain include inherited previous-trip delay and carried within-trip delay. This domain is operationally meaningful, but it is interpreted as one component of the associative model rather than the sole basis for explaining delay.

### 3.5 Domain E: Vessel Characteristics

The vessel domain captures systematic differences associated with vessel capacity and operating role. These predictors help assess whether some delay patterns are associated with the characteristics of the vessel being used rather than only with route structure or environmental conditions. Variables in this domain include vessel capacity and related vessel-level operational identifiers retained in the final specification.


## 4) Data Preparation

Before model estimation, the stop-level modeling table was prepared to support a consistent and interpretable associative model. This section summarizes the key preparation steps: variable construction, missing-value treatment for environmental fields, and helper functions used for model fitting and summary extraction.

### 4.1 Data Cleaning and Variable Construction

The input table was converted into a stop-level analytic dataset. We created a trip date field from the trip identifier, checked which candidate predictors were available, and retained variables needed for the outcome and domain-based model structure.

### 4.2 Environmental Variable Imputation

Missing environmental values were imputed using a local time-slot rolling rule. The primary fill used observations from the same scheduled hour within a centered seven-day window around the trip date. If no value was available, the method fell back to the same scheduled hour group within the same centered window.

This approach keeps environmental predictors in the model without relying only on complete-case filtering. It also keeps the imputed values tied to nearby dates and comparable operating periods.

```{r impute, include=FALSE}
build_week_fill_map <- function(day_slot_tbl, group_col, date_col = "trip_date", value_col = "slot_day_median", window_days = 3) {
  if (nrow(day_slot_tbl) == 0) return(tibble())
  split_tbl <- split(day_slot_tbl, day_slot_tbl[[group_col]], drop = TRUE)
  out <- lapply(split_tbl, function(gdf) {
    gdf <- gdf[order(gdf[[date_col]]), , drop = FALSE]
    d <- gdf[[date_col]]
    v <- gdf[[value_col]]
    week_fill <- vapply(seq_along(v), function(i) {
      idx <- abs(as.integer(d - d[i])) <= window_days
      val <- suppressWarnings(median(v[idx], na.rm = TRUE))
      ifelse(is.finite(val), val, NA_real_)
    }, numeric(1))
    gdf$week_fill <- week_fill
    gdf[, c(group_col, date_col, "week_fill"), drop = FALSE]
  })
  bind_rows(out)
}

impute_by_local_slot_week <- function(data, var_name, window_days = 3) {
  x <- data
  miss_before <- sum(is.na(x[[var_name]]))
  if (miss_before == 0) {
    return(list(data = x, summary = tibble(variable = var_name, n_missing_before = 0L, n_filled_primary = 0L, n_filled_fallback = 0L, n_missing_after = 0L)))
  }

  # Primary: same local hour within centered 7-day window (±3 days by default).
  primary_daily <- x %>%
    filter(!is.na(.data[[var_name]]), !is.na(trip_date), !is.na(prev_stop_sched_hour)) %>%
    group_by(prev_stop_sched_hour, trip_date) %>%
    summarise(slot_day_median = median(.data[[var_name]], na.rm = TRUE), .groups = "drop")
  primary_fill <- build_week_fill_map(primary_daily, group_col = "prev_stop_sched_hour", value_col = "slot_day_median", window_days = window_days) %>%
    rename(fill_primary = week_fill)

  # Fallback: same local hour-group within centered 7-day window.
  fallback_daily <- x %>%
    filter(!is.na(.data[[var_name]]), !is.na(trip_date), !is.na(prev_stop_sched_hour_group)) %>%
    group_by(prev_stop_sched_hour_group, trip_date) %>%
    summarise(slot_day_median = median(.data[[var_name]], na.rm = TRUE), .groups = "drop")
  fallback_fill <- build_week_fill_map(fallback_daily, group_col = "prev_stop_sched_hour_group", value_col = "slot_day_median", window_days = window_days) %>%
    rename(fill_fallback = week_fill)

  x <- x %>%
    left_join(primary_fill, by = c("prev_stop_sched_hour", "trip_date")) %>%
    left_join(fallback_fill, by = c("prev_stop_sched_hour_group", "trip_date"))

  was_missing <- is.na(x[[var_name]])
  fill_primary_idx <- was_missing & !is.na(x$fill_primary)
  fill_fallback_idx <- was_missing & is.na(x$fill_primary) & !is.na(x$fill_fallback)

  x[[var_name]][fill_primary_idx] <- x$fill_primary[fill_primary_idx]
  x[[var_name]][fill_fallback_idx] <- x$fill_fallback[fill_fallback_idx]
  x <- x %>% select(-fill_primary, -fill_fallback)

  list(
    data = x,
    summary = tibble(
      variable = var_name,
      n_missing_before = miss_before,
      n_filled_primary = sum(fill_primary_idx, na.rm = TRUE),
      n_filled_fallback = sum(fill_fallback_idx, na.rm = TRUE),
      n_missing_after = sum(is.na(x[[var_name]]))
    )
  )
}

impute_vars <- intersect(
  c(
    "env_temperature_c",
    "env_wind_speed_mps",
    "env_visibility_m",
    "env_precipitation_mm",
    "env_narrows_current_residual",
    "env_battery_water_level_obs",
    "env_narrows_current_obs"
  ),
  names(df)
)

impute_summary_list <- list()
df_imp <- df
for (v in impute_vars) {
  res <- impute_by_local_slot_week(df_imp, v, window_days = week_window_days)
  df_imp <- res$data
  impute_summary_list[[length(impute_summary_list) + 1]] <- res$summary
}
imputation_summary <- bind_rows(impute_summary_list)
```

```{r imputation-summary, echo=FALSE}
imputation_summary %>%
  mutate(
    pct_missing_before = round(n_missing_before / nrow(df) * 100, 2),
    pct_missing_after = round(n_missing_after / nrow(df_imp) * 100, 2)
  ) %>%
  kable(
    caption = "Summary of Environmental Variable Imputation",
    digits = 2
  )
```

### 4.3 Modeling Utilities

A small set of reusable helper functions was created to prepare model inputs, fit linear models, extract model metrics, and generate standardized coefficient summaries.
```{r helpers, include=FALSE}
prepare_model_df <- function(data, outcome, predictors) {
  data %>% select(all_of(c(outcome, predictors))) %>% drop_na()
}

fit_linear <- function(data, model_name, outcome, predictors) {
  d <- prepare_model_df(data, outcome, predictors)
  fml <- as.formula(paste(outcome, "~", paste(predictors, collapse = " + ")))
  fit <- lm(fml, data = d)
  pred <- predict(fit, newdata = d)
  rmse <- sqrt(mean((d[[outcome]] - pred)^2))
  g <- glance(fit)

  list(
    metrics = tibble(
      model_name = model_name,
      model_type = "linear",
      n = nrow(d),
      aic = g$AIC,
      bic = g$BIC,
      rsq = g$r.squared,
      adj_rsq = g$adj.r.squared,
      rmse = rmse,
      pseudo_rsq = NA_real_,
      accuracy = NA_real_,
      sensitivity = NA_real_,
      specificity = NA_real_,
      auc = NA_real_
    ),
    coefs = tidy(fit) %>% mutate(model_name = model_name, model_type = "linear"),
    formula_text = paste(deparse(fml), collapse = "")
  )
}

```

```{r standardized_helpers, include=FALSE}
standardize_for_lm <- function(data, outcome, predictors) {
  d <- data %>%
    select(all_of(c(outcome, predictors))) %>%
    drop_na()

  num_vars <- names(d)[sapply(d, is.numeric)]

  d_std <- d %>%
    mutate(across(all_of(num_vars), ~ as.numeric(scale(.x))))

  d_std
}

fit_standardized_linear <- function(data, model_name, outcome, predictors) {
  d_std <- standardize_for_lm(data, outcome, predictors)
  fml <- as.formula(paste(outcome, "~", paste(predictors, collapse = " + ")))
  fit <- lm(fml, data = d_std)

  broom::tidy(fit) %>%
    filter(term != "(Intercept)") %>%
    mutate(
      model_name = model_name,
      abs_estimate = abs(estimate)
    ) %>%
    arrange(desc(abs_estimate))
}
```


## 5) Variable Screening and Final Specification

This section explains how the final associative-model specification was selected from the broader predictor pool. The goal was not to include every available variable, but to keep interpretable coverage across the five explanatory domains while reducing redundancy among closely related predictors.

### 5.1 Initial Candidate Variables

The initial predictor pool was intentionally broad so that each explanatory domain was represented. Candidate variables included route and stop structure, environmental conditions, temporal indicators, operational propagation terms, and vessel/load-related characteristics. This broader pool provided the starting point for a smaller, more interpretable final specification.
```{r domain-predictors, include=FALSE}
domain_a_predictors <- intersect(c(
  "route_id",
  "direction",
  "stop_sequence",
  "is_terminal_stop",
  "prev_stop_shared_stop_flag",
  "curr_stop_shared_stop_flag"
), names(df_imp))

domain_b_predictors <- intersect(c(
  "env_temperature_c",
  "env_wind_speed_mps",
  "env_visibility_m",
  "env_precipitation_mm",
  "env_narrows_current_residual",
  "env_battery_water_level_obs",
  "env_narrows_current_obs"
), names(df_imp))

domain_c_predictors <- intersect(c(
  "prev_stop_sched_hour",
  "schedule_season",
  "schedule_daytype"
), names(df_imp))

domain_d_predictors <- intersect(c(
  "inherited_prev_trip_delay",
  "carried_within_trip_delay"
), names(df_imp))

domain_e_predictors <- intersect(c(
  "vessel_capacity",
  "vessel_trip_seq_in_chain",
  "prev_stop_boardings",
  "prev_stop_alightings",
  "prev_stop_onboard",
  "prev_stop_load_factor"
), names(df_imp))

final_predictors <- unique(c(
  domain_a_predictors,
  domain_b_predictors,
  domain_c_predictors,
  domain_d_predictors,
  domain_e_predictors
))
```
### 5.2 Redundancy and Multicollinearity Checks

Because several candidate predictors measured closely related aspects of stop position, ridership, and operating context, we reviewed pairwise correlations among numeric predictors before fitting the final associative model. The goal was not to eliminate all correlation, but to identify clusters of variables that conveyed largely overlapping information and would make coefficient interpretation less clear.


```{r corr_setup, include=FALSE}
numeric_candidate_predictors <- intersect(
  candidate_predictors,
  names(df_imp)[sapply(df_imp, is.numeric)]
)

corr_vars <- df_imp %>%
  select(all_of(numeric_candidate_predictors)) %>%
  select(where(~ !all(is.na(.))))

corr_matrix <- cor(corr_vars, use = "pairwise.complete.obs")
```
The figure below highlights selected variables reviewed during specification design. These variables are related to stop position and previous-stop load conditions, so they were checked carefully before finalizing the model.
```{r corr_plot_focus, fig.width=6, fig.height=5}
focus_vars <- c(
  "stop_sequence",
  "prev_stop_sequence",
  "prev_stop_onboard",
  "prev_stop_load_factor"
)


focus_corr <- corr_matrix[focus_vars, focus_vars]

focus_plot_df <- as.data.frame(as.table(focus_corr)) %>%
  rename(var1 = Var1, var2 = Var2, corr = Freq) %>%
  mutate(label = ifelse(var1 == var2, "", sprintf("%.2f", corr)))

focus_plot <- ggplot(focus_plot_df, aes(x = var1, y = var2, fill = corr)) +
  geom_tile(color = "white", linewidth = 0.5) +
  geom_text(aes(label = label), size = 4) +
  scale_fill_gradient2(
    low = "#1A9988", mid = "white", high = "#EB5600",
    midpoint = 0,
    limits = c(-1, 1)
  ) +
  coord_fixed() +
  theme_minimal() +
  theme(
    panel.grid = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(size = 13, face = "bold", hjust = 0.5)
  ) +
  labs(
    title = "Correlation Among Selected Candidate Predictors",
    x = NULL,
    y = NULL,
    fill = "Correlation"
  )

focus_plot
```
To support specification decisions more systematically, we also identified highly correlated numeric predictor pairs using an absolute correlation threshold of 0.7.
```{r corr_pairs}
corr_pairs <- as.data.frame(as.table(corr_matrix)) %>%
  rename(var1 = Var1, var2 = Var2, corr = Freq) %>%
  filter(var1 != var2) %>%
  mutate(abs_corr = abs(corr)) %>%
  filter(abs_corr >= 0.7) %>%
  rowwise() %>%
  mutate(pair_id = paste(sort(c(var1, var2)), collapse = " || ")) %>%
  ungroup() %>%
  distinct(pair_id, .keep_all = TRUE) %>%
  arrange(desc(abs_corr)) %>%
  select(var1, var2, corr, abs_corr)

corr_pairs %>%
  head(10) %>%
  kable(caption = "Highly correlated candidate predictor pairs (|r| >= 0.7)")
```
### 5.3 Final Variable Set

The final variable set was selected to preserve coverage across the five explanatory domains while avoiding unnecessary duplication. When multiple variables represented similar information, we prioritized predictors that were easier to interpret operationally and more directly connected to the domain being represented.

After reviewing correlation patterns, we removed redundant candidate variables and retained a reduced final specification for the combined associative model.
```{r vif_setup, include=FALSE}
final_predictors_trimmed <- setdiff(
  final_predictors,
  c("prev_stop_sequence", "prev_stop_sched_hour_group")
)

tibble(
  model = "Final combined associative model",
  n_predictors = length(final_predictors_trimmed)
) %>%
  kable(caption = "Predictor count in the final combined associative specification")
```
```{r vif_run, include=FALSE}
fit_vif_model <- function(data, outcome, predictors) {
  vars_needed <- unique(c(outcome, predictors))
  model_df <- data %>%
    select(all_of(vars_needed)) %>%
    drop_na()

  formula_str <- paste(outcome, "~", paste(predictors, collapse = " + "))
  lm(as.formula(formula_str), data = model_df)
}

extract_vif_table <- function(model, model_name) {
  vif_vals <- car::vif(model)

  if (is.matrix(vif_vals)) {
    out <- as.data.frame(vif_vals)
    out$term <- rownames(out)
    rownames(out) <- NULL
    out$model <- model_name
    out %>%
      rename(gvif = GVIF) %>%
      mutate(adjusted_vif = gvif^(1 / (2 * Df))) %>%
      select(model, term, Df, gvif, adjusted_vif)
  } else {
    tibble(
      model = model_name,
      term = names(vif_vals),
      vif = as.numeric(vif_vals)
    ) %>%
      arrange(desc(vif))
  }
}

vif_model_final <- fit_vif_model(df_imp, outcome_cont, final_predictors_trimmed)
vif_table_final <- extract_vif_table(vif_model_final, "Final combined associative model")
```
We then checked variance inflation factors for the final combined specification. For multi-level categorical predictors, we report the adjusted GVIF, which is more comparable across terms with different degrees of freedom.
```{r vif_table}
if ("adjusted_vif" %in% names(vif_table_final)) {
  vif_table_final %>%
    arrange(desc(adjusted_vif)) %>%
    head(10) %>%
    mutate(
      gvif = round(gvif, 2),
      adjusted_vif = round(adjusted_vif, 2)
    ) %>%
    kable(caption = "Top VIF results for the final combined associative specification")
} else {
  vif_table_final %>%
    arrange(desc(vif)) %>%
    head(10) %>%
    mutate(vif = round(vif, 2)) %>%
    kable(caption = "Top VIF results for the final combined associative specification")
}
```


## 6) Estimation Strategy
The associative model was estimated in two parts. First, we fit domain-specific linear models to evaluate how much stop-level delay variation was associated with each explanatory domain on its own. Second, we fit a combined associative model that included predictors from all five domains simultaneously. This strategy allows the analysis to distinguish domain-level explanatory contribution from the behavior of the full final specification.

### 6.1 Domain-Specific Models
To assess the explanatory role of each factor group, we estimated separate linear models for the five domains defined earlier: structural context, environmental conditions, temporal patterns, operational propagation, and vessel characteristics. Each model used the same dependent variable—stop-level arrival lateness in minutes—but included only the predictors assigned to that domain.

These domain-specific models were not intended as competing final models. Instead, they were used to show how much variation in delay could be associated with each domain when considered on its own.

```{r fit}
df_model <- df_imp

model_results <- list(
  domain_a_structural = fit_linear(df_model, "domain_a_structural", outcome_cont, domain_a_predictors),
  domain_b_environment = fit_linear(df_model, "domain_b_environment", outcome_cont, domain_b_predictors),
  domain_c_temporal = fit_linear(df_model, "domain_c_temporal", outcome_cont, domain_c_predictors),
  domain_d_operational = fit_linear(df_model, "domain_d_operational", outcome_cont, domain_d_predictors),
  domain_e_vessel = fit_linear(df_model, "domain_e_vessel", outcome_cont, domain_e_predictors),
  combined_associative = fit_linear(df_model, "combined_associative", outcome_cont, final_predictors_trimmed)
)
```


### 6.2 Combined Associative Model


After estimating the domain-specific models, we fit a combined associative model using the final predictor set across all five domains. This combined model serves as the final associative specification in the markdown. Its purpose is to provide a single interpretable model while preserving the ability to discuss structural, environmental, temporal, operational, and vessel-related contributions within one framework.
### 6.3 Comparison Logic

Model comparison focused on interpretability rather than model selection in a predictive sense. For each domain-specific model, we report summary fit statistics to show the amount of delay variation associated with that domain alone. We then compare those results with the combined associative model to show how the full specification performs when all domains are considered together.

This comparison logic supports the main purpose of the associative model: distinguishing baseline structural effects, external environmental conditions, temporal rhythms, operational delay propagation, and vessel-related differences within a single explanatory framework.


## 7) Results: Explanatory Contribution by Domain
The domain-specific models show that stop-level delay variation is not evenly distributed across explanatory domains. Operational propagation accounts for the largest share of explained variance, followed by structural context. Environmental conditions, temporal patterns, and vessel characteristics each explain a smaller portion of variation on their own, but together they add important interpretive context beyond propagated delay alone. The combined associative model achieves the strongest overall fit, indicating that these domains are complementary rather than redundant.
```{r}
model_comparison_summary <- bind_rows(lapply(model_results, `[[`, "metrics")) %>%
  mutate(
  domain = dplyr::recode(
    model_name,
    combined_associative = "Combined Associative Model",
    domain_a_structural = "Domain A: Structural Context",
    domain_b_environment = "Domain B: Environmental Conditions",
    domain_c_temporal = "Domain C: Temporal Patterns",
    domain_d_operational = "Domain D: Operational Propagation",
    domain_e_vessel = "Domain E: Vessel Characteristics"
  )
) %>%
  arrange(match(domain, c(
    "Domain A: Structural Context",
    "Domain B: Environmental Conditions",
    "Domain C: Temporal Patterns",
    "Domain D: Operational Propagation",
    "Domain E: Vessel Characteristics",
    "Combined Associative Model"
  )))

model_comparison_summary %>%
  select(domain, n, rsq, adj_rsq, rmse, aic, bic) %>%
  mutate(
    rsq = round(rsq, 4),
    adj_rsq = round(adj_rsq, 4),
    rmse = round(rmse, 4),
    aic = round(aic, 0),
    bic = round(bic, 0)
  ) %>%
  kable(caption = "Explained Variation by Domain and Final Combined Associative Model")
```

### 7.1 Structural Context

The structural-context model explains a meaningful share of stop-level delay variation on its own (R² = 0.1329). This indicates that route and stop configuration establish an important baseline pattern of delay across the system. In substantive terms, some portions of the ferry network appear structurally more delay-prone than others even before environmental or operational carry-forward effects are considered.


### 7.2 Environmental Conditions
```{r}
env_feature_table <- fit_standardized_linear(
  df_model,
  "domain_b_environment",
  outcome_cont,
  domain_b_predictors
)

env_feature_table %>%
  select(term, estimate, std.error, statistic, p.value, abs_estimate) %>%
  mutate(
    estimate = round(estimate, 4),
    std.error = round(std.error, 4),
    statistic = round(statistic, 2),
    p.value = round(p.value, 4),
    abs_estimate = round(abs_estimate, 4)
  ) %>%
  kable(caption = "Standardized Coefficients for Environmental Predictors")
```
```{r}
structural_feature_table <- fit_standardized_linear(
  df_model,
  "domain_a_structural",
  outcome_cont,
  domain_a_predictors
)

structural_feature_table %>%
  select(term, estimate, std.error, statistic, p.value, abs_estimate) %>%
  mutate(
    estimate = round(estimate, 4),
    std.error = round(std.error, 4),
    statistic = round(statistic, 2),
    p.value = round(p.value, 4),
    abs_estimate = round(abs_estimate, 4)
  ) %>%
  kable(caption = "Standardized Coefficients for Structural Predictors")
```

```{r}
vessel_feature_table <- fit_standardized_linear(
  df_model,
  "domain_e_vessel_demand",
  outcome_cont,
  domain_e_predictors
)

vessel_feature_table %>%
  select(term, estimate, std.error, statistic, p.value, abs_estimate) %>%
  mutate(
    estimate = round(estimate, 4),
    std.error = round(std.error, 4),
    statistic = round(statistic, 2),
    p.value = round(p.value, 4),
    abs_estimate = round(abs_estimate, 4)
  ) %>%
  kable(caption = "Standardized Coefficients for Demand and Vessel Predictors")
```

```{r}
operational_feature_table <- fit_standardized_linear(
  df_model,
  "domain_d_operational",
  outcome_cont,
  domain_d_predictors
)

operational_feature_table %>%
  select(term, estimate, std.error, statistic, p.value, abs_estimate) %>%
  mutate(
    estimate = round(estimate, 4),
    std.error = round(std.error, 4),
    statistic = round(statistic, 2),
    p.value = round(p.value, 4),
    abs_estimate = round(abs_estimate, 4)
  ) %>%
  kable(caption = "Standardized Coefficients for Operational Propagation Predictors")
```


```{r}
temporal_feature_table <- fit_standardized_linear(
  df_model,
  "domain_c_temporal",
  outcome_cont,
  domain_c_predictors
)

temporal_feature_table %>%
  select(term, estimate, std.error, statistic, p.value, abs_estimate) %>%
  mutate(
    estimate = round(estimate, 4),
    std.error = round(std.error, 4),
    statistic = round(statistic, 2),
    p.value = round(p.value, 4),
    abs_estimate = round(abs_estimate, 4)
  ) %>%
  kable(caption = "Standardized Coefficients for Temporal Predictors")
```

```{r}

selected_predictors <- intersect(c(

  "inherited_prev_trip_delay",
  "route_id",

  "schedule_season",
  "prev_stop_sched_hour_group",
  "schedule_daytype",
  "env_temperature_c",

  "env_wind_speed_mps",
  "env_precipitation_mm",
  "env_battery_water_level_obs",

  "stop_sequence"
), names(df_model))

selected_model_df <- prepare_model_df(
  df_model,
  outcome_cont,
  selected_predictors
)

selected_formula <- as.formula(
  paste(
    outcome_cont,
    "~",
    paste(selected_predictors, collapse = " + "),
    "+ prev_stop_sched_hour_group:schedule_daytype"
  )
)

selected_fit <- lm(selected_formula, data = selected_model_df)

selected_metrics <- broom::glance(selected_fit) %>%
  transmute(
    model = "Selected compact associative model",
    n = nobs(selected_fit),
    rsq = r.squared,
    adj_rsq = adj.r.squared,
    rmse = sqrt(mean(residuals(selected_fit)^2)),
    aic = AIC(selected_fit),
    bic = BIC(selected_fit)
  )

selected_metrics %>%
  mutate(
    rsq = round(rsq, 4),
    adj_rsq = round(adj_rsq, 4),
    rmse = round(rmse, 4),
    aic = round(aic, 0),
    bic = round(bic, 0)
  ) %>%
  kable(caption = "Selected Compact Associative Model Fit")
```

```{r selected_compact_model_standardized_coefficients}
# --- Standardized coefficient strength for selected compact model ---

x_mat <- model.matrix(selected_formula, data = selected_model_df)
x_mat <- x_mat[, colnames(x_mat) != "(Intercept)", drop = FALSE]

x_sd <- apply(x_mat, 2, sd, na.rm = TRUE)
x_mat <- x_mat[, is.finite(x_sd) & x_sd > 0, drop = FALSE]

x_scaled <- scale(x_mat)
y_scaled <- as.numeric(scale(selected_model_df[[outcome_cont]]))

std_df <- as.data.frame(x_scaled, check.names = FALSE)
std_df$.y_scaled <- y_scaled

selected_std_fit <- lm(.y_scaled ~ ., data = std_df)

selected_std_coef <- broom::tidy(selected_std_fit) %>%
  filter(term != "(Intercept)") %>%
  mutate(
    standardized_strength = abs(estimate),
    abs_statistic = abs(statistic)
  ) %>%
  arrange(desc(standardized_strength))

selected_std_coef %>%
  select(term, standardized_strength, std.error, abs_statistic, p.value) %>%
  mutate(
    standardized_strength = round(standardized_strength, 4),
    std.error = round(std.error, 4),
    abs_statistic = round(abs_statistic, 2),
    p.value = round(p.value, 4)
  ) %>%
  kable(caption = "Standardized Coefficient Strength for Selected Compact Model")
```

```{r}
route_summary <- df_model %>%
  group_by(route_id) %>%
  summarise(
    avg_delay = mean(.data[[outcome_cont]], na.rm = TRUE),
    avg_expected_water_min = mean(expected_water_min, na.rm = TRUE),
    n = n(),
    .groups = "drop"
  )

ggplot(route_summary, aes(x = avg_expected_water_min, y = avg_delay, label = route_id)) +
  geom_point(size = 3) +
  geom_text(nudge_y = 0.15) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    title = "Route Travel Time vs Average Lateness",
    x = "Average expected water travel time (minutes)\n(proxy for route length / exposure)",
    y = "Average lateness (minutes)"
  ) +
  theme_minimal()
```

```{r excessive_delay_components_plot, fig.width=8, fig.height=4}
delay_components <- tibble::tibble(
  component = c(
    "Inherited previous delay",
    "Carried within-trip delay",
    "New delay at current stop"
  ),
  minutes = c(6.24, 1.85, 1.68)
) %>%
  mutate(
    component = factor(
      component,
      levels = c(
        "Inherited previous delay",
        "Carried within-trip delay",
        "New delay at current stop"
      )
    )
  )

ggplot(delay_components, aes(x = "Excessive delay example", y = minutes, fill = component)) +
  geom_col(width = 0.45) +
  geom_text(
    aes(label = paste0(minutes, " min")),
    position = position_stack(vjust = 0.5),
    color = "white",
    fontface = "bold",
    size = 4
  ) +
  scale_fill_manual(
    values = c(
      "Inherited previous delay" = "#EB5600",
      "Carried within-trip delay" = "#1A9988",
      "New delay at current stop" = "#C7C7C7"
    )
  ) +
  labs(
    title = "Where Does Excessive Delay Come From?",
    subtitle = "Most excessive delay is inherited or carried forward, not newly created at the current stop",
    x = NULL,
    y = "Minutes of delay",
    fill = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    legend.position = "bottom",
    plot.title = element_text(face = "bold")
  )
```


```{r delay_accumulation_by_stop_sequence, fig.width=8, fig.height=5}
delay_accumulation <- df_model %>%
  group_by(route_id, stop_sequence) %>%
  summarise(
    avg_lateness = mean(.data[[outcome_cont]], na.rm = TRUE),
    n = n(),
    .groups = "drop"
  ) %>%
  filter(n >= 100)

ggplot(delay_accumulation, aes(x = stop_sequence, y = avg_lateness, color = route_id)) +
  geom_line(linewidth = 0.9) +
  geom_point(size = 1.8) +
  labs(
    title = "Delay Accumulation Along Route",
    subtitle = "Average lateness tends to change as trips move through stop sequence",
    x = "Stop sequence",
    y = "Average lateness (minutes)",
    color = "Route"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold"),
    legend.position = "bottom"
  )
```


```{r time_of_day_delay_plot, fig.width=8, fig.height=5}

possible_lateness_cols <- c(
  if (exists("outcome_cont")) outcome_cont else character(0),
  "arr_lateness",
  "arr_late_min",
  "arrival_lateness",
  "arrival_time_variance",
  "arrivaltimevariance",
  "ArrivalTimeVariance",
  "delay_min",
  "lateness"
)

lateness_col <- possible_lateness_cols[possible_lateness_cols %in% names(df_model)][1]

if (is.na(lateness_col)) {
  stop("Could not find the lateness column. Please run: names(df_model)[grepl('late|delay|variance|arrival', names(df_model), ignore.case = TRUE)]")
}

# 2. Create summary table
time_delay_summary <- df_model %>%
  mutate(lateness_for_plot = .data[[lateness_col]]) %>%
  filter(
    !is.na(prev_stop_sched_hour_group),
    !is.na(schedule_daytype),
    !is.na(lateness_for_plot)
  ) %>%
  group_by(prev_stop_sched_hour_group, schedule_daytype) %>%
  summarise(
    avg_lateness = mean(lateness_for_plot, na.rm = TRUE),
    n = n(),
    .groups = "drop"
  )

# 3. Plot
time_delay_summary <- df_model %>%
  mutate(lateness_for_plot = .data[[lateness_col]]) %>%
  filter(
    !is.na(prev_stop_sched_hour_group),
    !is.na(schedule_daytype),
    !is.na(lateness_for_plot)
  ) %>%
  group_by(prev_stop_sched_hour_group, schedule_daytype) %>%
  summarise(
    avg_lateness = mean(lateness_for_plot, na.rm = TRUE),
    n = n(),
    .groups = "drop"
  ) %>%
  mutate(
    prev_stop_sched_hour_group = factor(
      prev_stop_sched_hour_group,
      levels = c("5_8", "9_11", "12_14", "15_17", "18_22")
    )
  )

ggplot(
  time_delay_summary,
  aes(
    x = prev_stop_sched_hour_group,
    y = avg_lateness,
    fill = schedule_daytype
  )
) +
  geom_col(position = position_dodge(width = 0.55), width = 0.38) +
  scale_fill_manual(
    values = c(
      "WEEKDAY" = "#EB5600",
      "WEEKEND" = "#1A9988"
    )
  ) +
  labs(
    title = "Average Lateness by Time Bin and Day Type",
    x = "Scheduled time bin",
    y = "Average lateness (minutes)",
    fill = "Day type"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold"),
    legend.position = "bottom"
  )
```

```{r time_of_day_delay_plot, fig.width=8, fig.height=5}
ggplot(
  time_delay_summary,
  aes(
    x = prev_stop_sched_hour_group,
    y = avg_lateness,
    fill = schedule_daytype
  )
) +
  geom_col(position = position_dodge(width = 0.75), width = 0.6) +
  scale_fill_manual(
    values = c(
      "WEEKDAY" = "#EB5600",
      "WEEKEND" = "#1A9988"
    )
  ) +
  labs(
    title = "Average Lateness by Time Bin and Day Type",
    x = "Scheduled time bin",
    y = "Average lateness (minutes)",
    fill = "Day type"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold"),
    legend.position = "bottom"
  )
```

---


```{r temperature_lateness_plot, fig.width=8, fig.height=5}
ggplot(temp_delay_summary, aes(x = avg_temperature, y = avg_lateness)) +
  geom_point(color = "#1A9988", size = 2.5) +
  geom_smooth(method = "lm", se = FALSE, color = "#EB5600") +
  labs(
    title = "Temperature and Average Lateness",
    x = "Average temperature (°C)",
    y = "Average lateness (minutes)"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold")
  )
```

```{r}
# --- Top 10 coefficient bar chart for Domain B / C / E ---

# Add demand/load variables into Domain E for this chart
domain_e_plot_predictors <- intersect(c(
  domain_e_predictors,
  "prev_stop_boardings",
  "prev_stop_alightings",
  "prev_stop_onboard",
  "prev_stop_load_factor"
), names(df_model))

# Predictors to show in the chart: Domain B + C + updated E
bce_predictors <- unique(c(
  domain_b_predictors,
  domain_c_predictors,
  domain_e_plot_predictors
))

# Rebuild final predictors for this plot model
plot_model_predictors <- unique(c(
  domain_a_predictors,
  domain_b_predictors,
  domain_c_predictors,
  domain_d_predictors,
  domain_e_plot_predictors
))

# Standardize numeric predictors
numeric_predictors <- plot_model_predictors[
  sapply(df_model[plot_model_predictors], is.numeric)
]

df_model_std <- df_model %>%
  mutate(across(
    all_of(numeric_predictors),
    ~ as.numeric(scale(.x))
  ))

# Fit a direct lm model
plot_formula <- as.formula(
  paste(outcome_cont, "~", paste(plot_model_predictors, collapse = " + "))
)

combined_std_fit <- lm(plot_formula, data = df_model_std)

# Domain lookup
domain_lookup <- tibble(
  variable = bce_predictors,
  domain = case_when(
    variable %in% domain_b_predictors ~ "Environmental Conditions",
    variable %in% domain_c_predictors ~ "Temporal Patterns",
    variable %in% domain_e_plot_predictors ~ "Vessel / Demand-Load Context",
    TRUE ~ "Other"
  )
)

# Readable labels
label_lookup <- c(
  env_temperature_c = "Temperature",
  env_wind_speed_mps = "Wind speed",
  env_visibility_m = "Visibility",
  env_precipitation_mm = "Precipitation",
  env_battery_water_level_obs = "Water level",
  env_battery_water_level_anomaly = "Water-level anomaly",
  env_narrows_current_obs = "Current",
  env_narrows_current_residual = "Current residual",

  prev_stop_sched_hour = "Previous scheduled hour",
  prev_stop_sched_hour_group = "Previous scheduled hour group",
  schedule_season = "Schedule season",
  schedule_daytype = "Schedule day type",

  vessel_capacity = "Vessel capacity",
  vessel_trip_seq_in_chain = "Vessel trip sequence",
  prev_stop_boardings = "Boardings",
  prev_stop_alightings = "Alightings",
  prev_stop_onboard = "Onboard count",
  prev_stop_load_factor = "Load factor"
)

# Helper: match coefficient term back to base variable
get_base_var <- function(term, vars) {
  hits <- vars[vapply(vars, function(v) startsWith(term, v), logical(1))]
  if (length(hits) == 0) return(NA_character_)
  hits[which.max(nchar(hits))]
}

# Build plotting dataframe
# --- Build coefficient table for plotting ---

coef_plot_df <- broom::tidy(combined_std_fit) %>%
  filter(term != "(Intercept)") %>%
  rowwise() %>%
  mutate(variable = get_base_var(term, bce_predictors)) %>%
  ungroup() %>%
  filter(!is.na(variable)) %>%
  mutate(
    domain = case_when(
      variable %in% domain_b_predictors ~ "Environmental Conditions",
      variable %in% domain_c_predictors ~ "Temporal Patterns",
      variable %in% domain_e_plot_predictors ~ "Demand and Vessel Context",
      TRUE ~ "Other"
    )
  ) %>%
  filter(domain != "Other") %>%
  filter(
    !str_detect(term, "^schedule_season"),
    !variable %in% c(
      "prev_stop_load_factor",   # remove load factor
      "prev_stop_onboard",       # optional: remove onboard if you feel redundant
      "vessel_trip_seq_in_chain" # optional: keep chart cleaner
    )
  ) %>%
  mutate(
    clean_label = if_else(
      variable %in% names(label_lookup),
      unname(label_lookup[variable]),
      variable
    ),
    clean_label = case_when(
      variable == "schedule_daytype" ~ str_replace(term, variable, "Schedule day type "),
      variable == "prev_stop_sched_hour_group" ~ str_replace(term, variable, "Previous scheduled hour "),
      TRUE ~ clean_label
    ),
    abs_estimate = abs(estimate)
  )

# --- Select top terms by domain ---
coef_plot_top <- bind_rows(
  coef_plot_df %>%
    filter(domain == "Environmental Conditions") %>%
    slice_max(abs_estimate, n = 4, with_ties = FALSE),

  coef_plot_df %>%
    filter(domain == "Temporal Patterns") %>%
    slice_max(abs_estimate, n = 3, with_ties = FALSE),

  coef_plot_df %>%
    filter(domain == "Demand and Vessel Context") %>%
    slice_max(abs_estimate, n = 4, with_ties = FALSE)
) %>%
  distinct(term, .keep_all = TRUE) %>%
  arrange(abs_estimate)

# --- Plot absolute coefficients only ---
ggplot(
  coef_plot_top,
  aes(
    x = reorder(clean_label, abs_estimate),
    y = abs_estimate,
    fill = domain
  )
) +
  geom_col(width = 0.72) +
  coord_flip() +
  scale_fill_manual(
    values = c(
      "Environmental Conditions" = "#1A9988",
      "Temporal Patterns" = "#7A7A7A",
      "Demand and Vessel Context" = "#EB5600"
    )
  ) +
  labs(
    title = "Monitorable Signals Beyond Prior Delay",
    x = NULL,
    y = "Absolute standardized coefficient",
    fill = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "bottom",
    panel.grid.major.y = element_blank(),
    plot.title = element_text(face = "bold")
  )
```
```{r}
# --- Raw coefficient strength for Domain B / C / E ---

bce_predictors <- unique(c(
  domain_b_predictors,
  domain_c_predictors,
  domain_e_predictors
))

domain_lookup_bce <- tibble(
  variable = bce_predictors,
  domain = case_when(
    variable %in% domain_b_predictors ~ "Domain B: Environmental Conditions",
    variable %in% domain_c_predictors ~ "Domain C: Temporal Patterns",
    variable %in% domain_e_predictors ~ "Domain E: Demand and Vessel Context",
    TRUE ~ "Other"
  )
)

label_lookup <- c(
  env_temperature_c = "Temperature",
  env_wind_speed_mps = "Wind speed",
  env_visibility_m = "Visibility",
  env_precipitation_mm = "Precipitation",
  env_battery_water_level_obs = "Water level",
  env_battery_water_level_anomaly = "Water-level anomaly",
  env_narrows_current_obs = "Current",
  env_narrows_current_residual = "Current residual",

  prev_stop_sched_hour = "Scheduled hour",
  schedule_season = "Schedule season",
  schedule_daytype = "Schedule day type",

  vessel_capacity = "Vessel capacity",
  vessel_trip_seq_in_chain = "Vessel trip sequence",
  prev_stop_boardings = "Boardings",
  prev_stop_alightings = "Alightings",
  prev_stop_onboard = "Onboard count",
  prev_stop_load_factor = "Load factor"
)

get_base_var <- function(term, vars) {
  hits <- vars[vapply(vars, function(v) startsWith(term, v), logical(1))]
  if (length(hits) == 0) return(NA_character_)
  hits[which.max(nchar(hits))]
}

combined_model_df <- prepare_model_df(
  df_model,
  outcome_cont,
  final_predictors_trimmed
)

combined_formula <- as.formula(
  paste(outcome_cont, "~", paste(final_predictors_trimmed, collapse = " + "))
)

combined_raw_fit <- lm(combined_formula, data = combined_model_df)

coef_bce_raw <- broom::tidy(combined_raw_fit) %>%
  filter(term != "(Intercept)") %>%
  rowwise() %>%
  mutate(variable = get_base_var(term, bce_predictors)) %>%
  ungroup() %>%
  filter(!is.na(variable)) %>%
  left_join(domain_lookup_bce, by = "variable") %>%
  mutate(
    feature_label = if_else(
      variable %in% names(label_lookup),
      unname(label_lookup[variable]),
      variable
    ),
    feature_label = case_when(
      variable %in% c("schedule_daytype", "schedule_season") ~
        str_replace(term, variable, paste0(feature_label, " ")),
      TRUE ~ feature_label
    ),
    coefficient_strength = abs(estimate),
    abs_statistic = abs(statistic)
  ) %>%
  arrange(desc(coefficient_strength)) %>%
  select(
    domain,
    feature_label,
    term,
    coefficient_strength,
    std.error,
    abs_statistic,
    p.value
  )

coef_bce_raw %>%
  mutate(
    coefficient_strength = round(coefficient_strength, 4),
    std.error = round(std.error, 4),
    abs_statistic = round(abs_statistic, 2),
    p.value = round(p.value, 4)
  ) %>%
  kable(caption = "Raw Coefficient Strength for Domain B / C / E Features")
```



### 7.3 Temporal Patterns

The temporal model explains a modest share of variation (R² = 0.0129), indicating that day type, scheduled hour, and season capture recurring service rhythms but do not dominate the delay process on their own. These results suggest that delay does vary systematically across time, but that those temporal patterns are much weaker than the operational propagation and structural-context domains.

### 7.4 Operational Propagation

The operational-propagation model explains by far the largest share of variation among the individual domains (R² = 0.5887). This confirms that delay accumulation within and across trips is the strongest single explanatory component in the associative model. In other words, stop-level delay is highly path-dependent: once lateness has entered the operating chain, a substantial portion of subsequent delay can be understood as propagated rather than newly generated at each stop.

### 7.5 Vessel Characteristics

The vessel-characteristics model explains the smallest share of variation among the five domains (R² = 0.0085). This suggests that vessel-related differences are present but relatively limited when considered in isolation. Vessel effects therefore appear to be secondary to route structure, external conditions, and propagated delay in explaining stop-level lateness.



## 8) Final Combined Associative Model

### 8.1 Overall Model Fit

The final combined associative model explains a substantial share of stop-level delay variation (R² = 0.7295), clearly outperforming any single domain-specific model. This indicates that the five explanatory domains are complementary rather than interchangeable. While operational propagation provides the strongest single source of explanatory power, the full associative model benefits from combining structural, environmental, temporal, operational, and vessel-related information within one specification.

```{r}
# --- Standardized coefficient strength for Domain B / C / E ---

combined_model_df <- prepare_model_df(
  df_model,
  outcome_cont,
  final_predictors_trimmed
)

combined_formula <- as.formula(
  paste(outcome_cont, "~", paste(final_predictors_trimmed, collapse = " + "))
)

# Build dummy-expanded model matrix
x_mat <- model.matrix(combined_formula, data = combined_model_df)

# Remove intercept
x_mat <- x_mat[, colnames(x_mat) != "(Intercept)", drop = FALSE]

# Remove zero-variance columns
x_sd <- apply(x_mat, 2, sd, na.rm = TRUE)
x_mat <- x_mat[, is.finite(x_sd) & x_sd > 0, drop = FALSE]

# Standardize all predictors and outcome
x_scaled <- scale(x_mat)
y_scaled <- as.numeric(scale(combined_model_df[[outcome_cont]]))

# Fit standardized model
std_lm_df <- as.data.frame(x_scaled, check.names = FALSE)
std_lm_df$.y_scaled <- y_scaled

combined_std_fit <- lm(.y_scaled ~ ., data = std_lm_df)

coef_bce_standardized <- broom::tidy(combined_std_fit) %>%
  filter(term != "(Intercept)") %>%
  rowwise() %>%
  mutate(variable = get_base_var(term, bce_predictors)) %>%
  ungroup() %>%
  filter(!is.na(variable)) %>%
  left_join(domain_lookup_bce, by = "variable") %>%
  mutate(
    feature_label = if_else(
      variable %in% names(label_lookup),
      unname(label_lookup[variable]),
      variable
    ),
    feature_label = case_when(
      variable %in% c("schedule_daytype", "schedule_season") ~
        str_replace(term, variable, paste0(feature_label, " ")),
      TRUE ~ feature_label
    ),
    standardized_strength = abs(estimate),
    abs_statistic = abs(statistic)
  ) %>%
  arrange(desc(standardized_strength)) %>%
  select(
    domain,
    feature_label,
    term,
    standardized_strength,
    std.error,
    abs_statistic,
    p.value
  )

coef_bce_standardized %>%
  mutate(
    standardized_strength = round(standardized_strength, 4),
    std.error = round(std.error, 4),
    abs_statistic = round(abs_statistic, 2),
    p.value = round(p.value, 4)
  ) %>%
  kable(caption = "Standardized Coefficient Strength for Domain B / C / E Features")
```
```{r}
# --- Standardized coefficient strength bar chart ---

coef_strength_plot_df <- coef_bce_standardized %>%
  mutate(
    plot_label = case_when(
      feature_label == "Schedule day type WEEKEND" ~ "Weekend service",
      feature_label == "Previous scheduled hour" ~ "Scheduled hour",
      TRUE ~ feature_label
    ),
    domain_clean = case_when(
      domain == "Domain B: Environmental Conditions" ~ "Environmental Conditions",
      domain == "Domain C: Temporal Patterns" ~ "Temporal Patterns",
      domain == "Domain E: Demand and Vessel Context" ~ "Demand and Vessel Context",
      TRUE ~ domain
    )
  ) %>%
  slice_max(standardized_strength, n = 12, with_ties = FALSE) %>%
  arrange(standardized_strength)

ggplot(
  coef_strength_plot_df,
  aes(
    x = reorder(plot_label, standardized_strength),
    y = standardized_strength,
    fill = domain_clean
  )
) +
  geom_col(width = 0.5) +
  coord_flip() +
  scale_fill_manual(
    values = c(
      "Environmental Conditions" = "#1A9988",
      "Temporal Patterns" = "#C7C7C7",
      "Demand and Vessel Context" = "#EB5600"
    )
  ) +
  labs(
    title = "Standardized Signal Strength Across Monitorable Features",
    x = NULL,
    y = "Standardized coefficient strength",
    fill = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "bottom",
    legend.key.size = unit(0.45, "cm"),
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    plot.title = element_text(face = "bold", size = 14),
    axis.text.y = element_text(size = 10),
    axis.title.x = element_text(size = 11)
  )
```


### 8.2 Relative Importance Across Domains

The combined results suggest a clear hierarchy across domains. Operational propagation is the dominant explanatory component, followed by structural context. Environmental conditions and temporal patterns each explain a smaller but still measurable share of variation, while vessel characteristics contribute the least on their own. This pattern supports a layered interpretation of delay: some variation reflects structural baseline differences across the system, some reflects external operating conditions, and a large share reflects delay that is already being carried forward through the service chain.

### 8.3 Key Interpretable Signals

#### Environmental feature signals

Within the environmental domain, temperature is the strongest individual predictor by a substantial margin. Wind speed, observed water level, and observed current form a second tier of environmental signals, while visibility, current residual, and precipitation appear weaker. This shows that the environmental block is not uniform: even though it explains a relatively small share of total variance overall, some external-condition variables are clearly more informative than others.

#### Operational feature signals

Within the operational domain, both inherited previous-trip delay and carried within-trip delay remain strong positive signals, reinforcing the conclusion that stop-level delay is heavily cumulative. In practical terms, this means that once lateness enters the operating chain, it tends to persist across subsequent stops and linked trips.

#### Structural feature signals

Structural context remains the second most important explanatory domain after operational propagation. This indicates that route and stop configuration establish an important baseline pattern of delay across the system. In substantive terms, some portions of the ferry network appear systematically more delay-prone than others even before short-run operational carry-forward effects are considered.

### 8.4 Interpretation for Operations and Client Use

For the client, the associative model is best understood as a diagnostic tool rather than a forecasting tool. It helps explain why delay patterns differ across the system and separates three broad types of influence: structural baseline conditions, external operating conditions, and propagated operational delay. In this sense, the associative model supports interpretation and diagnosis, while the predictive model remains the main framework for forward-looking operational forecasting.

From an operational perspective, the results suggest three practical takeaways. First, propagated lateness is the dominant short-run driver of delay, so once the system is behind schedule, subsequent stops are likely to continue absorbing that delay. Second, structural context matters, which means that some routes and stop configurations begin from a less favorable operating baseline. Third, external conditions such as temperature, wind, water level, and current do not dominate total variance, but they still provide useful context for understanding when water-side operating conditions may be less favorable.


## 9) Save Outputs

```{r save}
write_csv(model_comparison_summary, file.path(output_dir, "model_comparison_summary.csv"))
write_csv(imputation_summary, file.path(output_dir, "ais_env_imputation_summary.csv"))
write_csv(env_feature_table, file.path(output_dir, "environmental_feature_table_standardized.csv"))

tibble(
  file = c(
    "model_comparison_summary.csv",
    "ais_env_imputation_summary.csv",
    "environmental_feature_table_standardized.csv"
  )
) %>% 
  kable(caption = "Saved Outputs for Associative Model Section")
```
