---
title: "Swiftly GPS Data Exploration 2"
author: "Tyler Maynard"
date: "2026-04-04"
output: 
  html_document:
    code_folding: hide
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE)
```

## Dataset Overview

The Swiftly GPS dataset contains location data every 10-60 seconds from the NYC Ferry AVL system. Columns include vessel, route, trip, block, and coordinates. This will be especially useful in calculating the speed of the vehicle at various points in time, helping identify locations, routes, and trips where the system struggles with on-time performance.

## Libraries

```{r libraries, warning=FALSE, message=FALSE}
library(tidyverse)    
library(sf)
library(kableExtra)
library(grid)
library(gridExtra)
library(viridis)
library(lubridate)
library(ggthemes)
library(ggspatial)
library(leaflet)
library(spatstat)
library(spdep)
library(htmlwidgets)
library(concaveman)
library(patchwork)
library(hms)
library(lwgeom)
library(duckdb)
library(data.table)
library(tidytransit)
library(gtfsio)

```

## Swiftly Data

The Swiftly GPS data set is available by month in CSV files. The following chunk of code reads all CSV files from January 2024 to June 2025, filters to the dates included in the sample week, and parses out each record's date and time into separate columns.

NOTE: The January 2024 data had three missing columns (none of which are important for this analysis), so those needed to be filtered out of the other files.

```{r data_import, warning=FALSE, message=FALSE}

# Setting Box Drive Root
box_root_path <- "C:\\Users\\tjama\\Box\\NYC_Ferry"

# Original csv

jan2024_original <- fread(file.path(box_root_path, "\\data\\NYCF_data_compiled\\swiftly_gps\\swiftly_gps", "swiftly_gps.2024_01.csv"))
feb2024_original <- fread(file.path(box_root_path, "\\data\\NYCF_data_compiled\\swiftly_gps\\swiftly_gps", "swiftly_gps.2024_02.csv"))
mar2024_original <- fread(file.path(box_root_path, "\\data\\NYCF_data_compiled\\swiftly_gps\\swiftly_gps", "swiftly_gps.2024_03.csv"))
apr2024_original <- fread(file.path(box_root_path, "\\data\\NYCF_data_compiled\\swiftly_gps\\swiftly_gps", "swiftly_gps.2024_04.csv"))
may2024_original <- fread(file.path(box_root_path, "\\data\\NYCF_data_compiled\\swiftly_gps\\swiftly_gps", "swiftly_gps.2024_05.csv"))
jun2024_original <- fread(file.path(box_root_path, "\\data\\NYCF_data_compiled\\swiftly_gps\\swiftly_gps", "swiftly_gps.2024_06.csv"))
jul2024_original <- fread(file.path(box_root_path, "\\data\\NYCF_data_compiled\\swiftly_gps\\swiftly_gps", "swiftly_gps.2024_07.csv"))
aug2024_original <- fread(file.path(box_root_path, "\\data\\NYCF_data_compiled\\swiftly_gps\\swiftly_gps", "swiftly_gps.2024_08.csv"))
sep2024_original <- fread(file.path(box_root_path, "\\data\\NYCF_data_compiled\\swiftly_gps\\swiftly_gps", "swiftly_gps.2024_09.csv"))
oct2024_original <- fread(file.path(box_root_path, "\\data\\NYCF_data_compiled\\swiftly_gps\\swiftly_gps", "swiftly_gps.2024_10.csv"))
nov2024_original <- fread(file.path(box_root_path, "\\data\\NYCF_data_compiled\\swiftly_gps\\swiftly_gps", "swiftly_gps.2024_11.csv"))
dec2024_original <- fread(file.path(box_root_path, "\\data\\NYCF_data_compiled\\swiftly_gps\\swiftly_gps", "swiftly_gps.2024_12.csv"))
jan2025_original <- fread(file.path(box_root_path, "\\data\\NYCF_data_compiled\\swiftly_gps\\swiftly_gps", "swiftly_gps.2025_01.csv"))
feb2025_original <- fread(file.path(box_root_path, "\\data\\NYCF_data_compiled\\swiftly_gps\\swiftly_gps", "swiftly_gps.2025_02.csv"))
mar2025_original <- fread(file.path(box_root_path, "\\data\\NYCF_data_compiled\\swiftly_gps\\swiftly_gps", "swiftly_gps.2025_03.csv"))
apr2025_original <- fread(file.path(box_root_path, "\\data\\NYCF_data_compiled\\swiftly_gps\\swiftly_gps", "swiftly_gps.2025_04.csv"))
may2025_original <- fread(file.path(box_root_path, "\\data\\NYCF_data_compiled\\swiftly_gps\\swiftly_gps", "swiftly_gps.2025_05.csv"))
jun2025_original <- fread(file.path(box_root_path, "\\data\\NYCF_data_compiled\\swiftly_gps\\swiftly_gps", "swiftly_gps.2025_06.csv"))

all_first <- rbindlist(list(
  feb2024_original, mar2024_original, apr2024_original,
  may2024_original, jun2024_original, jul2024_original,
  aug2024_original, sep2024_original, oct2024_original,
  nov2024_original, dec2024_original, jan2025_original,
  feb2025_original, mar2025_original, apr2025_original,
  may2025_original, jun2025_original
), fill = TRUE)  # fill = TRUE handles mismatched columns automatically

all_first <- all_first %>%
  select(-data.data.unpredictablestatus.type, -data.data.unpredictablestatus.details.blockId, -data.data.unpredictablestatus.details.serviceDate)

all_table <- rbindlist(list(
  jan2024_original, all_first
), fill = TRUE)  # fill = TRUE handles mismatched columns automatically

rm(box_root_path, all_first, jan2024_original, feb2024_original, mar2024_original, apr2024_original,
  may2024_original, jun2024_original, jul2024_original,
  aug2024_original, sep2024_original, oct2024_original,
  nov2024_original, dec2024_original, jan2025_original,
  feb2025_original, mar2025_original, apr2025_original,
  may2025_original, jun2025_original)

# Removing extra 'data.' from column names, removing unneeded columns
all_table <- all_table %>%
  rename_with(~ gsub("data.", "", .x)) %>%
  select(-(1:3), -(7))

# Remove NAs in route column
all_table <- all_table %>%
  filter(routeId %in% c("AS", "ER", "RES", "RS", "RWS", "SB", "SG")) %>%
  select("time","lat","lon","routeId","tripId","directionId","vehicleId","blockId","schedAdhMsec")

# Parse dates and times
all_table[, date_time := ymd_hms(time, tz = "America/New_York")]

all_table[, `:=`(
  date = as.Date(date_time),
  time = format(date_time, "%H:%M:%S"),
  hour = hour(date_time),
  dayOfWeek = lubridate::wday(date_time, label = TRUE, abbr = FALSE, week_start = 1),
  directionId = case_when(
    directionId == "1" ~ "NB",
    TRUE ~ "SB"
  ),
  joinId = paste(
    lubridate::wday(date_time, label = TRUE, abbr = FALSE, week_start = 1),
    routeId, tripId, directionId,
    sep = " "
  )
)]

all_table <- all_table[!is.na(tripId)][order(vehicleId, date_time)]

all_table <- all_table %>%
  select("date_time","date","dayOfWeek","time","hour","joinId","routeId","tripId","directionId","vehicleId","blockId","schedAdhMsec","lat","lon"
) %>%
  mutate(
    lat1 = lat,
    lon1 = lon
  )

```

## Shapefiles

As a way to add more of a spatial context to this analysis, the landings and routes (old system were extracted from the NYC Ferry and NYC OpenData websites.

```{r sf_objects, warning=FALSE, message=FALSE}

# Download file in GitHub and change path
landings <- read_sf("C:\\Users\\tjama\\Practicum\\Landings\\landings.shp") %>% 
  st_transform(32118)

# Download file in GitHub and change path
routes_new <- read_sf("C:\\Users\\tjama\\Practicum\\Landings\\routes.shp") %>%
  group_by(route_id) %>%
  summarise() %>% 
  st_transform(32118)

routes <- read_sf("C:\\Users\\tjama\\Practicum\\Landings\\NYCF_Routes.shp") %>%
  st_transform(32118) %>%
  mutate(
    route_id = case_when(
      Route_Name == "Astoria" ~ "AS",      
      Route_Name == "East River" ~ "ER", 
      Route_Name == "South Brooklyn" ~ "SB",   
      Route_Name == "St. George" ~ "SG",
      Route_Name == "Soundview" ~ "SV",      
      Route_Name == "Rockaway" ~ "RW",      
      TRUE ~ "GI"
    )
  )

```

## Data Cleaning

The following chunk does more data cleaning, creates sf object for each of the data points, and computes some new fields for feature engineering. Some features created in this chunk include the following:

-   Knots

    -   Based on the time and distance of the previous node of the same trip.

-   Key identifiers

    -   A number of these were created for different analyses, but generally at a base level include the date, trip, and route.

-   OTP

    -   This column classifies the on-time performance at each node based on the schedule adherence column (schedAdhMsec).

-   Excessively Late

    -   This is a binary field that assigns each record to a label of either excessively late (more than 10 minutes past the expected time) or not.

-   First Record Late

    -   Indicates if the trip starts out five minutes past its scheduled time.

-   Cumulative Distance and Time

    -   Computes the total distance and time duration of a trip.

To visualize the exact boundaries at which we se higher concentrations of excessively late data points, a concave polygon was created. This object could be useful (or not) for the future, but exists here if needed for future analyses.


```{r cleaning_up, warning=FALSE, message=FALSE}

all <- st_as_sf(all_table, coords = c("lon", "lat"), crs = 4326) %>%
  st_transform(32118)

xy <- st_coordinates(all)

# convert back to data.table, dropping geometry
all <- as.data.table(all)
all[, geometry := NULL]  # drop sf geometry column

# now add x and y
all[, `:=`(
  x = xy[, 1],
  y = xy[, 2]
)]

group_cols <- c("joinId")

setorder(all, joinId, date_time)

all[, `:=`(
  isFirst = seq_len(.N) == 1,
  isSecond = seq_len(.N) == 2,
  dist_to_prev_m = fifelse(
    seq_len(.N) == 1,
    0,
    round(sqrt((x - shift(x))^2 + (y - shift(y))^2), 2)
  ),
  time_diff_sec = fifelse(
    seq_len(.N) == 1,
    0,
    as.numeric(difftime(date_time, shift(date_time), units = "secs"))
  ),
  countDailyTrip = .N
), by = group_cols]

all[, `:=`(
  cum_dist_m = cumsum(dist_to_prev_m),
  cum_time_sec = cumsum(time_diff_sec)
), by = group_cols]

all[, m_per_sec := fifelse(
  is.na(time_diff_sec) | time_diff_sec <= 0,
  0,
  round(dist_to_prev_m / time_diff_sec, 2)
)]

all[, knots := m_per_sec * 1.94384]

all[, `:=`(
  date_route_trip_direction_vehicle_dow = paste(date, routeId, tripId, directionId, vehicleId, dayOfWeek, sep = " "),
  key_identifier = paste(date, blockId, routeId, tripId, directionId, dayOfWeek, sep = " "),
  dailyTrip = paste(dayOfWeek, routeId, tripId, sep = " "),
  OTP = fcase(
    schedAdhMsec < -60000, "Early",
    schedAdhMsec > 600000, "Excessively Late",
    schedAdhMsec > 300000, "Late",
    default = "Ontime"
  ),
  timeOfDay = hour(date_time)
)]

all[, excessivelyLate := fifelse(OTP == "Excessively Late", 1L, 0L)]

all[, firstRecordLate := fifelse(
  isSecond & OTP %chin% c("Late", "Excessively Late"),
  "Yes",
  "No"
)]

all[, OTP_prev := shift(OTP), by = group_cols]

all[, OTP_changed := fifelse(is.na(OTP_prev), FALSE, OTP != OTP_prev)]

all[, OTP_transition := fcase(
  isFirst, "Start",
  is.na(OTP_prev), "Start",
  OTP == OTP_prev, "No Change",
  default = paste0(OTP_prev, " -> ", OTP)
)]

final_cols <- c(
  "dailyTrip", "key_identifier", "date_route_trip_direction_vehicle_dow",
  "date", "time", "OTP", "OTP_transition",
  "excessivelyLate", "timeOfDay", "schedAdhMsec", "dayOfWeek",
  "dist_to_prev_m", "time_diff_sec", "knots", "lat1", "lon1"
)

all[, c("x", "y", "m_per_sec") := NULL]

setcolorder(
  all,
  c(
    final_cols,
    setdiff(names(all), c(final_cols, "date_time")),
    "date_time"
  )
)

```

The all_long computes the average distance and time traveled for each trip for each day, and includes the 7-day lag values for them. In situations where there data did not exist for 7 days prior (e.g., the first week of data available), the values for that day were used.

```{r pivot_vessel2, warning=FALSE, message=FALSE}

all_long <- all %>%
  group_by(date, joinId) %>%
  summarise(
    total_distance = max(cum_dist_m),
    total_time = max(cum_time_sec),
    knots = total_distance / total_time * 1.94384,
    .groups = "drop"
  )

lag_table <- all_long %>%
  select(date, joinId, total_distance, total_time, knots) %>%
  rename(
    lag_date = date,
    lag_7d_distance = total_distance,
    lag_7d_time = total_time,
    lag_7d_knots = knots
  )

all_long <- all_long %>%
  mutate(lag_date = date - 7) %>%
  left_join(lag_table, by = c("joinId", "lag_date")) %>%
  mutate(
    lag_7d_distance = coalesce(lag_7d_distance, total_distance),
    lag_7d_time = coalesce(lag_7d_time, total_time),
    lag_7d_knots = coalesce(lag_7d_knots, knots),
    change_in_distance = (total_distance - lag_7d_distance) / lag_7d_distance,
    change_in_time = (total_time - lag_7d_time) / lag_7d_time,
    change_in_knots = (knots - lag_7d_knots) / lag_7d_knots
  )

all_long <- all_long %>%
  mutate(
    change_in_distance = round(((total_distance - lag_7d_distance) / lag_7d_distance * 100), 2),
    change_in_time = round(((total_time - lag_7d_time) / total_time * 100), 2),
    change_in_knots = round(((knots - lag_7d_knots) / lag_7d_knots * 100), 2)
  )

```

## Segmentation of New Routes



```{r new_route_segments, warning=FALSE, message=FALSE}

gtfs <- import_gtfs("C:\\Users\\tjama\\Practicum\\GTFS\\gtfs.zip")

stops       <- gtfs$stops
stop_times  <- gtfs$stop_times
trips       <- gtfs$trips
routes      <- gtfs$routes
shapes      <- gtfs$shapes

# representative trip for each shape
rep_trips <- trips %>%
  filter(!is.na(shape_id)) %>%
  group_by(route_id, direction_id, shape_id) %>%
  slice(1) %>%
  ungroup() %>%
  select(trip_id, route_id, direction_id, shape_id)

segment_list <- lapply(seq_len(nrow(rep_trips)), function(j) {

  target_trip  <- rep_trips$trip_id[j]
  target_shape <- rep_trips$shape_id[j]
  target_route <- rep_trips$route_id[j]
  target_dir   <- rep_trips$direction_id[j]

  # shape line
  shape_line <- shapes %>%
    filter(shape_id == target_shape) %>%
    arrange(shape_pt_sequence) %>%
    st_as_sf(coords = c("shape_pt_lon", "shape_pt_lat"), crs = 4326) %>%
    summarise(do_union = FALSE) %>%
    st_cast("LINESTRING")

  # stops for this trip
  trip_stops <- stop_times %>%
    filter(trip_id == target_trip) %>%
    arrange(stop_sequence) %>%
    left_join(stops, by = "stop_id") %>%
    st_as_sf(coords = c("stop_lon", "stop_lat"), crs = 4326)

  # skip bad cases
  if (nrow(trip_stops) < 2 || nrow(shape_line) == 0) return(NULL)

  # project CRS
  shape_line <- st_transform(shape_line, 32118)
  trip_stops <- st_transform(trip_stops, 32118)

  # snap stops to line
  trip_stops_snap <- st_snap(trip_stops, shape_line, tolerance = 100)

  # locate stops along line
  locs <- sf::st_line_project(
    st_geometry(shape_line),
    st_geometry(trip_stops_snap),
    normalized = TRUE
  ) |> as.numeric()

  # if lengths mismatch, skip
  if (length(locs) != nrow(trip_stops_snap)) return(NULL)

  trip_stops_snap <- trip_stops_snap %>%
    mutate(loc = locs) %>%
    arrange(stop_sequence)

  # build stop-to-stop segments
  segs <- lapply(seq_len(nrow(trip_stops_snap) - 1), function(i) {
    lwgeom::st_linesubstring(
      st_geometry(shape_line)[[1]],
      trip_stops_snap$loc[i],
      trip_stops_snap$loc[i + 1]
    )
  })

  st_sf(
    trip_id      = target_trip,
    route_id     = target_route,
    direction_id = target_dir,
    shape_id     = target_shape,
    from_stop_id = trip_stops_snap$stop_id[-nrow(trip_stops_snap)],
    to_stop_id   = trip_stops_snap$stop_id[-1],
    from_seq     = trip_stops_snap$stop_sequence[-nrow(trip_stops_snap)],
    to_seq       = trip_stops_snap$stop_sequence[-1],
    geometry     = st_sfc(segs, crs = st_crs(shape_line))
  ) %>%
    mutate(segment_length = as.numeric(st_length(geometry)))
})

segments_sf <- do.call(rbind, segment_list) %>%
  rename(
    "routeId" = route_id
  )

segments_sf1 <- segments_sf %>%
  st_join(segments_sf, landings, join = st_nearest_feature)

st_write(segments_sf, "new_routes_segmentation.geojson", delete_dsn = TRUE)

# 1. Join segments within buffer
joined <- landings %>%
  st_join(
    segments_sf,
    join = st_is_within_distance,
    dist = 1,
    suffix = c("", "_line")
  ) %>%
  filter(!is.na(routeId))

# 2. Compute distance to each matched segment
matched_segments <- segments_sf[match(joined$routeId, segments_sf$routeId), ]

joined$dist_to_route <- st_distance(joined, matched_segments, by_element = TRUE)

# 3. Keep up to 2 closest segments per landing per route (one each direction)
#    falls back to 1 if only 1 segment is within buffer (end of line)
joined <- joined %>%
  group_by(, routeId) %>%      # replace landing_id with your point ID
  arrange(dist_to_route) %>%
  slice_head(n = 2) %>%                  # up to 2 (one per direction)
  ungroup()

ggplot() +
  geom_sf(data = segments_sf, aes(color = factor(segment_length))) +
  geom_sf(data = landings) +
  theme_solid()

```

## Segmentation of Old Routes

```{r route_segments, warning=FALSE, message=FALSE}

gtfs <- import_gtfs("C:\\Users\\tjama\\Practicum\\GTFS\\gtfs.zip")

stops       <- gtfs$stops
stop_times  <- gtfs$stop_times
trips       <- gtfs$trips
routes      <- gtfs$routes
shapes      <- gtfs$shapes

# representative trip for each shape
rep_trips <- trips %>%
  filter(!is.na(shape_id)) %>%
  group_by(route_id, direction_id, shape_id) %>%
  slice(1) %>%
  ungroup() %>%
  select(trip_id, route_id, direction_id, shape_id)

segment_list <- lapply(seq_len(nrow(rep_trips)), function(j) {

  target_trip  <- rep_trips$trip_id[j]
  target_shape <- rep_trips$shape_id[j]
  target_route <- rep_trips$route_id[j]
  target_dir   <- rep_trips$direction_id[j]

  # shape line
  shape_line <- shapes %>%
    filter(shape_id == target_shape) %>%
    arrange(shape_pt_sequence) %>%
    st_as_sf(coords = c("shape_pt_lon", "shape_pt_lat"), crs = 4326) %>%
    summarise(do_union = FALSE) %>%
    st_cast("LINESTRING")

  # stops for this trip
  trip_stops <- stop_times %>%
    filter(trip_id == target_trip) %>%
    arrange(stop_sequence) %>%
    left_join(stops, by = "stop_id") %>%
    st_as_sf(coords = c("stop_lon", "stop_lat"), crs = 4326)

  # skip bad cases
  if (nrow(trip_stops) < 2 || nrow(shape_line) == 0) return(NULL)

  # project CRS
  shape_line <- st_transform(shape_line, 32118)
  trip_stops <- st_transform(trip_stops, 32118)

  # snap stops to line
  trip_stops_snap <- st_snap(trip_stops, shape_line, tolerance = 100)

  # locate stops along line
  locs <- sf::st_line_project(
    st_geometry(shape_line),
    st_geometry(trip_stops_snap),
    normalized = TRUE
  ) |> as.numeric()

  # if lengths mismatch, skip
  if (length(locs) != nrow(trip_stops_snap)) return(NULL)

  trip_stops_snap <- trip_stops_snap %>%
    mutate(loc = locs) %>%
    arrange(stop_sequence)

  # build stop-to-stop segments
  segs <- lapply(seq_len(nrow(trip_stops_snap) - 1), function(i) {
    lwgeom::st_linesubstring(
      st_geometry(shape_line)[[1]],
      trip_stops_snap$loc[i],
      trip_stops_snap$loc[i + 1]
    )
  })

  st_sf(
    trip_id      = target_trip,
    route_id     = target_route,
    direction_id = target_dir,
    shape_id     = target_shape,
    from_stop_id = trip_stops_snap$stop_id[-nrow(trip_stops_snap)],
    to_stop_id   = trip_stops_snap$stop_id[-1],
    from_seq     = trip_stops_snap$stop_sequence[-nrow(trip_stops_snap)],
    to_seq       = trip_stops_snap$stop_sequence[-1],
    geometry     = st_sfc(segs, crs = st_crs(shape_line))
  ) %>%
    mutate(segment_length = as.numeric(st_length(geometry)))
})

segments_sf <- do.call(rbind, segment_list) %>%
  rename(
    "routeId" = route_id
  )

segments_sf1 <- segments_sf %>%
  st_join(segments_sf, landings, join = st_nearest_feature)

st_write(segments_sf, "new_routes_segmentation.geojson", delete_dsn = TRUE)

# 1. Join segments within buffer
joined <- landings %>%
  st_join(
    segments_sf,
    join = st_is_within_distance,
    dist = 1,
    suffix = c("", "_line")
  ) %>%
  filter(!is.na(routeId))

# 2. Compute distance to each matched segment
matched_segments <- segments_sf[match(joined$routeId, segments_sf$routeId), ]

joined$dist_to_route <- st_distance(joined, matched_segments, by_element = TRUE)

# 3. Keep up to 2 closest segments per landing per route (one each direction)
#    falls back to 1 if only 1 segment is within buffer (end of line)
joined <- joined %>%
  group_by(, routeId) %>%      # replace landing_id with your point ID
  arrange(dist_to_route) %>%
  slice_head(n = 2) %>%                  # up to 2 (one per direction)
  ungroup()

ggplot() +
  geom_sf(data = segments_sf, aes(color = factor(segment_length))) +
  geom_sf(data = landings) +
  theme_solid()

```