---
title: "Swiftly GPS Data Exploration - 6/16/25 to 6/22/25"
author: "Tyler Maynard"
date: "2026-02-17"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Dataset Overview

The Swiftly GPS dataset contains location data every 10-60 seconds from the NYC Ferry AVL system. Columns include vessel, route, trip, block, and coordinates. This will be especially useful in calculating the speed of the vehicle at various points in time, helping identify locations, routes, and trips where the system struggles with on-time performance.   

## Libraries

```{r libraries, warning=FALSE, message=FALSE}
library(dplyr)
library(tidyverse)
library(tidyr)
library(sf)
library(kableExtra)
library(grid)
library(gridExtra)
library(viridis)
library(lubridate)
library(ggthemes)
library(ggplot2)
library(ggspatial)
library(leaflet)
library(spatstat)
library(spdep)
library(htmlwidgets)
library(concaveman)
library(purrr)
library(stringr)
library(patchwork)

```
install.packages('concaveman')

June 16 - 22 2025 Reference Week

## Establishing Service Levels

```{r special_service, warning=FALSE, message=FALSE}

special_service <- data.frame(
  dates = as.Date(c("2025-06-19")))

```

## Swiftly Data

```{r data_import, warning=FALSE, message=FALSE}
# Setting Box Drive Root
BOX_ROOT <- "C:\\Users\\tjama\\Box\\NYC_Ferry"

# Original csv
jun2025_original <- read.csv(file.path(BOX_ROOT, "\\data\\NYCF_data_compiled\\swiftly_gps\\swiftly_gps", "swiftly_gps.2025_06.csv"))

# Removing extra 'data.' from column names, removing unneeded columns
jun2025 <- jun2025_original %>%
  rename_with(~ gsub("data.", "", .x)) %>%
  select(-(1:3), -(7))

# Converting to datetime
jun2025$time <- ymd_hms(jun2025$time)
jun2025$timeProcessed <- ymd_hms(jun2025$timeProcessed)

# Seeing how many unique values each column has
jun2025 %>%
  summarise(across(everything(), n_distinct)) %>%
  pivot_longer(cols = everything()) %>%
  kable()

# Parse dates and times
jun2025 <- jun2025 %>%
  mutate(
    date_time = ymd_hms(time, tz = "America/New_York"),
    processed_date_time = ymd_hms(timeProcessed, tz = "America/New_York"),
    
    date = as.Date(date_time),
    time = format(date_time, "%H:%M:%S"),
    hour = substr(time,1,2),
    
    dateProcessed = as.Date(processed_date_time),
    timeProcessed = format(processed_date_time, "%H:%M:%S"),
    
    dayOfWeek = if_else(wday(date, week_start = 1) %in% c(6,7), "Weekend", "Weekday"),
    dayOfWeek = case_when(
      date %in% special_service$date ~ "Holiday",
      TRUE ~ dayOfWeek)) %>%
  select(date, time, dateProcessed, timeProcessed, dayOfWeek, headsign, tripShortName, everything(), date_time, processed_date_time)

jun2025_reference_week <- jun2025 %>%
  filter(date %in% c("2025-06-16","2025-06-17","2025-06-18","2025-06-19","2025-06-20","2025-06-21","2025-06-22")) %>%
  filter(!is.na(tripId)) %>%  
#  filter(directionId == 0) %>%
#  filter(routeId == "SB") %>%
 # filter(substr(time, 1, 2) %in% c("13", "14", "15", "16", "17", "18", "19", "20", "21")) %>%
  arrange(vehicleId, date_time)   

str(jun2025_reference_week$time)

```

## Data Cleaning

```{r cleaning_up, warning=FALSE, message=FALSE}
# Create sf points and coordinates from the SAME jun2025_reference_week
pts <- st_as_sf(jun2025_reference_week, coords = c("lon", "lat"), crs = 4326, remove = FALSE) %>%
  st_transform(32118)

xy <- st_coordinates(pts)

stopifnot(nrow(jun2025_reference_week) == nrow(xy))               

jun2025_reference_week <- jun2025_reference_week %>%
  mutate(x = xy[, 1], y = xy[, 2]) %>%
  group_by(date, blockId, routeId, tripId, directionId, dayOfWeek) %>%
  arrange(date_time, .by_group = TRUE) %>%
  mutate(
    directionId = case_when(
      directionId == 1 ~ "NB",
      TRUE ~ "SB"
    ),
    isFirst = row_number() == 1,
    isSecond = row_number() == 2,
    dist_to_prev_m = round(sqrt((x - lag(x))^2 + (y - lag(y))^2), 2),
    time_diff_sec  = as.numeric(difftime(date_time, lag(date_time), units = "secs")),
    m_per_sec      = if_else(is.na(time_diff_sec) | time_diff_sec <= 0,
                             0,
                             round(dist_to_prev_m / time_diff_sec, 2)),
    knots = m_per_sec * 1.94384,
    date_route_trip_direction_vehicle_dow     = paste(date, routeId, tripId, directionId, vehicleId, dayOfWeek, sep = " "),
    key_identifier = paste(date, blockId, routeId, tripId, directionId, dayOfWeek, sep = " "),
    
    dailyTrip = paste(dayOfWeek, routeId, tripId, sep = " "),
    countDailyTrip = n(),
    OTP            = case_when(
      schedAdhMsec < -60000 ~ "Early",
      schedAdhMsec > 600000 ~ "Excessively Late",
      schedAdhMsec > 300000 ~ "Late",
      TRUE ~ "Ontime"
    ),
    excessivelyLate = case_when(
      OTP == "Excessively Late" ~ 1,
      TRUE ~ 0
    ),
    timeOfDay = as.integer(hour(date_time)),
    firstRecordLate = if_else(isSecond == 1 & (OTP == "Late" | OTP == "Excessively Late" ), "Yes", "No"),
      # previous OTP within the grouped+ordered trip
  OTP_prev = lag(OTP),
  # a simple "did it change?"
  OTP_changed = coalesce(OTP != OTP_prev, FALSE),
  # transition label like "Ontime -> Late"
  OTP_transition = case_when(
    isFirst ~ "Start",
    is.na(OTP_prev) ~ "Start",
    OTP == OTP_prev ~ "No Change",
    TRUE ~ paste0(OTP_prev, " -> ", OTP)
  ),
  ) %>%
  ungroup() %>%
  select(-x, -y, -m_per_sec) %>%
  select(dailyTrip, key_identifier, date_route_trip_direction_vehicle_dow, date, time, dateProcessed, timeProcessed, OTP, OTP_transition, excessivelyLate, timeOfDay, schedAdh, dayOfWeek,
         dist_to_prev_m, time_diff_sec, knots, headsign, tripShortName,
         everything(), date_time, processed_date_time)

summary_table <- data.frame(
  Statistic = names(summary(jun2025_reference_week$knots)),
  Value     = as.numeric(summary(jun2025_reference_week$knots)),
  row.names = NULL
)

 jun2025_reference_week_sf <- st_as_sf(jun2025_reference_week, coords = c("lon", "lat"), crs = 4326, remove = FALSE) %>%
  st_transform(32118)

kable(summary_table, digits = 2, caption = "Summary of Knots") %>%
  kable_styling(
    bootstrap_options = c("striped", "hover"),
    full_width = FALSE
  ) %>%
  column_spec(1:2, width = "150px")

jun2025_reference_week %>%
  count(excessivelyLate, sort = TRUE)
  
# Seeing how many unique values each column has
jun2025 %>%
  summarise(across(everything(), n_distinct)) %>%
  pivot_longer(cols = everything()) %>%
  kable()

jun2025_reference_week_excessively_lates <- jun2025_reference_week %>%
  filter(OTP == "Excessively Late")
  
jun2025_reference_week_excessively_lates_sf <- st_as_sf(jun2025_reference_week_excessively_lates, coords = c("lon", "lat"), crs = 4326, remove = FALSE) %>%
  st_transform(32118)

hull <- concaveman(jun2025_reference_week_excessively_lates_sf)
  
jun2025_reference_week %>%
  count(routeId)
```

## OTP by Direction

```{r er, warning=FALSE, message=FALSE}

jun2025_reference_week_dir <- jun2025_reference_week %>%
  select(dailyTrip, directionId, OTP, routeId) %>%
  filter(OTP == "Excessively Late")

jun2025_reference_week_dir %>%
  count(directionId, routeId)

```

<!-- ## East River Route -->

<!-- ```{r er, message=FALSE, warning=FALSE} -->

<!-- jun2025_reference_week <- jun2025_reference_week %>% -->
<!--   filter(routeId == "ER") -->

<!-- ``` -->

## Bar Charts

```{r bar_charts, warning=FALSE, message=FALSE}


excessivelyLatePlotVariables <- 
  jun2025_reference_week %>%
  as.data.frame() %>%
  select(excessivelyLate, dayOfWeek, routeId, timeOfDay, blockId, directionId, firstRecordLate) %>%
  gather(variable, value, -excessivelyLate) %>%
  filter(!is.na(value))

# Replace with dailyTrip for original
jun2025_reference_week_wide <- jun2025_reference_week %>%
  count(key_identifier, OTP) %>%
  pivot_wider(
    names_from = OTP, 
    values_from = n,
    values_fill = 0,
  ) %>% mutate(Total = rowSums(across(-key_identifier)))


jun2025_reference_week_wide

jun2025_reference_week_percents <- jun2025_reference_week_wide %>%
  mutate(round(across(-c(key_identifier, Total),
                ~ . / Total * 100), 2)) %>%
  pivot_longer(
    cols = -c(key_identifier, Total),
    names_to = "OTP",
    values_to = "Percent"
  ) %>%  filter(OTP == "Excessively Late", Percent > 0) %>%
  mutate(Percent_Rounded = round(Percent, 0))


mean_val <- mean(jun2025_reference_week_percents$Percent_Rounded)
median_val <- median(jun2025_reference_week_percents$Percent_Rounded)

ggplot() +
  geom_histogram(data = jun2025_reference_week_percents, color = "white", fill = "#006bb6", linewidth = 0.3, aes(x = Percent_Rounded), binwidth = 10) +
  geom_vline(xintercept = mean_val, color = "red2",linetype = "dashed", linewidth = 1) +
  geom_vline(xintercept = median_val, color = "darkgreen",linetype = "dashed", linewidth = 1) +
  annotate("text", x = mean_val, y = 15, label = "Mean", color = "red2", angle = 90, vjust = -0.5) +
  annotate("text", x = median_val, y = 15, label = "Median", color = "darkgreen", angle = 90, vjust = -0.5) +
 # geom_bar(data = jun2025_reference_week_percents, color = "#828282", fill = "#006bb6", linewidth = 0.3, aes(x = Percent_Rounded)) +
  labs(
    title = "Distribution of Excessively Late Trips by Time Spent Excessively Late",
    x = "Percent",
    y = "Count"
  )+
  theme_excel_new() +
  theme(
    axis.title = element_text(size = 14),
    plot.title = element_text(size = 20),
    plot.subtitle = element_text(size = 16),
    plot.title.position = "plot"
  )


plot_df <- excessivelyLatePlotVariables %>%
  group_by(excessivelyLate, variable, value) %>%
  summarise(count = n(), .groups = "drop") %>%
  group_by(excessivelyLate, variable) %>%
  mutate(prop = count / sum(count)) %>%
  ungroup()

ggplot(plot_df, aes(x = ifelse(variable == "timeOfDay",
                               as.numeric(value),
                               value), y = prop, fill = as.factor(excessivelyLate))) +
  geom_col(position = "dodge") +
  facet_wrap(~ variable, scales = "free_x") +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_manual(values = c("0" = "limegreen", "1" = "darkred"),
                    labels = c("Not Excessively Late","Excessively Late"),
                    name = "") +
  labs(x = NULL, y = "Percent within group") +
  theme_excel_new()

```

```{r bar_chart_by_route, warning=FALSE, message=FALSE}


jun2025_reference_week <- jun2025_reference_week %>%
  mutate(day_route = paste(dayOfWeek, routeId, sep = " "))

speed_by_route <- jun2025_reference_week %>%
  group_by(day_route, dayOfWeek, routeId) %>%
  summarise(
    avgKnots = mean(knots, na.rm = TRUE),
    medKnots = median(knots, na.rm = TRUE),
    n = dplyr::n(),
    .groups = "drop"
  )

ggplot() +
  geom_col(data = speed_by_route, aes(x = day_route, y = avgKnots, fill = dayOfWeek)) +
  labs(
    title = "Average Speed by Route and Service Level",
    x = "Service Level Month",
    y = "Average Knots"
  ) + 
  scale_fill_manual(
    values = c(
      "Holiday" = "limegreen",
      "Weekday" = "#006bb6",
      "Weekend" = "darkred"
    )
  ) +
  theme_excel_new()  +
  theme(
    plot.title = element_text(size = 20),
    plot.subtitle = element_text(size = 16),
    plot.title.position = "plot",
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

```


## Data Based on Identifier

```{r long_data, warning=FALSE, message=FALSE}

jun2025_reference_week_summary <- jun2025_reference_week %>%
  group_by(date_route_trip_direction_vehicle_dow) %>%
  summarise(
    trip_minutes = as.integer(
      difftime(max(date_time), min(date_time), units = "mins")
    ),
    .groups = "drop"
  ) 
```

## Shapefiles

```{r sf_objects, warning=FALSE, message=FALSE}

landings <- read_sf("C:\\Users\\tjama\\Practicum\\Landings\\landings.shp")

routes <- read_sf("C:\\Users\\tjama\\Practicum\\Landings\\routes.shp") %>%
  group_by(route_id) %>%
  summarise()

boroughs<- read_sf("C:\\Users\\tjama\\Practicum\\Boundaries\\boroughs.shp")


landings <- st_transform(landings, 32118)
landings_east_river <- landings %>%
  filter(!stop_id %in% c("37","39","43","45","46","48","49","50",
    "51","52","55","56","57","58","61","62",
    "16","88","103","104","105","142",
    "145","149","150","151"))
routes <- st_transform(routes, 32118)
boroughs <- st_transform(boroughs, 32118)

```

## Standardizing Plots

```{r bounding_boxes, warning=FALSE, message=FALSE}

bbox <- st_bbox(c(
  xmin = -74.30603027343751,
  ymin = 40.492915026895815,
  xmax = -73.64959716796876,
  ymax = 40.919739051062216
), crs = 4326) 

bbox_area <- st_as_sfc(bbox) %>%
  st_transform(32118)

bbox_service_area <- st_bbox(boroughs)

bbox_hull <- st_bbox(hull)

bbox_east_river <- st_bbox(landings_east_river)
bbox_expanded_east_river <- bbox_east_river
bbox_expanded_east_river["xmin"] <- bbox_east_river["xmin"] - 15000
bbox_expanded_east_river["ymin"] <- bbox_east_river["ymin"] - 15000
bbox_expanded_east_river["xmax"] <- bbox_east_river["xmax"] + 15000
bbox_expanded_east_river["ymax"] <- bbox_east_river["ymax"] + 15000
bbox_poly_east_river <- st_as_sfc(bbox_expanded_east_river)

```

## Kernel Density

```{r kd, dev='ragg_png', dpi=300, warning=FALSE, message=FALSE}

water <- read_sf("C:\\Users\\tjama\\Practicum\\Boundaries\\Hydrology\\Shape\\WBDHU2.shp") %>%
  st_transform(32118) %>%
  summarise(geometry = st_union(geometry)) %>%
  st_intersection(st_transform(st_as_sfc(st_bbox(bbox_expanded_east_river)), 32118))

fishnet <- 
  st_make_grid(water, cellsize = 250) %>%
  st_sf() %>%
  mutate(uniqueID = rownames(.))

jun2025_net <- st_join(
  fishnet, jun2025_reference_week_excessively_lates_sf, join = st_intersects) %>%
  group_by(uniqueID) %>%
  summarise(mean_speed = mean(knots, na.rm=TRUE),
            n = n(), .groups="drop",
) %>%
  filter(!is.na(mean_speed))

nb_queen <- poly2nb(jun2025_net, queen=TRUE)
lw_queen <- nb2listw(nb_queen, style = "W", zero.policy = TRUE)

local_moran <- localmoran(jun2025_net$mean_speed, lw_queen)
jun2025_net$Ii <- local_moran[,1]
jun2025_net$pvalue <- local_moran[,5]

jun2025_ppp <- as.ppp(st_coordinates(jun2025_reference_week_excessively_lates_sf), W = st_bbox(jun2025_net))
jun2025_KD <- density.ppp(jun2025_ppp, sigma = 1000)

#kd <- st_as_stars

as.data.frame(jun2025_KD) %>%
  st_as_sf(coords = c("x", "y"), crs = st_crs(jun2025_net)) %>%
  aggregate(., jun2025_net, mean) %>%
   ggplot() +
     geom_sf(aes(fill=value), color = NA, linewidth=0) +
     geom_sf(data = landings, color = "black", size = 1, alpha = 0.5) + 
     scale_fill_viridis(name = "Density") +
     labs(title = "Kernel density of 2017 burglaries") +
     theme_solid()

as.data.frame(jun2025_KD) %>%
  st_as_sf(coords = c("x", "y"), crs = st_crs(jun2025_net)) %>%
  aggregate(., jun2025_net, mean) %>%
  ggplot() +
    geom_sf(data = boroughs, color = "white") + 
  geom_sf(aes(fill = value), color = NA, linewidth = 0) +
#  geom_sf(data = routes, aes(color = as.factor(route_id)), linewidth = 0.5, fill = NA) +
  geom_sf(data = landings, color = "black", size = 1, alpha = 0.5) +
  coord_sf(
    xlim = c(bbox_east_river["xmin"] -15000, bbox_east_river["xmax"] +15000),
    ylim = c(bbox_east_river["ymin"] -15000, bbox_east_river["ymax"] +15000),
    expand = FALSE
  ) +
  facet_wrap(~ hour) +
  scale_fill_viridis_c(name = "Density") +
  scale_color_discrete(name = "Route") +
  labs(title = "Kernel density of 2017 burglaries") +
  theme_solid()

```

## Mean Speed

```{r mean_speed, warning=FALSE, message=FALSE}

ggplot(jun2025_net) +
  geom_sf(data = boroughs, color = "white") + 
  geom_sf(data = landings, color = "black", size = 2, alpha = 0.5) +
  geom_sf(aes(fill = mean_speed), color = NA) +
    coord_sf(
      xlim = c(bbox_hull["xmin"] - 100, bbox_hull["xmax"] + 100),
      ylim = c(bbox_hull["ymin"] - 200, bbox_hull["ymax"] + 200),
      expand = FALSE
    ) +
  scale_fill_viridis_c(option = "D", name = "Knots") +
  labs(
    title = "Mean Vessel Speed by Cell"
  ) +
  theme_solid() +
  theme(
    plot.title = element_text(size = 16),
    plot.subtitle = element_text(size = 16),
    plot.title.position = "plot")

```

## Kernel Density Hourly

```{r kd, dev='ragg_png', dpi=300, warning=FALSE, message=FALSE}
  
jun2025_net_13 <- st_join(
  fishnet, jun2025_reference_week_excessively_lates_sf, join = st_intersects) %>%
  group_by(uniqueID) %>%
  filter(hour == "13") %>%
  summarise(mean_speed = mean(knots, na.rm=TRUE),
            n = n(), .groups="drop",)
jun2025_net_13_ppp <- as.ppp(st_coordinates(jun2025_net_13), W = st_bbox(jun2025_net))
jun2025_net_13_KD <- density.ppp(jun2025_net_13_ppp, sigma = 1000)

jun2025_net_14 <- st_join(
  fishnet, jun2025_reference_week_excessively_lates_sf, join = st_intersects) %>%
  group_by(uniqueID) %>%
  filter(hour == "14") %>%
  summarise(mean_speed = mean(knots, na.rm=TRUE),
            n = n(), .groups="drop",)
jun2025_net_14_ppp <- as.ppp(st_coordinates(jun2025_net_14), W = st_bbox(jun2025_net))
jun2025_net_14_KD <- density.ppp(jun2025_net_14_ppp, sigma = 1000)
  
jun2025_net_15 <- st_join(
  fishnet, jun2025_reference_week_excessively_lates_sf, join = st_intersects) %>%
  group_by(uniqueID) %>%
  filter(hour == "15") %>%
  summarise(mean_speed = mean(knots, na.rm=TRUE),
            n = n(), .groups="drop",)
jun2025_net_15_ppp <- as.ppp(st_coordinates(jun2025_net_15), W = st_bbox(jun2025_net))
jun2025_net_15_KD <- density.ppp(jun2025_net_15_ppp, sigma = 1000)

jun2025_net_16 <- st_join(
  fishnet, jun2025_reference_week_excessively_lates_sf, join = st_intersects) %>%
  group_by(uniqueID) %>%
  filter(hour == "16") %>%
  summarise(mean_speed = mean(knots, na.rm=TRUE),
            n = n(), .groups="drop",)
jun2025_net_16_ppp <- as.ppp(st_coordinates(jun2025_net_16), W = st_bbox(jun2025_net))
jun2025_net_16_KD <- density.ppp(jun2025_net_16_ppp, sigma = 1000)
  
jun2025_net_17 <- st_join(
  fishnet, jun2025_reference_week_excessively_lates_sf, join = st_intersects) %>%
  group_by(uniqueID) %>%
  filter(hour == "17") %>%
  summarise(mean_speed = mean(knots, na.rm=TRUE),
            n = n(), .groups="drop",)
jun2025_net_17_ppp <- as.ppp(st_coordinates(jun2025_net_17), W = st_bbox(jun2025_net))
jun2025_net_17_KD <- density.ppp(jun2025_net_17_ppp, sigma = 1000)

jun2025_net_18 <- st_join(
  fishnet, jun2025_reference_week_excessively_lates_sf, join = st_intersects) %>%
  group_by(uniqueID) %>%
  filter(hour == "18") %>%
  summarise(mean_speed = mean(knots, na.rm=TRUE),
            n = n(), .groups="drop",)
jun2025_net_18_ppp <- as.ppp(st_coordinates(jun2025_net_18), W = st_bbox(jun2025_net))
jun2025_net_18_KD <- density.ppp(jun2025_net_18_ppp, sigma = 1000)
  
jun2025_net_19 <- st_join(
  fishnet, jun2025_reference_week_excessively_lates_sf, join = st_intersects) %>%
  group_by(uniqueID) %>%
  filter(hour == "19") %>%
  summarise(mean_speed = mean(knots, na.rm=TRUE),
            n = n(), .groups="drop",)
jun2025_net_19_ppp <- as.ppp(st_coordinates(jun2025_net_19), W = st_bbox(jun2025_net))
jun2025_net_19_KD <- density.ppp(jun2025_net_19_ppp, sigma = 1000)

jun2025_net_20 <- st_join(
  fishnet, jun2025_reference_week_excessively_lates_sf, join = st_intersects) %>%
  group_by(uniqueID) %>%
  filter(hour == "20") %>%
  summarise(mean_speed = mean(knots, na.rm=TRUE),
            n = n(), .groups="drop",)
jun2025_net_20_ppp <- as.ppp(st_coordinates(jun2025_net_20), W = st_bbox(jun2025_net))
jun2025_net_20_KD <- density.ppp(jun2025_net_20_ppp, sigma = 1000)
  
jun2025_net_21 <- st_join(
  fishnet, jun2025_reference_week_excessively_lates_sf, join = st_intersects) %>%
  group_by(uniqueID) %>%
  filter(hour == "21") %>%
  summarise(mean_speed = mean(knots, na.rm=TRUE),
            n = n(), .groups="drop",)
jun2025_net_21_ppp <- as.ppp(st_coordinates(jun2025_net_21), W = st_bbox(jun2025_net))
jun2025_net_21_KD <- density.ppp(jun2025_net_21_ppp, sigma = 1000)



as.data.frame(kd_df) %>%
  st_as_sf(coords = c("x", "y"), crs = st_crs(jun2025_net)) %>%
  aggregate(., jun2025_net, mean) %>%
  ggplot() +
    geom_sf(data = boroughs, color = "white") + 
  geom_sf(aes(fill = value), color = NA, linewidth = 0) +
  geom_sf(data = landings, color = "black", size = 1, alpha = 0.5) +
 # facet_wrap(~ hour, ncol = 6) +
  coord_sf(
    xlim = c(bbox_east_river["xmin"] -15000, bbox_east_river["xmax"] +15000),
    ylim = c(bbox_east_river["ymin"] -15000, bbox_east_river["ymax"] +15000),
    expand = FALSE
  ) +
  scale_fill_viridis_c(name = "Density") +
  scale_color_discrete(name = "Route") +
  labs(title = "Kernel Density of Excessively Late Records") +
  theme_solid()


```

## Kernel Plots

```{r kd_plots, dev='ragg_png', dpi=300, warning=FALSE, message=FALSE}

plot13 <- as.data.frame(jun2025_net_13_KD) %>%
  st_as_sf(coords = c("x", "y"), crs = st_crs(jun2025_net_13)) %>%
  aggregate(., jun2025_net_13, mean) %>%
  ggplot() +
    geom_sf(data = boroughs, color = "white") + 
  geom_sf(aes(fill = value), color = NA, linewidth = 0) +
#  geom_sf(data = landings, color = "black", size = 1, alpha = 0.5) +
  coord_sf(
    xlim = c(bbox_hull["xmin"] -100, bbox_hull["xmax"] +100),
    ylim = c(bbox_hull["ymin"] -100, bbox_hull["ymax"] +100),
    expand = FALSE
  ) +
  scale_fill_viridis_c(name = "") +
  scale_color_discrete(name = "Route") +
  labs(title = "Kernel Density - 1:00PM") +
  theme_solid() +
  theme(
    plot.title = element_text(size = 16),
    plot.subtitle = element_text(size = 16),
    plot.title.position = "plot"
  )

plot14 <- as.data.frame(jun2025_net_14_KD) %>%
  st_as_sf(coords = c("x", "y"), crs = st_crs(jun2025_net_14)) %>%
  aggregate(., jun2025_net_14, mean) %>%
  ggplot() +
    geom_sf(data = boroughs, color = "white") + 
  geom_sf(aes(fill = value), color = NA, linewidth = 0) +
#  geom_sf(data = landings, color = "black", size = 1, alpha = 0.5) +
  coord_sf(
    xlim = c(bbox_hull["xmin"] -100, bbox_hull["xmax"] +100),
    ylim = c(bbox_hull["ymin"] -100, bbox_hull["ymax"] +100),
    expand = FALSE
  ) +
  scale_fill_viridis_c(name = "") +
  scale_color_discrete(name = "Route") +
  labs(title = "Kernel Density - 2:00PM") +
  theme_solid() +
  theme(
    plot.title = element_text(size = 16),
    plot.subtitle = element_text(size = 16),
    plot.title.position = "plot"
  )

plot15 <- as.data.frame(jun2025_net_15_KD) %>%
  st_as_sf(coords = c("x", "y"), crs = st_crs(jun2025_net_15)) %>%
  aggregate(., jun2025_net_15, mean) %>%
  ggplot() +
    geom_sf(data = boroughs, color = "white") + 
  geom_sf(aes(fill = value), color = NA, linewidth = 0) +
#  geom_sf(data = landings, color = "black", size = 1, alpha = 0.5) +
  coord_sf(
    xlim = c(bbox_hull["xmin"] -100, bbox_hull["xmax"] +100),
    ylim = c(bbox_hull["ymin"] -100, bbox_hull["ymax"] +100),
    expand = FALSE
  ) +
  scale_fill_viridis_c(name = "") +
  scale_color_discrete(name = "Route") +
  labs(title = "Kernel Density - 3:00PM") +
  theme_solid() +
  theme(
    plot.title = element_text(size = 16),
    plot.subtitle = element_text(size = 16),
    plot.title.position = "plot"
  )

plot16 <- as.data.frame(jun2025_net_16_KD) %>%
  st_as_sf(coords = c("x", "y"), crs = st_crs(jun2025_net_16)) %>%
  aggregate(., jun2025_net_16, mean) %>%
  ggplot() +
    geom_sf(data = boroughs, color = "white") + 
  geom_sf(aes(fill = value), color = NA, linewidth = 0) +
#  geom_sf(data = landings, color = "black", size = 1, alpha = 0.5) +
  coord_sf(
    xlim = c(bbox_hull["xmin"] -100, bbox_hull["xmax"] +100),
    ylim = c(bbox_hull["ymin"] -100, bbox_hull["ymax"] +100),
    expand = FALSE
  ) +
  scale_fill_viridis_c(name = "") +
  scale_color_discrete(name = "Route") +
  labs(title = "Kernel Density - 4:00PM") +
  theme_solid() +
  theme(
    plot.title = element_text(size = 16),
    plot.subtitle = element_text(size = 16),
    plot.title.position = "plot"
  )

plot17 <- as.data.frame(jun2025_net_17_KD) %>%
  st_as_sf(coords = c("x", "y"), crs = st_crs(jun2025_net_17)) %>%
  aggregate(., jun2025_net_17, mean) %>%
  ggplot() +
    geom_sf(data = boroughs, color = "white") + 
  geom_sf(aes(fill = value), color = NA, linewidth = 0) +
#  geom_sf(data = landings, color = "black", size = 1, alpha = 0.5) +
  coord_sf(
    xlim = c(bbox_hull["xmin"] -100, bbox_hull["xmax"] +100),
    ylim = c(bbox_hull["ymin"] -100, bbox_hull["ymax"] +100),
    expand = FALSE
  ) +
  scale_fill_viridis_c(name = "") +
  scale_color_discrete(name = "Route") +
  labs(title = "Kernel Density - 5:00PM") +
  theme_solid() +
  theme(
    plot.title = element_text(size = 16),
    plot.subtitle = element_text(size = 16),
    plot.title.position = "plot"
  )

plot18 <- as.data.frame(jun2025_net_18_KD) %>%
  st_as_sf(coords = c("x", "y"), crs = st_crs(jun2025_net_18)) %>%
  aggregate(., jun2025_net_18, mean) %>%
  ggplot() +
    geom_sf(data = boroughs, color = "white") + 
  geom_sf(aes(fill = value), color = NA, linewidth = 0) +
#  geom_sf(data = landings, color = "black", size = 1, alpha = 0.5) +
  coord_sf(
    xlim = c(bbox_hull["xmin"] -100, bbox_hull["xmax"] +100),
    ylim = c(bbox_hull["ymin"] -100, bbox_hull["ymax"] +100),
    expand = FALSE
  ) +
  scale_fill_viridis_c(name = "") +
  scale_color_discrete(name = "Route") +
  labs(title = "Kernel Density - 6:00PM") +
  theme_solid() +
  theme(
    plot.title = element_text(size = 16),
    plot.subtitle = element_text(size = 16),
    plot.title.position = "plot"
  )

plot19 <- as.data.frame(jun2025_net_19_KD) %>%
  st_as_sf(coords = c("x", "y"), crs = st_crs(jun2025_net_19)) %>%
  aggregate(., jun2025_net_19, mean) %>%
  ggplot() +
    geom_sf(data = boroughs, color = "white") + 
  geom_sf(aes(fill = value), color = NA, linewidth = 0) +
#  geom_sf(data = landings, color = "black", size = 1, alpha = 0.5) +
  coord_sf(
    xlim = c(bbox_hull["xmin"] -100, bbox_hull["xmax"] +100),
    ylim = c(bbox_hull["ymin"] -100, bbox_hull["ymax"] +100),
    expand = FALSE
  ) +
  scale_fill_viridis_c(name = "") +
  scale_color_discrete(name = "Route") +
  labs(title = "Kernel Density - 7:00PM") +
  theme_solid() +
  theme(
    plot.title = element_text(size = 16),
    plot.subtitle = element_text(size = 16),
    plot.title.position = "plot"
  )

plot20 <- as.data.frame(jun2025_net_20_KD) %>%
  st_as_sf(coords = c("x", "y"), crs = st_crs(jun2025_net_20)) %>%
  aggregate(., jun2025_net_20, mean) %>%
  ggplot() +
    geom_sf(data = boroughs, color = "white") + 
  geom_sf(aes(fill = value), color = NA, linewidth = 0) +
#  geom_sf(data = landings, color = "black", size = 1, alpha = 0.5) +
  coord_sf(
    xlim = c(bbox_hull["xmin"] -100, bbox_hull["xmax"] +100),
    ylim = c(bbox_hull["ymin"] -100, bbox_hull["ymax"] +100),
    expand = FALSE
  ) +
  scale_fill_viridis_c(name = "") +
  scale_color_discrete(name = "Route") +
  labs(title = "Kernel Density - 8:00PM") +
  theme_solid() +
  theme(
    plot.title = element_text(size = 16),
    plot.subtitle = element_text(size = 16),
    plot.title.position = "plot"
  )

plot21 <- as.data.frame(jun2025_net_21_KD) %>%
  st_as_sf(coords = c("x", "y"), crs = st_crs(jun2025_net_21)) %>%
  aggregate(., jun2025_net_21, mean) %>%
  ggplot() +
    geom_sf(data = boroughs, color = "white") + 
  geom_sf(aes(fill = value), color = NA, linewidth = 0) +
#  geom_sf(data = landings, color = "black", size = 1, alpha = 0.5) +
  coord_sf(
    xlim = c(bbox_hull["xmin"] -100, bbox_hull["xmax"] +100),
    ylim = c(bbox_hull["ymin"] -100, bbox_hull["ymax"] +100),
    expand = FALSE
  ) +
  scale_fill_viridis_c(name = "") +
  scale_color_discrete(name = "Route") +
  labs(title = "Kernel Density - 9:00PM") +
  theme_solid() +
  theme(
    plot.title = element_text(size = 16),
    plot.subtitle = element_text(size = 16),
    plot.title.position = "plot"
  )

plots <- list(plot13,
              plot14,
              plot15,
              plot16,
              plot17,
              plot18,
              plot19,
              plot20,
              plot21
)

wrap_plots(plots, ncol = 3)
theme(legend.position = "right")


# Actual Plot

kd_list <- list(
  '1PM' = jun2025_net_13_KD,
  '2PM' = jun2025_net_14_KD,
  '3PM' = jun2025_net_15_KD,
  '4PM' = jun2025_net_16_KD,
  '5PM' = jun2025_net_17_KD,
  '6PM' = jun2025_net_18_KD,
  '7PM' = jun2025_net_19_KD,
  '8PM' = jun2025_net_20_KD,
  '9PM' = jun2025_net_21_KD
)

all_values <- unlist(lapply(kd_list, \(kd) as.data.frame(kd)$value))
global_limits <- range(all_values, na.rm = TRUE)

plots <- imap(kd_list, ~{
  df <- as.data.frame(.x)  # columns: x, y, value

  sf_pts <- st_as_sf(df, coords = c("x","y"), crs = st_crs(jun2025_net))

  # aggregate ONLY the value column so the output column name stays clean
  agg <- aggregate(sf_pts["value"], jun2025_net, mean, na.rm = TRUE)

  ggplot() +
    geom_sf(data = boroughs, color = "white") +
    geom_sf(data = agg, aes(fill = value), color = NA, linewidth = 0) +
    coord_sf(
      xlim = c(bbox_hull["xmin"] - 100, bbox_hull["xmax"] + 100),
      ylim = c(bbox_hull["ymin"] - 100, bbox_hull["ymax"] + 100),
      expand = FALSE
    ) +
    scale_fill_viridis_c(
      name = "Kernel Density",
      limits = global_limits,
      oob = scales::squish
    ) +
    labs(title = paste("KD -", .y), subtitle = "Excessively Lates", caption = "250m Cell Size") +
    theme_solid() +
  theme(
    plot.title = element_text(size = 16),
    plot.subtitle = element_text(size = 12),
    plot.caption = element_text(size = 12),
    plot.title.position = "plot"
  )
})

wrap_plots(plots, ncol = 3) +
  plot_layout(guides = "collect") &
  theme(legend.position = "right")




```

## Plotting By Hour

```{r kd_hourly, dev='ragg_png', dpi=300, warning=FALSE, message=FALSE}

jun2025_net_hourly <- jun2025_net %>%
  st_drop_geometry() %>%
  distinct(uniqueID) %>%                      # base cells
  tidyr::crossing(hour = 0:23) %>%            # repeat each cell for each hour
  left_join(
    jun2025_net %>% select(uniqueID, geometry) %>% distinct(),
    by = "uniqueID"
  ) %>%
  st_as_sf()

kd_hourly <- as.data.frame(jun2025_KD) %>%
  st_as_sf(coords = c("x", "y"), crs = st_crs(jun2025_net)) %>%
  aggregate(., jun2025_net_hourly, mean)

ggplot() +
  geom_sf(data = boroughs, color = "white") +
  geom_sf(data = kd_hourly, aes(fill = value), color = NA, linewidth = 0) +
  geom_sf(data = landings, color = "black", size = 1, alpha = 0.5) +
  coord_sf(
    xlim = c(bbox_east_river["xmin"] - 15000, bbox_east_river["xmax"] + 15000),
    ylim = c(bbox_east_river["ymin"] - 15000, bbox_east_river["ymax"] + 15000),
    expand = FALSE
  ) +
  facet_wrap(~ hour) +
  scale_fill_viridis_c(name = "Density") +
  labs(title = "Kernel density by hour") +
  theme_solid()

```

## Moran's I

mean(jun2025_net$n_records)

table(jun2025_net$pvalue < 0.05, useNA = "ifany")


```{r moran, warning=FALSE, message=FALSE}

jun2025_net$z_value <- scale(jun2025_net$mean_speed)[,1]


jun2025_net$lag_z <- lag.listw(lw_queen, jun2025_net$z_value, zero.policy = TRUE)


jun2025_net$cluster <- NA

sig <- jun2025_net$pvalue < 0.05

jun2025_net$cluster[sig & jun2025_net$z > 0 & jun2025_net$lag_z > 0] <- "High-High"
jun2025_net$cluster[sig & jun2025_net$z < 0 & jun2025_net$lag_z < 0] <- "Low-Low"
jun2025_net$cluster[sig & jun2025_net$z > 0 & jun2025_net$lag_z < 0] <- "High-Low"
jun2025_net$cluster[sig & jun2025_net$z < 0 & jun2025_net$lag_z > 0] <- "Low-High"

ggplot(jun2025_net) +
  geom_sf(aes(fill = cluster), color = NA) +
  coord_sf(
    xlim = c(bbox_area["xmin"], bbox_area["xmax"]),
    ylim = c(bbox_area["ymin"], bbox_area["ymax"]),
    expand = FALSE
  ) +
  scale_fill_manual(
    values = c(
      "High-High" = "red",
      "Low-Low"   = "blue",
      "High-Low"  = "orange",
      "Low-High"  = "lightblue"
    ),
    na.value = "grey90"
  ) +
  geom_sf(data = landings, color = "black", size = 1, alpha = 0.5) + 
  labs(title = "Local Moran’s I Cluster Map") +
  theme_minimal()



```

```{r label_standards, warning=FALSE, message=FALSE}

ranges_quartiles <- quartiles_sf %>%
  group_by(group) %>%
  summarise(
    min_val = round(min(knots, na.rm = TRUE), 2),
    max_val = round(max(knots, na.rm = TRUE), 2),
    n = n()
  )

labels_vec <- paste0(
  ranges_quartiles$group,
  " (",
  ranges_quartiles$min_val,
  " – ",
  ranges_quartiles$max_val,
  " m/s, n = ",
  ranges_quartiles$n,
  ")"
)

names(labels_vec) <- ranges_quartiles$group

ranges_Q1_Q2 <- Q1_Q2_sf %>%
  group_by(group) %>%
  summarise(
    min_val = round(min(knots, na.rm = TRUE), 2),
    max_val = round(max(knots, na.rm = TRUE), 2),
    n = n()
  )

labels_vec_Q1_Q2 <- paste0(
  ranges_Q1_Q2$group,
  " (",
  ranges_Q1_Q2$min_val,
  " – ",
  ranges_Q1_Q2$max_val,
  " m/s, n = ",
  ranges_Q1_Q2$n,
  ")"
)

names(labels_vec_Q1_Q2) <- ranges_Q1_Q2$group


```

## Plots

```{r plots, warning=FALSE, message=FALSE}
ggplot()+
  geom_sf(data = boroughs, color = "white") +  
  geom_sf(data = landings, color = "black", size = 3) +
   labs(
    title = "NYC Ferry Landings"
  ) +
  theme_solid() +
  theme(
    plot.title = element_text(size = 16),
    plot.title.position = "plot"
  )
```

```{r plot_transitions, warning=FALSE, message=FALSE}

jun2025_reference_week_late_sf <- jun2025_reference_week_sf %>%
  filter(OTP_transition %in% c("Early -> Late", "Late -> Excessively Late", "Ontime -> Late", "Ontime -> Excessively Late"
))

ggplot()+
  geom_sf(data = boroughs, color = "white") +  
  geom_sf(data = landings, color = "black", size = 2, alpha = 0.5) +
  geom_sf(data = jun2025_reference_week_late_sf,
          aes(color = OTP_transition), size = 2, alpha = 0.2) +
  coord_sf(
    xlim = c(bbox_east_river["xmin"] - 10000, bbox_east_river["xmax"]+ 10000),
    ylim = c(bbox_east_river["ymin"] - 10000, bbox_east_river["ymax"] + 10000),
    expand = FALSE
  ) +
   labs(
    title = "OTP Transition Locations"
  ) +
  scale_color_manual(
    name = "Transition Type",
    values = c(
      "Early -> Late" = "orange",
      "Late -> Excessively Late" = "red"),
    labels = c(
      "Early -> Late" = "Early to Late",
      "Late -> Excessively Late" = "Late to Excessively Late"
    )
  ) +
  facet_wrap(~OTP_transition) +
  theme_solid() +
  theme(
    plot.title = element_text(size = 20),
    plot.title.position = "plot"
  )

ggplot()+
  geom_sf(data = boroughs, color = "white") +  
  geom_sf(data = landings, color = "black", size = 2, alpha = 0.2) +
  geom_sf(data = jun2025_reference_week_late_sf,
          aes(color = OTP_transition), size = 2, alpha = 0.2) +
  coord_sf(
    xlim = c(bbox_east_river["xmin"] - 10000, bbox_east_river["xmax"]+ 10000),
    ylim = c(bbox_east_river["ymin"] - 10000, bbox_east_river["ymax"] + 10000),
    expand = FALSE
  ) +
   labs(
    title = "OTP Transition Locations"
  ) +
  scale_color_manual(
    name = "Transition Type",
    values = c(
      "Early -> Late" = "#1e3888",
      "Ontime -> Late" = "#47a8bd",
      "Ontime -> Excessively Late" = "#ffad69",
      "Late -> Excessively Late" = "#9c3848"),
    labels = c(
      "Early -> Late" = "Early to Late",
      "Ontime -> Late" = "Ontime to Late",
      "Ontime -> Excessively Late" = "Ontime to Excessively Late",
      "Late -> Excessively Late" = "Late to Excessively Late"
    )
  ) +
  facet_wrap(~OTP_transition) +
  theme_solid() +
  theme(
    plot.title = element_text(size = 20),
    plot.title.position = "plot"
  )

unique(jun2025_reference_week_sf$OTP_transition)
```

```{r plot_excessively_lates, warning=FALSE, message=FALSE}
ggplot()+
  geom_sf(data = boroughs, color = "white") +  
 # geom_sf(data = landings, color = "black", size = 3) +
  geom_sf(data = jun2025_reference_week_sf, aes(fill = OTP_transition, color = "red", size = 10, linewidth = 2)) +
 # geom_sf(data = jun2025_reference_week_excessively_lates_sf, color = "blue", size = 1) +
  coord_sf(
    xlim = c(bbox_east_river["xmin"] - 10000, bbox_east_river["xmax"]+ 10000),
    ylim = c(bbox_east_river["ymin"] - 10000, bbox_east_river["ymax"] + 10000),
    expand = FALSE
  ) +
   labs(
    title = "Excessivlely Late 'Zone'"
  ) +
  theme_solid() +
  theme(
    plot.title = element_text(size = 20),
    plot.title.position = "plot"
  )

```

```{r plot_speed_quartiles, warning=FALSE, message=FALSE}
ggplot() +
  geom_sf(data = boroughs, color = "white") +  
  geom_sf(data = landings, color = "black", size = 2, alpha = 0.1) +
  geom_sf(data = quartiles_sf,
          aes(color = group),
          size = 1, alpha = 0.01) +
  coord_sf(
    xlim = c(bbox_quartiles["xmin"], bbox_quartiles["xmax"]),
    ylim = c(bbox_quartiles["ymin"], bbox_quartiles["ymax"]),
    expand = FALSE
  ) +
  scale_color_manual(
    values = c("Q1" = "#1e3888",
               "Q2" = "#47a8bd",
               "Q3" = "#ffad69",
               "Q4" = "#9c3848"),
    labels = labels_vec,
    name = "Speed Quartile"
  ) +
  guides(color = guide_legend(override.aes = list(alpha = 1, size = 2))) +
  facet_wrap(~ group) +
  labs(
    title = "Small Multiple of Speed Quartiles",
    subtitle = "January 02, 2024"
  ) +
  theme_solid() +
  theme(
    plot.title = element_text(size = 16),
    plot.subtitle = element_text(size = 16),
    plot.title.position = "plot"
  )
```

```{r plot_higher_quartiles, warning=FALSE, message=FALSE}

ggplot() +
  geom_sf(data = boroughs, color = "white") +  
  geom_sf(data = landings, color = "black", size = 2, alpha = 0.1) +
  geom_sf(data = Q1_Q2_sf,
          aes(color = group),
          size = 1, alpha = 0.01) +
  coord_sf(
    xlim = c(bbox_Q1_Q2["xmin"], bbox_Q1_Q2["xmax"]),
    ylim = c(bbox_Q1_Q2["ymin"], bbox_Q1_Q2["ymax"]),
    expand = FALSE
  ) +
  scale_color_manual(
    values = c("Q1" = "#1e3888",
               "Q2" = "#47a8bd"),
    labels = labels_vec_Q1_Q2,
    name = "Speed Quartile"
  ) +
  guides(color = guide_legend(override.aes = list(alpha = 1, size = 2))) +
  facet_wrap(~ group) +
  labs(
    title = "Small Multiple of Speed Quartiles Q1 and Q2",
    subtitle = "January 02, 2024"
  ) +
  theme_solid() +
  theme(
    plot.title = element_text(size = 16),
    plot.subtitle = element_text(size = 16),
    plot.title.position = "plot"
  )
```
  
```{r plot_focused_higher_quartiles, warning=FALSE, message=FALSE}
  ggplot() +
  geom_sf(data = boroughs, color = "white") +  
  geom_sf(data = landings, color = "black", size = 2, alpha = 0.1) +
  geom_sf(data = Q1_Q2_sf,
          aes(color = group),
          size = 1, alpha = 0.01) +
  coord_sf(
    xlim = c(bbox_east_river["xmin"], bbox_east_river["xmax"]),
    ylim = c(bbox_east_river["ymin"], bbox_east_river["ymax"]),
    expand = FALSE
  ) +
  scale_color_manual(
    values = c("Q1" = "#1e3888",
               "Q2" = "#47a8bd"),
    labels = labels_vec,
    name = "Speed Quartile"
  ) +
  facet_wrap(~ group) +
  labs(
    title = "Small Multiple of Speed Quartiles Q1 and Q2 - Focused",
    subtitle = "January 02, 2024"
  ) +
  theme_solid() +
  theme(
    plot.title = element_text(size = 16),
    plot.subtitle = element_text(size = 16),
    plot.title.position = "plot"
  )
```

## Buffers

```{r buffers, warning=FALSE, message=FALSE}

landings_buffers_100ft <- st_buffer(landings, dist = 30.48) %>%
  st_transform(32118)

counts <- lengths(st_intersects(landings_buffers_100ft, jun2025_reference_week_excessively_lates_sf))

landings_buffers_100ft$record_count <- counts

```

```{r plot_buffers, warning=FALSE, message=FALSE}

ggplot() +
  geom_sf(data = boroughs, color = "white") +  
  geom_sf(data = landings, color = "black", size = 2, alpha = 0.2) +
  geom_sf(
    data = landings_buffers_100ft,
    aes(linewidth = record_count),
    fill = "#006bb6",
    color = alpha("#006bb6", 0.5)) +
  scale_linewidth(range = c(1, 10), name = "Count of Records") +
   labs(
    title = "Count of Records Around Landings"
  ) +
  theme_solid() +
  theme(
    plot.title = element_text(size = 20),
    plot.title.position = "plot"
  )
  
 
ggplot() +
  geom_sf(data = boroughs, color = "white") +  
  geom_sf(data = landings, color = "black", size = 2, alpha = 0.2) + 
  geom_sf(data = quartiles_sf,
          aes(color = group),
          size = 1, alpha = 0.01) +
  coord_sf(
    xlim = c(bbox_quartiles["xmin"], bbox_quartiles["xmax"]),
    ylim = c(bbox_quartiles["ymin"], bbox_quartiles["ymax"]),
    expand = FALSE
  ) +
  scale_color_manual(
    values = c("Q1" = "#1e3888",
               "Q2" = "#47a8bd",
               "Q3" = "#ffad69",
               "Q4" = "#9c3848"),
    labels = labels_vec,
    name = "Speed Quartile"
  ) +
  facet_wrap(~ group) +
  theme_solid()

ggplot() +
  geom_sf(data = boroughs, color = "white") +  
  geom_sf(data = landings, color = "black", size = 2, alpha = 0.5) + 
  geom_sf(data = quartiles_sf,
          aes(color = moreThanTenMin),
          size = 1, alpha = 0.5) +
  coord_sf(
    xlim = c(bbox_quartiles["xmin"], bbox_quartiles["xmax"]),
    ylim = c(bbox_quartiles["ymin"], bbox_quartiles["ymax"]),
    expand = FALSE
  ) +
  scale_color_manual(
    values = c("No" = "#47a8bd",
               "Yes" = "#9c3848"),
    labels = labels_vec,
    name = "OTP Over Ten Minutes Late"
  ) +
  facet_wrap(~ group) +
  theme_solid()

  
```

## Split by OTP

```{r more_than_ten_minutes_late, message=FALSE, warning=FALSE}

moreThanTenMinutesLate <- jun2025_reference_week %>% filter(OTP == "Excessively Late")
notMoreThanTenMinutesLate <- jun2025_reference_week %>% filter(OTP != "Excessively Late")

moreThanTenMinutesLate_sf <- st_as_sf(moreThanTenMinutesLate, coords = c("lon", "lat"), crs = 4326, remove = FALSE)

notMoreThanTenMinutesLate_sf <- st_as_sf(notMoreThanTenMinutesLate, coords = c("lon", "lat"), crs = 4326, remove = FALSE)
```

## Leaflet

```{r leaflet_plots, warning=FALSE, message=FALSE}

points_plot <- st_transform(landings, 4326)
buffers_plot <- st_transform(landings_buffers_100ft, 4326)
routes_plot <- st_transform(routes, 4326)
boroughs_plot <- st_transform(boroughs, 4326)

# leaflet() %>%
#   addProviderTiles("Esri.WorldImagery") %>%
#   addPolygons(
#     data = boroughs_plot,
#     color = "#828282",
#     weight = 1,
#     fillOpacity = 0.75
#   ) %>%  
#   addPolylines(
#     data = routes_plot,
#     color = "#006bb6",
#     weight = 2, 
#     opacity = 0.2
#   ) %>%
#   addPolygons(
#     data = buffers_plot,
#     color = "#f58426",
#     weight = 1,
#     fillOpacity = 0.75
#   ) %>%  
#   addCircleMarkers(
#     data = points_plot,
#     radius = 3,
#     color = "#f58426",
#     opacity = 0.6
#   )

quartiles_leaflet <- st_transform(quartiles_sf, 4326)
Q1_leaflet <- st_transform(Q1_sf, 4326)
Q2_leaflet <- st_transform(Q2_sf, 4326)
Q3_leaflet <- st_transform(Q3_sf, 4326)
Q4_leaflet <- st_transform(Q4_sf, 4326)


leaflet() %>%
  addProviderTiles("Esri.WorldImagery") %>%
  addPolygons(
    data = boroughs_plot,
    color = "#828282",
    weight = 1,
    fillOpacity = 0.75,
    group = "Boroughs"
  ) %>%
  addPolylines(
    data = routes_plot,
    color = "#006bb6",
    weight = 2,
    opacity = 0.2,
    group = "Routes"
  ) %>%
  addPolygons(
    data = buffers_plot,
    color = "#f58426",
    weight = 10,
    fillOpacity = 0.75,
    group = "Landings",
    label = ~paste0("Records: ", record_count)
  ) %>%
  addCircleMarkers(
    data = points_plot,
    radius = 3,
    color = "#f58426",
    opacity = 0.6,
    group = "Landings"
  ) %>%
  addCircleMarkers(data = Q1_leaflet, radius = 1, color = "#1e3888", opacity = 0.6, fillOpacity = 0.4, group = "Q1", label = ~paste0("Speed: ", round(knots,2)),
    popup = ~paste0("<b>Speed:</b> ", round(knots, 2), " m/s",
  "<br><b>Route:</b> ", routeId,
  "<br><b>Trip:</b> ", tripId,
  "<br><b>Vessel</b> ", vehicleId,
  "<br><b>Time:</b> ", time)) %>%
  addCircleMarkers(data = Q2_leaflet, radius = 1, color = "#47a8bd", opacity = 0.6, fillOpacity = 0.4, group = "Q2", label = ~paste0("Speed: ", round(knots,2)),
    popup = ~paste0("<b>Speed:</b> ", round(knots, 2), " m/s",
  "<br><b>Route:</b> ", routeId,
  "<br><b>Trip:</b> ", tripId,
  "<br><b>Vessel</b> ", vehicleId,
  "<br><b>Time:</b> ", time)) %>%
  addCircleMarkers(data = Q3_leaflet, radius = 1, color = "#ffad69", opacity = 0.6, fillOpacity = 0.4, group = "Q3", label = ~paste0("Speed: ", round(knots,2)),
    popup = ~paste0("<b>Speed:</b> ", round(knots, 2), " m/s",
  "<br><b>Route:</b> ", routeId,
  "<br><b>Trip:</b> ", tripId,
  "<br><b>Vessel</b> ", vehicleId,
  "<br><b>Time:</b> ", time)) %>%
  addCircleMarkers(data = Q4_leaflet, radius = 1, color = "#9c3848", opacity = 0.6, fillOpacity = 0.4, group = "Q2", label = ~paste0("Speed: ", round(knots,2)),
    popup = ~paste0("<b>Speed:</b> ", round(knots, 2), " m/s",
  "<br><b>Route:</b> ", routeId,
  "<br><b>Trip:</b> ", tripId,
  "<br><b>Vessel</b> ", vehicleId,
  "<br><b>Time:</b> ", time)) %>%
  addLayersControl(
    overlayGroups = c("Boroughs", "Routes", "Landings", "Q1", "Q2", "Q3", "Q4"),
    baseGroups = c("Imagery"),
    options = layersControlOptions(collapsed = FALSE)
  )

leaflet() %>%
  addProviderTiles("Esri.WorldImagery", group = "Imagery") %>%
  addPolygons(data = boroughs_plot, color = "#828282", weight = 1, fillOpacity = 0.75, group = "Boroughs") %>%
  addPolylines(data = routes_plot, color = "#006bb6", weight = 2, opacity = 0.2, group = "Routes") %>%
  addPolygons(data = buffers_plot, color = "#f58426", weight = 10, fillOpacity = 0.75, group = "Landings",
              label = ~paste0("Records: ", record_count)) %>%
  addCircleMarkers(data = points_plot, radius = 3, color = "#f58426", opacity = 0.6, group = "Landings") %>%
  addCircleMarkers(
    data = moreThanTenMinutesLate_sf, radius = 1, color = "#47a8bd",
    opacity = 0.6, fillOpacity = 0.4, group = "Not Excessively Late",
    label = ~paste0("Speed: ", round(knots, 2)),
    popup = ~paste0("<b>Speed:</b> ", round(knots, 2), " m/s",
                    "<br><b>Route:</b> ", routeId,
                    "<br><b>Trip:</b> ", tripId,
                    "<br><b>Vessel</b> ", vehicleId,
                    "<br><b>Time:</b> ", time)
  ) %>%
  addCircleMarkers(
    data = notMoreThanTenMinutesLate_sf, radius = 1, color = "#9c3848",
    opacity = 0.6, fillOpacity = 0.4, group = "Excessively Late",
    label = ~paste0("Speed: ", round(knots, 2)),
    popup = ~paste0("<b>Speed:</b> ", round(knots, 2), " m/s",
                    "<br><b>Route:</b> ", routeId,
                    "<br><b>Trip:</b> ", tripId,
                    "<br><b>Vessel</b> ", vehicleId,
                    "<br><b>Time:</b> ", time)
  ) %>%
  addLayersControl(
    overlayGroups = c("Boroughs", "Routes", "Landings", "Not Excessively Late", "Excessively Late"),
    baseGroups = c("Imagery"),
    options = layersControlOptions(collapsed = FALSE)
  ) %>%
  htmlwidgets::onRender("
    function(el, x) {
      var map = this;

      function logBounds() {
        var b = map.getBounds();
        console.log('bbox:',
          'xmin', b.getWest(),
          'ymin', b.getSouth(),
          'xmax', b.getEast(),
          'ymax', b.getNorth()
        );
      }

      map.whenReady(function() {
        logBounds();               // log once on load
        map.on('moveend', logBounds);
      });
    }
  ") %>%
  addControl(
    html = "<div class='map-title'>My Map Title</div>",
    position = "topleft"
  )

```